---
title: "Comparison_with_lmerseq"
format: html
editor: visual
bibliography: references.bib
---

## Introduction

Here, a basic comparison is made between lmerseq [@vestal2022] and Seqwrap using the [@pillon2022] gene expression data

## Data extraction/preprocessing

Here the Pillion gene expression data is downloaded and transformed using VST as required by lmerseq

```{r}
#| label: libraries and data preprocessing
#| code-fold: true
#| echo: false
library(lmerSeq)
library(DESeq2)
library(edgeR)
library(seqwraphelper)
library(dplyr)
library(tidyverse)
library(seqwrap)
library(purrr)
library(glmmSeq)
library(ggplot2)

# download the data from seqwraphelper
dat <- seqwraphelper::geo_data(gse = "GSE202295")




# Filter low expressed gened and combine after filtering
keep <- filterByExpr(
  dat$countdata[,-1],
  min.count = 10,
  min.total.count = 15,
  large.n = 10,
  min.prop = 0.7,
  group = paste(dat$metadata$group, dat$metadata$time))

countdat <-  dat$countdata[keep,]



metadat <- dat$metadata



# Check if sequence_ID matches. If False, make to match 
all(colnames(countdat)[-1] == metadat$seq_sample_id)



# Normalize the count data using DESeq2's VST
dds <- DESeqDataSetFromMatrix(countData = countdat[,-1],
                              colData = metadat,
                              design = ~ time * group)

dds <- DESeq(dds)
vsd.fixed <- varianceStabilizingTransformation(dds, blind=F)
vst_expr <- assay(vsd.fixed)





# Add as gene_ids to the variance_stabilised_counts
rownames(vst_expr) <- countdat$geneid

# subset to only the first fifteen genes
vst_expr <- vst_expr[1:15,]
```

Lmerseq has a lmerseq.fit() function that builds the model. Lmerseq by default uses 2 cores. This will be set in the other packages too

```{r}
#| label: lmerseq
#| code-fold: true
#| echo: false

# fit the model using lmerseq
fit.lmerSeq <- lmerSeq.fit(form = ~ time * group  + (1|id),
                           expr_mat = vst_expr,
                           sample_data = metadat,
                           REML = T)




coefs <- c("timepost", "timerec", "groupT2D", "timepost:groupT2D", "timerec:groupT2D")


lmerseq_summary <- coefs %>%
  map_df(~ lmerSeq.summary(
    lmerSeq_results = fit.lmerSeq,
    coefficient = .x,
    p_adj_method = "BH",
    ddf = "Satterthwaite",
    sort_results = FALSE
  )$summary_table %>%
    mutate(Coefficient = .x)
  )


saveRDS(lmerseq_summary, "data/lmerseq_model_summary.RDS")
```

### Seqwrap

Building same model with seqwrap

```{r}
#| label: Seqwrap
#| code-fold: true
#| echo: false

# make data suitable for seqwrap as it accepts only dataframes or lists
vst_expr_df <- data.frame(vst_expr) %>%
  tibble::rownames_to_column(var = "target")


# first validate input using seqwrap compose
seqwrap_val <- seqwrap_compose(
  data = vst_expr_df,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = lmerTest:::lmer,
  arguments = list(
    formula = y ~ time * group  + (1|id))
)


  seqwrap_model <- seqwrap(seqwrap_val,
                        return_models = FALSE,
                        cores = 2)

seqwrap_summary_df <- seqwrap_summarise(seqwrap_model)$summaries


saveRDS(seqwrap_summary_df, "data/seqwrap_model_summary.RDS")
```

### Visualisation

The estimates and pvalues of both models are to be visualised

```{r}
#| label: Visualisation
#| code-fold: true
#| echo: false


# Select
seqwrap_df <- seqwrap_summary_df %>%
  subset(!term %in% c("(Intercept)", "sd__(Intercept)", "sd__Observation")) %>%
  dplyr::select("target",  "estimate", "p.value", "term" ) %>%
  mutate(source = "Seqwrap")


lmerseq_df <- lmerseq_summary %>%
  dplyr::select(gene, Estimate, p_val_raw, Coefficient) %>%
  mutate(source = "lmerseq") %>%
  #rename to match the columns from seqwrap summary
  rename(target = gene,
         estimate = Estimate,
         p.value = p_val_raw,
         term = Coefficient)


merged <- merge(seqwrap_df, lmerseq_df, by = c("target", "term"), suffixes = c("_seqwrap", "_lmerseq"))

ggplot(merged, aes(x = estimate_seqwrap, y = estimate_lmerseq, color = target)) +
   geom_point(size = 2) +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
   facet_wrap(~ term)
   labs(title = "Comparison of Estimates", x = "seqwrap Estimate", y = "lmerseq Estimate") +
   theme_minimal()





   
   
   ggplot(merged, aes(x = p.value_seqwrap, y = p.value_lmerseq, color = target)) +
     geom_point(size = 2) +
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
     facet_wrap(~ term)
   labs(title = "Comparison of p values", x = "seqwrap p values", y = "lmerseq pvalues") +
     theme_minimal()


```

```{}
```
