---
title: "Shetty et al. (2024) Re-analysis"
format: html
editor: visual
execute: 
  echo: false
editor_options: 
  chunk_output_type: console
---

## TODO

- Find missing samples
- Re-do analysis using formal mixed-effects model 



```{r}
# Load the packages
library(seqwrap)
library(dplyr)
library(tidyverse)
library(lme4)
library(seqwraphelper)
library(glmmTMB)
library(nlme)
```

First extract the alcohol study data in seqwraphelper. Link to raw data <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE232408>

Link to publication <https://pmc.ncbi.nlm.nih.gov/articles/PMC11087488/#>*ad93*

```{r}


# gene level data 
gene_data <- seqwraphelper::shetty_gene



metadata <- seqwraphelper::shetty_metadata |> 
  # drop the tissue colume
  dplyr::select(-tissue, -Run) |> 
  mutate(sex = factor(sex, levels =c("Male", "Female")),
         dose = factor(dose, levels = c("Placebo", "MediumDose", "HighDose")),
         period = factor(period, levels = c("1", "2","3")),
         time = factor(time, levels = c("0", "3")),
         study_site = factor(study_site, levels = c("1", "2")))


## Extract and combine sequence data 
sequence_data <- metadata |> 
  mutate(sequence = paste(period, dose, sep = "_")) |> 
  
  distinct(sequence, participant, period) |> 
  
  summarise(.by = participant, 
            sequence = paste(sequence, collapse = "-"))


## Extract carryover 
metadata2 <- metadata |> 
  
  inner_join(sequence_data) |> 
  # Add carryover
  arrange(participant, period) |> 
  mutate(.by = participant, 
         carryover = lag(dose), 
         carryover = if_else(is.na(carryover), "First", carryover)) |> 
  # Fix time and period to factor variables 
  mutate(period = paste0("p", period), 
         time = paste0("t", time), 
         dose = factor(dose, levels = c("Placebo", "MediumDose", "HighDose")))
           
  

## Check if samples are in meta data 
# In case order of sample Ids do not match, reorder to match
all(colnames(gene_data)[-1] %in% metadata$seq_sample_id)


# Check if everything matches except the gene_id
all(match(colnames(gene_data[,-1]), metadata2$seq_sample_id))
```

As our primary interest lies in the relationship between alcohol dosage and gene expression, we want to control for the order of treatments, or any carryover effects between treatments, with the (average) interaction between `time` and `dose` as the main focus for our model. In Shetty et al. [-@shetty_peripheral_2024], participants were used as the clustering variable for random-effects intercepts while dose, time and sequence were used as fixed effects. An alternative formulation could include a factor indicating the previous treatment to model carryover effects. 



```{r}

# Filter away low-expression genes
cpm_values <- t(t(gene_data[, -1] ) / colSums(gene_data[, -1] )) * 1e6
keep <- rowSums(cpm_values > 1) >= 50

gene_filtered <- gene_data[keep, ]

temp <- slice_sample(gene_filtered, n = 100) |> 
  bind_rows(filter(gene_filtered, gene_id == "ENSG00000180537"))


m1 <- seqwrap_compose(data = temp, 
                metadata = metadata2, 
                modelfun = glmmTMB::glmmTMB, 
                samplename = "seq_sample_id",
                arguments = list(formula = y ~ sex + period + dose*time + (1|participant), 
                                 family = glmmTMB::nbinom1()))


m2 <- seqwrap_compose(data = temp, 
                metadata = metadata2, 
                modelfun = glmmTMB::glmmTMB, 
                samplename = "seq_sample_id",
                arguments = list(formula = y ~ sex + carryover + dose*time + (1|participant), 
                                 family = glmmTMB::nbinom1()))








m1_results <- seqwrap(m1, 
                      return_models = TRUE)

m2_results <- seqwrap(m2, 
                      return_models = TRUE)





summary(m1_results@models[[1]])
summary(m2_results@models[[2]])


m1_sum <- seqwrap_summarise(m1_results)
m2_sum <- seqwrap_summarise(m2_results)

m1_sum$summaries |> 
  filter(target == "RNF182") |> 
  print(n = 22)

m2_sum$summaries |> 
  filter(target == "ENSG00000180537") |> 
  print(n = 22)



```








Analysing the data using GlmmTmb . Here, we analyse the gene counts using a negative binomial mixed-effects model . We include dose, time ,sex and the interaction between dose and time as fixed effects. For random effects, we allow each participant to have its own baseline and time trend but assume no systematic relationship between the two. We also include study site as a source of random variation.

Allowing correlation between time and participant (1+ time \| participant) caused convergence issues a significant number of genes

```{r}

# initialise arguments using the algorithms syntax
args <- list(formula = y ~ dose*time + sex +  period + (1+ time || participant) + (1|study_site) ,
             family = glmmTMB::nbinom1)


glmm_container <- seqwrap_compose(data = gene_data,
                             metadata = metadata,
                             samplename = "seq_sample_id",
                             modelfun = glmmTMB::glmmTMB ,
                             arguments = args,
                          #   summary_fun = sumfun,
                          #  eval_fun = evalfun
                            )

glmm_model <- seqwrap(glmm_container,
                 #summary_fun = sumfun,
                 #eval_fun = evalfun,
                 return_models = F,
                # subset = 1:19,
                 cores = 8)

saveRDS(glmm_model, "data/seqwrap_glmm_model.RDS")

glmm_model_summary <- seqwrap_summarise(glmm_model)


glmm_model_summary$summaries


glmm_model_summary$evaluations

seqwrapResults(glmm_model_summary$summaries)
```

Building a model using lme4

```{r}


lmer_container <- seqwrap_compose(
  data = gene_data,
  metadata = metadata,
  samplename = "seq_sample_id",
  modelfun = lme4::glmer,
  arguments = list(formula = y ~ dose*time + sex +  period + (1+ time || participant) + (1|study_site),
  family = poisson),
)

lmer_model <- seqwrap(
  lmer_container,
  return_models = F,
  #subset = 1:15,
  cores = 2
)
lmer_model_summary <- seqwrap_summarise(lmer_model)

saveRDS(lmer_model, "data/seqwrap_lmer_model.RDS")

lmer_model_summary$summaries


lmer_model_summary$evaluations
```
