---
title: "seqwrap: an R package for flexible iterative fitting of high-dimensional data"
subtitle: "Supplementary material"
author:
  - name: Chidimma Echebiri
    affiliation:
      - name: University of Inland Norway
        department: Department of Biotechnology, PhD programme in applied ecology and biotechnology
  - name: Rafi Ahmad
    affiliation: 
      - name: University of Inland Norway
        department: Department of Biotechnology
  - name: Stian Ellefsen
    affiliation: 
      - name: University of Inland Norway
        department: Department of Public Health and Sport Sciences 
  - name: Daniel Hammarström
    email: daniel.hammarstrom@inn.no
    affiliation: 
      - name: University of Inland Norway
        department: Department of Public Health and Sport Sciences 
    attributes:
      corresponding: true
format:
  typst:
    toc: true
    include-in-header:
      text: |
        #show outline: it => {
          v(2em)  // Add vertical space before TOC
          it
        }
csl: resources/bmc-bioinformatics.csl
bibliography: resources/bibliography.bib
execute: 
  cache: true
---

## Fitting models with `seqwrap`

The `seqwrap` package is designed to iterate over multiple targets in a high-dimensional data set to fit a user-defined regression model using a user-selected model fitting function. Each target may represent, e.g., a gene, an exon, a protein, a methylation site, or any other outcome from a high-throughput omics platform. Each target is measured across a set of samples, with a metadata set used to describe study covariates. In the basic case, data and metadata are combined in each iteration to fit the selected model. 

Dungan et al. [@dungan2022] evaluated a senolytic treatment (dasatinib and quercetin) in mouse skeletal muscle subjected to mechanical overload through synergist ablation. The study included sham surgery and a vehicle control, allowing analysis of the effect of senolytic treatment and the interaction between mechanical overloading and senolytic treatment. RNA-seq was performed on 20 animals (5 per group). We will start by downloading the data (raw count matrix) and filtering low-expression genes to prepare the count data. Next, we will prepare the metadata with the relevant study covariates.


```{r}
#| labels: packages-and-data 
#| message: false
#| warning: false

# Load packages
library(tidyverse)
library(seqwrap)
library(edgeR)
library(gt)
library(GEOquery)


# The data set is available at GEO accession number GSE195707  
url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE195nnn/GSE195707/suppl/GSE195707%5Fcount%5Fmatrix%2Etxt%2Egz"

# Download the count file 
temp <- tempfile(fileext = ".txt.gz")
download.file(url, temp, mode = "wb")
counts <- read.table(temp, header = TRUE, sep = "\t")
unlink(temp)

# Process the counts
counts <- counts |> 
  # Change estimated counts to integers
  mutate(across(where(is.numeric), ~as.integer(round(.)))) |> 
  # Filter rows with min less than 10
  filter(apply(across(where(is.numeric)), 1, min) >= 10)


# Create a meta data df 
gse <- getGEO("GSE195707", GSEMatrix = TRUE)


metadata <- data.frame(gse$GSE195707_series_matrix.txt.gz, 
                       row.names = NULL) |> 
  dplyr::select(seq_sample_id = description, 
                treatment = characteristics_ch1.1, 
                surgery = characteristics_ch1.2) |> 
  mutate(treatment = gsub("treatment: ", "", treatment), 
         surgery = gsub("surgery: ", "", surgery), 
         treatment = factor(treatment, levels = c("Vehicle", "Senolytic")), 
         surgery = factor(surgery, levels = c("Sham", "Overload"))) 
           
```

The metadata contains the sample ID corresponding to columns in the counts data set. We are using `seq_sample_id` as the column name here. The first rows of the prepared metadata are displayed in @tbl-meta-dungan.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-meta-dungan
#| tbl-cap: "Selected rows from the Dungan et al. metadata data frame."


metadata |> 
  slice_head(n = 2, by = c(treatment, surgery)) |> 
  gt()

# Save seq_sample_id for display
samples <- metadata |> 
  slice_head(n = 2, by = c(treatment, surgery)) |> 
  pull(seq_sample_id)


```

The count data matrix can be manipulated as a data frame, with the first column indicating the targets (in this case, gene symbols). Importantly, additional column names have names corresponding to the `seq_sample_id` variable in the metadata. As an example, the first rows of the count data frame are shown for the same samples as in the metadata data frame (see @tbl-dungan-counts). `seqwrap` expects that targets are unique so before going further we should validate this in the data. When doing so we find that five targets have duplicate rows in the data set, for simplicity, we will remove them.   

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-dungan-counts
#| tbl-cap: "Selected rows from the Dungan et al. counts data frame."

counts |> 
  slice_head(n = 10) |> 
  dplyr::select(gene_name, all_of(samples)) |> 
  gt()

```



```{r}

# Find targets with more than a unique row in the data
rm_targets <- counts |> 
  summarise(.by = gene_name, 
            n = n()) |> 
  filter(n > 1) |> 
  pull(gene_name)

# Filter out non-uniqe rows
counts <- counts |> 
  filter(!(gene_name %in% rm_targets))

```


```{r}
#| echo: false

# Count the number of targets
n_genes <- nrow(counts)

```


The data suggests a negative binomial model where the linear predictor can be used to assess the difference between senolytic treatment and control (treat; $\beta_1g$), mechanical overloading compared to sham surgery (mov; $\beta_2g$), and their interaction ($\beta_3g$) for gene *g* in 1 to `r n_genes` and row *i* in 1 to 20 of the meta data (@eq-m1). We want to control for differences in sequencing depth by including an offset (*s*~i~). We will use the TMM method [@robinson2010] to calculate the effective library size.


$$
\begin{align}
y_{g[i]} &\sim \operatorname{NB2}(\mu_{g[i]}, \theta_g) \\
\operatorname{log}(\mu_{g[i]}) &= \beta_0 + \beta_{1g}\text{treat}_i + \beta_{2g}\text{mov}_i + \beta_{3g}(\text{treat}_i \times  \text{mov}_i) + \operatorname{log}(s_i)
\end{align}
$$ {#eq-m1}

Below, we are using the `edgeR` package to calculate the library size normalization factor. Effective library sizes are then expressed as relative to the median effective library and log transformed.

```{r}

# Use EdgeR to calculate the TMM
y <- edgeR::DGEList(counts[,-1])
y <- edgeR::calcNormFactors(y)

# Combine the data into metadata
metadata <- metadata |> 
  inner_join(
    y$samples |> 
  tibble::rownames_to_column("seq_sample_id")
  ) |> 
   mutate(
     efflibsize = (lib.size * norm.factors) / median(lib.size * norm.factors),
     efflibsize = log(efflibsize)
   ) 


```


`seqwrap_compose` is used to combine the data, metadata and the selected modelling function with its arguments into a `swobject`. A negative binomial model can be fitted using e.g., `glmmTMB` from the `glmmTMB package (glmmTMB::glmmTMB`) or `MASS::glm.nb`, here we will use `glmmTMB` as it can be extended in subsequent modelling steps. 

The specified `glmmTMB` function (`modelfun` slot) needs a formula and a family function, these are supplied in the arguments slot. The count data are supplied in the `data` slot and the metadata in the `metadata` slot.  Notice that we use the namespace operator (`::`) to specify selected functions to avoid difficulties in parallel processing. The `samplename` indicates the column name in the metadata that gives sample names.   


```{r}

m1.1 <- seqwrap_compose(
  modelfun = glmmTMB::glmmTMB,
  data = counts, 
  metadata = metadata, 
  arguments = list(formula = y ~ treatment * surgery + offset(efflibsize), 
                family = glmmTMB::nbinom2),
  samplename = "seq_sample_id"
  )

```

Next, we will run the model using `seqwrap`. We recommend using `subset` to specify a smaller number of targets when prototyping the model. Here, we will assess whether the model output matches expectations. By specifying `return_models = TRUE`, all model objects will be returned in the results object, which can be inspected if needed. Here we are using a single core for fitting the models, specified in the `cores` argument.

```{r}

m1.1_temp <- seqwrap(
  m1.1, 
  cores = 1, 
  subset = 1:24, 
  return_models = TRUE, 
  verbose = FALSE
)

```

Following iterative fitting using `seqwrap`, we can summarize the models using `seqwrap_summarise`, which produces a list of two summaries. Below, we are saving the summaries in an object for later access. Running the `seqwrap_summarise` function also gives us an overview of the model and successfully summarized targets.

In the `summaries` slot, model parameter values are returned (see @tbl-dungan-sum-temp). These are produced using `broom.mixed::tidy` which is a general function to retrieve model parameters from many regression model packages in R. 

In the `evaluations` slot, each model is evaluated using tests from the [DHARMa package](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html) (see @tbl-dungan-eval-temp). 

```{r}

m1.1_temp_sum <- seqwrap_summarise(m1.1_temp, verbose = FALSE)

```

```{=typst}
#set text(size: 8pt)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-dungan-sum-temp
#| tbl-cap: "The first ten rows from summarised preliminary models (model summaries)."


m1.1_temp_sum$summaries |> 
  slice_head(n = 10) |> 
  gt() 

```

```{=typst}
#set text(size: 11pt)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-dungan-eval-temp
#| tbl-cap: "The first ten rows from summarised preliminary models (model evaluations). Values are p-values from the DHARMa package (see [here for details](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html))."


m1.1_temp_sum$evaluations |> 
  slice_head(n = 10) |> 
  gt()

```


### Specifying summary and evaluation function

We can specify custom functions for evaluating the model object. `seqwrap` allows for two slots (summaries and evaluations). Each function should take the model object as an argument, and for compatibility with `seqwrap_summarise`, return a data frame. For our next step, we would like to retrieve the dispersion parameter from each model. We will place this parameter in the evaluation function. We can extract the dispersion parameter on the log scale from the `glmmTMB` model object. We will construct a function around the extraction and include the observed counts from the data:


```{r}

dispersion_fun <- function(x) {
  
    out <- data.frame(
      dispersion =  data.frame(summary(x$sdr))["betadisp",1],
      dispersion.se = data.frame(summary(x$sdr))["betadisp",2],
      log_mu = log(mean(x$frame$y))
      )
  
    return(out)
  
}
```

Using our preliminary models, we can validate the function for one model.

```{r}
dispersion_fun(m1.1_temp@models[[1]])
```

The output is, as expected, a data frame with three columns. The dispersion parameter is expected to be on the log scale. To validate this we can directly access the estimated dispersion parameter from the model object using `sigma()`, which return the dispersion parameter on the natural scale. 

```{r}
log(sigma(m1.1_temp@models[[1]]))
```


### Running the full model

When we are happy with preliminary performance on a subset of targets, we can initialize `seqwrap` to iterate over all targets. We are updating the previous model with the new evaluation function directly in `seqwrap`. Notice that we also remove the `subset` and set `return_models` to `FALSE`. Saving all model objects will be too memory-consuming.

We should expect some errors from modelling many thousands of targets. As seen in the note from `seqwrap`, some models are associated with warnings from the modelling algorithm and/or errors from the evaluation function.


```{r}


m1.2_results <- seqwrap(
  m1.1,
  eval_fun = dispersion_fun,
  cores = 12, 
  return_models = FALSE,
  verbose = FALSE
)


```


Errors can be inspected by accessing the errors slot. From the inspection below, warnings are at least to some extent related to model convergence.

```{r}

warnings <- m1.2_results@errors|> 
  # filter the non-null elements
  filter(map_lgl(warnings_fit, ~ !is.null(.x[[1]]))) |> 
  # save in object
  pull(warnings_fit)

warnings[1]

```


### Using target data for prior information in `glmmTMB` 

A major benefit of canonical R packages for RNA-seq data modelling is the sharing of information between targets through shrinkage and empirical Bayes strategies. We suggest (see the main text) using priors in `glmmTMB` as convenient way of flexibly regularize models using an empirical Bayes strategy. In `glmmTMB`, the `priors` argument can be used to set priors on parameters. We will use the estimated parameters from the first model to set up the priors.

First, we need to extract the parameters. From the evaluation function we get the average log count and the dispersion parameter estimate. From the summaries, we will get the other parameters (summarized in @fig-dungan-estimates).


```{r}

m1.2_sum <- seqwrap_summarise(m1.2_results, verbose = FALSE)

```



```{r}
#| fig-cap: "Estimates from modelling the Dungan data. The dispersion trend is visualized using the default `ggplot`2::geom_smooth`."
#| echo: false
#| message: false
#| warning: false
#| label: fig-dungan-estimates

library(cowplot)
library(ggtext)


p1 <- m1.2_sum$evaluations |> 
  ggplot(aes(log_mu, dispersion)) + 
  geom_point(alpha = 0.2) + 
  geom_smooth(se = FALSE, color = "orange") + 
  theme_classic() +
  labs(x = "Average raw counts log(&mu;) ", 
       y = "Dispersion log(&theta;)") + 
  theme(axis.title.x = element_markdown(), 
        axis.title.y = element_markdown())
  

p2 <- m1.2_sum$summaries |> 
  
  ggplot(aes(estimate)) + 
  geom_density() + 
  facet_wrap(~ term, scales = "free") +
  theme_classic() + 
  labs(x = "Estimate", y = "")


plot_grid(
  plot_grid(NULL, p1, NULL, rel_heights = c(0.2, 1, 0.2), ncol = 1),
            p2, 
          ncol = 2, 
          rel_widths = c(1, 1.5))

```


Except for the dispersion values, we can easily summarize parameter values by their mean and standard deviation. A more elaborate summary of the dispersion values could include an estimate of the mean (log) dispersion, conditional on the average log counts. As the dispersion prior will be slightly different for each target (based on the observed counts), we will combine priors into target-specific data frames and store them as a list. `seqwrap` will make target-specific list elements available in each iteration, making it possible to build the prior data frame for `glmmTMB`.

```{r}

# Summarise parameters from the main output
param_sum <- m1.2_sum$summaries |> 
  summarise(.by = term, 
            m = mean(estimate), 
            s = sd(estimate)) 

# Fit a model to the log dispersion values
trend_model <- loess(dispersion ~ log_mu,
                         data = m1.2_sum$evaluations,
                         span = 0.7)


# Predict dispersion for each gene based on log raw counts
# and combine into a prior.
dispersion_prior <- data.frame(gene =  counts[,1],
                               pred = round(
                                 predict(trend_model,
                                         newdata = data.frame(
                                           log_mu =  log(rowMeans(counts[,-1]))
                                           )
                                   ), 3),
                                 s = round(
                                   trend_model$s, 3)
  ) |>
    mutate(prior = paste0("normal(", pred, ",", s, ")"))

# Combine parameters in a data frame...
Priors_df <- data.frame(prior = paste0("normal(",
                                       round(param_sum$m, 2),
                                       ",",
                                       round(param_sum$s,2 ), 
                                       ")"),
                        class = rep("fixef", 4),
                        coef = param_sum$term)


# ... extract a prediction for the mean dispersion parameter per target
# and combine with other parameters  
Priors_list <- list()
for( j in 1:nrow( counts )) {


    df <- bind_rows(Priors_df,
                    data.frame(
                      prior =
                        dispersion_prior[dispersion_prior$gene ==
                                           counts[j,1],4],
                      class = "fixef_disp",
                      coef = "1"
                    )
    )

    Priors_list[[j]] <- df

}

```


Again, we will combine data, metadata, and arguments using `seqwrap_compose`. We are now also adding the priors to the arguments. The variables used in the priors data frame are made available from the target-specific list referenced in `targetdata = Prior_list`. Using `arguments = alist(... ` avoids evaluation of the list so that the data frame is not evaluated before being used in each iteration. Priors have been built corresponding what is expected by `glmmTMB` (see [Priors in glmmTMB](https://glmmtmb.github.io/glmmTMB/articles/priors.html)).

```{r}


m1.3 <- seqwrap_compose(
  modelfun = glmmTMB::glmmTMB,
  data = counts, 
  metadata = metadata, 
  eval_fun = dispersion_fun, 
  arguments =  alist(
    formula = y ~ treatment * surgery + offset(efflibsize), 
    family = glmmTMB::nbinom2,
      priors = data.frame(
        prior = prior,
        class = class,
        coef = coef)), 
  targetdata = Priors_list
)


m1.3_temp <- seqwrap(
  m1.3, 
  subset = 1:24, 
  return_models = TRUE, 
  verbose = FALSE
)


summary(m1.3_temp@models[[1]])

```

Running the prototyping seems to work, we will proceed with the full data set.

```{r}

m1.3_results <- seqwrap(
  m1.3, 
  cores = 12,
  return_models = FALSE, 
  verbose = FALSE
)

m1.3_sum <- seqwrap_summarise(m1.3_results, verbose = FALSE)

```

Regularization increases the number models without convergence issues, in fact all `r n_genes` targets have evaluations and summaries.

The summaries data frame is easily manipulated in subsequent analysis steps to, e.g., adjust p-values and create basic visualizations of the results (@fig-dungan-final-estimates).

```{r}
#| fig-cap: "A visualization of results from the regularized model of the Dungan et al. data."
#| echo: false
#| message: false
#| warning: false
#| label: fig-dungan-final-estimates



m1.3_sum$summaries |> 

  filter(term == "treatmentSenolytic:surgeryOverload") |> 
  mutate(fdr = p.adjust(p.value, method = "fdr"), 
         sig = if_else(fdr < 0.05, "significant", "non-significant"), 
         SYMBOL = if_else(sig == "significant", target, "")) |> 
  
  ggplot(aes(estimate, -log2(p.value), 
             color = sig)) + 
  geom_point(alpha = 0.4) + 
  
  geom_text(aes(label = SYMBOL), 
      show.legend = FALSE,
            position = position_nudge(y = 1)) +
  
  theme_classic() + 
  labs(x = "Estimate (log-scaled)", 
       y = "-log<sub>2</sub>(p-value)", 
       color = "FDR < 0.05", 
       subtitle = "Overload:Senolytic treatment") + 
  theme(axis.title.y = element_markdown()) 

  
  


```


The above example shows how to build models with `seqwrap`. Importantly, we make no claims about the validity of the final model but urge the user to assess assumptions in their data and models as part of a principled workflow.


## Comparing `lmerSeq` and `seqwrap`

`lmerseq` is a package for fitting RNA-seq data using mixed-effects models assuming Gaussian errors on transformed count data. Here, we perform a basic comparison between `lmerseq` [@vestal_lmerseq_2022] and `seqwrap`, using a subset from the gene expression data set presented in the main text from Pillon et al. [@pillon2022].

### Data preprocessing

Here the Pillion gene expression data is downloaded and transformed using VST transform from `DESeq2` as suggested  in [@vestal_lmerseq_2022].

```{r}
#| label: comp-data
#| echo: true
#| message: false
#| warning: false

# Need installation from github: https://github.com/stop-pre16/lmerSeq
library(lmerSeq) 
library(DESeq2)
library(edgeR)
# Need installation from github: https://github.com/dhammarstrom/seqwraphelper
library(seqwraphelper)
library(dplyr)
library(tidyverse)
library(seqwrap)
library(purrr)
library(glmmSeq)
library(ggplot2)

# download the data from seqwraphelper
dat <- seqwraphelper::geo_data(gse = "GSE202295")


# Filter low expressed genes and combine after filtering
keep <- filterByExpr(
  dat$countdata[,-1],
  min.count = 10,
  min.total.count = 15,
  large.n = 10,
  min.prop = 0.7,
  group = paste(dat$metadata$group, dat$metadata$time))

countdat <-  dat$countdata[keep,]

metadat <- dat$metadata


# Check if sequence_ID matches.
all(colnames(countdat)[-1] == metadat$seq_sample_id)


# Transform the count data using DESeq2's VST
dds <- DESeqDataSetFromMatrix(countData = countdat[,-1],
                              colData = metadat,
                              design = ~ time * group)

dds <- DESeq(dds)
vsd.fixed <- varianceStabilizingTransformation(dds, blind=F)
vst_expr <- assay(vsd.fixed)





# Add as gene_ids to the VST-transformed counts
rownames(vst_expr) <- countdat$geneid

# subset to only the first fifteen genes
vst_expr <- vst_expr[1:15,]

```


### Fitting models in `lmerseq`

`lmerseq` has a `lmerseq.fit()` function for building and fitting the model.

```{r}
#| label: lmerseq
#| echo: true
#| message: false
#| warning: false

# fit the model using lmerseq
fit.lmerSeq <- lmerSeq.fit(form = ~ time * group  + (1|id),
                           expr_mat = vst_expr,
                           sample_data = metadat,
                           parallel = FALSE,
                           REML = TRUE)



# Extract coefficients from the models 
coefs <- c("timepost", "timerec", "groupT2D", "timepost:groupT2D", "timerec:groupT2D")

# Extract all coefficients from lmerseq models
lmerseq_summary <- coefs |> 
  map_df(~ lmerSeq.summary(
    lmerSeq_results = fit.lmerSeq,
    coefficient = .x,
    p_adj_method = "BH",
    ddf = "Satterthwaite",
    sort_results = FALSE
  )$summary_table |> 
    mutate(Coefficient = .x)
  )



```


### Fitting models in transformed data in `seqwrap`

Below we are building the same model with `seqwrap`. We are using `lmerTest::lmer` to get p-values from each model. `lmerseq` filters targets from the output due to singular fit, here we add a small function that evaluates singular fits in the `lme4`-based model. We will use only models with non-singular fits in the comparison.   

```{r}
#| label: seqwrap
#| echo: true
#| message: false
#| warnings: false

# make data suitable for seqwrap as it accepts only data frames or lists
vst_expr_df <- data.frame(vst_expr) %>%
  tibble::rownames_to_column(var = "target")

# Creating a evaluation function for finding singular fits
singular_eval <- function(x) {
  out <- data.frame(isSingular = lme4::isSingular(x))
  return(out)
}



# first validate input using seqwrap compose
seqwrap_val <- seqwrap_compose(
  data = vst_expr_df,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = lmerTest::lmer,
  arguments = list(
    formula = y ~ time * group  + (1|id)), 
  eval_fun = singular_eval
)


seqwrap_model <- seqwrap(seqwrap_val,
                        return_models = FALSE,
                        cores = 1, 
                        verbose = FALSE)



# Check for singular fits and list only targets without 
targets <- seqwrap_summarise(seqwrap_model, 
                             verbose = FALSE)$evaluations |> 
  filter(!isSingular) |> 
  pull(target)


seqwrap_summary_df <- seqwrap_summarise(seqwrap_model, 
                                        verbose = FALSE)$summaries |> 
  filter(target %in% targets)


```

### Comparing results from `lmerseq`and `seqwrap`

Since both `lmerseq`and `seqwrap` uses the same underlying model and data, the output estimates are the same. The estimates and p-values of both approaches are visualized in @fig-comp-visualisation

```{r}
#| label: fig-comp-visualisation
#| echo: false
#| fig-cap: "Comparing estimates from `lmerseq` and `seqwrap`"
#| fig-height: 8


# Select
seqwrap_df <- seqwrap_summary_df %>%
  subset(!term %in% c("(Intercept)", "sd__(Intercept)", "sd__Observation")) %>%
  dplyr::select("target",  "estimate", "p.value", "term" ) %>%
  mutate(source = "seqwrap")


lmerseq_df <- lmerseq_summary |>
  dplyr::select(gene, Estimate, p_val_raw, Coefficient) |>
  mutate(source = "lmerseq") |>
  #rename to match the columns from seqwrap summary
  dplyr::rename(target = gene,
         estimate = Estimate,
         p.value = p_val_raw,
         term = Coefficient)



p1 <- bind_rows(seqwrap_df, 
          lmerseq_df) |> 
  pivot_wider(names_from = source, 
              values_from = c(estimate, p.value)) |> 
 ggplot(aes(x = estimate_seqwrap, y = estimate_lmerseq, color = target)) +
   geom_point(size = 2) +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
   facet_wrap(~ term) +
   labs(subtitle = "Comparison of Estimates", x = "seqwrap Estimate", y = "lmerseq Estimate", 
        color = "") +
   theme_classic()



p2 <-  bind_rows(seqwrap_df, 
          lmerseq_df) |> 
  pivot_wider(names_from = source, 
              values_from = c(estimate, p.value)) |> 
   ggplot(aes(x = p.value_seqwrap, y = p.value_lmerseq, color = target)) +
     geom_point(size = 2) +
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
     facet_wrap(~ term) +
   labs(subtitle = "Comparison of p-values", x = "seqwrap p-values", y = "lmerseq p-values") +
     theme_classic() + 
  theme(legend.position = "none")


plot_grid(p1, p2, ncol = 1)

```


## Fitting the negative binomial and Poisson OLRE models using `seqwrap`

In the analyses presented in the main text, we used a negative binomial model as the point of departure for the analyses. The model was fitted in `seqwrap` using `glmmTMB` with the `nbinom2` family function. Below we show the code for fitting the model described in Eq. 1 (Negative binomial (non-informed)).

We used the TMM-method [@robinson2010] for calculating the effective library size before adding it to the metadata as log-transformed library sizes relative to the median-sized library. We are using the DHARMa package [@DHARMa] in a custom evaluation function (`sigma_summary2`), which also include additional metrics. 

```{r}
#| eval: false

### Code chunk not evaluated ###

# Preparing data 
# Download the data using seqwraphelper
# (see above for installation instructions)
dat <- seqwraphelper::geo_data(gse = "GSE202295")

# Filter low expressed genes and combine after filtering
keep <- filterByExpr(
  dat$countdata[,-1],
  min.count = 10,
  min.total.count = 15,
  large.n = 10,
  min.prop = 0.7,
  group = paste(dat$metadata$group, dat$metadata$time))



countdat <-  dat$countdata[keep,]
metadat <- dat$metadata


# Use EdgeR to calculate the TMM
y <- edgeR::DGEList(countdat[,-1])
y <- edgeR::calcNormFactors(y)

# Store effective library sizes 
libsize <- y$samples |>
    rownames_to_column(var = "seq_sample_id") |>
    select(- group)

# Combine all meta data
metadat <- dat$metadata |>
  inner_join(libsize, by = "seq_sample_id") |>
  mutate(group = factor(group, levels = c("NGT", "T2D")),
         time = factor(time, levels = c("basal", "post", "rec")),
         efflibsize = (lib.size * norm.factors)/median(lib.size * norm.factors),
         ln_efflibsize = log(efflibsize))

# Setting up the evaluation function
sigma_summary2 <- function(x) {

  if(is.null(x$fit$convergence)) {
    conv <- 1
  } else {
    conv <- x$fit$convergence[[1]]
  }


  ### Simulating scaled resid ###
  # This creates simulated residuals for test of uniformity and dispersion.
  # See
  # https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html
  # for details.
  residObj <- DHARMa::simulateResiduals(x, plot = FALSE)

  # Saving tests from DHARMa
  unif <- DHARMa::testUniformity(residObj, plot = FALSE)
  disp <- DHARMa::testDispersion(residObj, plot = FALSE)


  # Saving SD of Obs Level Rand Eff in the Poisson model

  tidy_coef <- broom.mixed::tidy(x)

  if (any(tidy_coef$group %in% "seq_sample_id")) {
    olre.sd <- tidy_coef |>
      dplyr::filter(group == "seq_sample_id") |>
      dplyr::pull(estimate)
  } else { olre.sd <- NA }



  # Combine all in a data frame
  out <- data.frame(dispersion =  data.frame(summary(x$sdr))["betadisp",1],
                    dispersion.se = data.frame(summary(x$sdr))["betadisp",2],
                    olre.sd = olre.sd,
                    log_mu = mean(predict(x, type = "link")),
                    convergence = conv,
                    pdHess = x$sdr$pdHess,
                    aic = AIC(x),
                    unif.p = unif$p.value,
                    unif.stat = unif$statistic,
                    disp.p = disp$p.value,
                    disp.stat = disp$statistic)
  return(out)

}




# Setting up the model
m1 <- seqwrap_compose(
  data = countdat,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = glmmTMB::glmmTMB,
  eval_fun = sigma_summary2,
  arguments = list(
    formula = y ~ time * group + offset(ln_efflibsize) + (1|id),
    family = glmmTMB::nbinom2)
  )

m1_results <- seqwrap(m1,
                      return_models = FALSE,
                      cores = parallel::detectCores())

```


The Poisson Observation-Level Random Effect (OLRE) model was fitted using `glmmTMB` using the `stats::poisson` family function. The same data and metadata as prepared above was used. The below code corresponds to Eq. 2 in the main text (Poisson OLRE (non-informed)). Notice that the sample id variable is used to introduce the observation-level random effect into the model.

```{r}
#| eval: false

### Code chunk not evaluated ###

m2 <- seqwrap_compose(
  data = countdat,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = glmmTMB::glmmTMB,
  eval_fun = sigma_summary2,
  targetdata = NULL,
  arguments = list(
    formula = y ~ time * group + 
      offset(ln_efflibsize) + 
      (1|id) +
      (1|seq_sample_id),
      family = stats::poisson)
  )


m2_results <- seqwrap(
      m2,
      return_models = FALSE,
      verbose = FALSE,
      subset = 1:16,
      cores = parallel::detectCores())

```


We regularized models by adding priors based on non-informed models. The general strategy was outlined above for the Dungan data example. Here follows the code for the negative binomial model. 

```{r}
#| eval: false

### Code not evaluated ###

## Summarise results from the non-informed model
m1_sums  <- seqwrap_summarise(m1_results)


# Prepare priors for the regularized version

  ## Fitting a model for the mean-dispersion relationship

  # Fit a trend to the dispersion data from m1
  # save the data in a convenient format. Using the log_mu_raw (average
  # observed counts) allows for a prior on dispersion also for genes with
  # unsuccessful fits in model 1. First we gather all dispersion data.

  dispersion_dat <- m1_sums$evaluation |>
    filter(target %in% targets)

  # Calculate the raw observed counts from the data
  raw_log_counts <- data.frame(
    target = as.character( countdat[, 1]),
    log_mu_raw = log(rowMeans(countdat[,-1]))
  )


  # Adding log raw counts to the dispersion df for modeling.
  dispersion_dat <- dispersion_dat |>
    inner_join(raw_log_counts)



  # The SE of the dispersion estimate is used as weights in the trended model
    trend_model <- loess(dispersion ~ log_mu_raw,
                         data = dispersion_dat,
                         span = 0.7,
                         weights = 1/(dispersion.se^2))


  # Predict dispersion for each gene based on log raw counts
  # and combine into a prior.

  dispersion_prior <- data.frame(gene =  countdat[,1],
                                 pred = round(
                                   predict(trend_model,
                                           newdata = data.frame(
                                             log_mu_raw =  log(
                                               rowMeans(
                                                 countdat[,-1]))
                                           )
                                   ), 3),
                                 s = round(
                                   trend_model$s, 3)
  ) |>
    mutate(prior = paste0("normal(", pred, ",", s, ")"))



  # Gather all distributions of estimates that can be used as prior information
  # in subsequent model.


  ## Preliminary plot ##
  #  ms1_sum$summaries |>
  #  filter(target %in% targets) |>
  #  select(target, term, estimate) |>
  #  ggplot(aes(estimate)) + geom_density() +
  #  facet_wrap(~ term, scales = "free")

  estimate_distributions <- m1_sums$summaries |>
    filter(target %in% targets) |>
    select(target, term, estimate) |>

    summarise(.by = term,
              m = mean(estimate),
              s = sd(estimate)) |>
    filter(!term %in%  c("(Intercept)", "sd__(Intercept)"))

  # Extract the random effects distribution to fit a gamma distribution
  random_sd_estimate <- m1_sums$summaries |>
    filter(target %in% targets,
           term == "sd__(Intercept)") |>
    select(target, term, estimate) |>
    pull(estimate)

  # The gamma distribution is parameterized using a shape and a rate
  # parameter. It looks like this prior will lead to a push towards 0,
  # we are adding a constant to push away from zero...
  mean_sd <- mean(random_sd_estimate)
  var_sd <- var(random_sd_estimate)
  shape_param <- 2 # mean_sd^2 / var_sd





  # Here we prepare priors for the fixed effects, in this version all fixed
  # effects, except the intercept, will have regularizing priors 
  # corresponding to the distributions of effects seen in the non-informed
  # models.
  Priors_df <- bind_rows(
    data.frame(prior = paste0("normal(",
                              round(estimate_distributions$m,2 ), 
                              ", ",
                              ,round(estimate_distributions$s,2 ), ")"),
               class = rep("fixef", 5),
               coef = estimate_distributions$term),
    data.frame(prior = paste0(
      "gamma(",
      round(mean_sd ,2),
      ",",
      2,
      ")"),
      class = "ranef",
      coef = "id")
  )

  # We want to use the mean-dispersion relationship to add a prior for the
  # dispersion parameter. This means that we need a target specific prior

  Priors_list <- list()
  for( j in 1:nrow( countdat)) {


    df <- bind_rows(Priors_df,
                    data.frame(
                      prior =
                        dispersion_prior[dispersion_prior$gene ==
                                           countdat[j,1],4],
                      class = "fixef_disp",
                      coef = "1"
                    )
    )

    Priors_list[[j]] <- df



  }

# Fit regularized model 
  
  m1.reg <- seqwrap_compose(
    data =  countdat,
    metadata =  metadat,
    samplename = "seq_sample_id",
    modelfun = glmmTMB::glmmTMB,
    eval_fun = sigma_summary2,
    targetdata =  Priors_list,
    arguments = alist(
      formula =  y ~ time * group + offset(ln_efflibsize) + (1|id),
      family = glmmTMB::nbinom2,
      priors = data.frame(
        prior = prior,
        class = class,
        coef = coef) ))

  m1.reg_results <- seqwrap(
    m1.reg,
    return_models = FALSE,
    verbose = FALSE,
    #  subset = 1:50, # for prototyping
    cores = parallel::detectCores())


```






## Simulation results

Below follows supplementary results from the simulation study. 

```{r}
#| label: fig-genecounts-simulated
#| fig-cap: "Number of simulated gene targets (true negative and true positives) across dispersion scenarios and sample sizes."
#| echo: false
#| message: false
#| warning: false


source("figures/figure-opts.R")

## Load data from simulations
source("R/simulation-functions.R")


# Extract simulations sorts out all simulations files and combines them.
# This is simulations with dispersion scenario 1.
sim_results <- extract_simulations(evaluations_path = "data_sim/evaluations",
                                   estimates_path = "data_sim/estimates",
                                   populationeffects_path = "data_sim/simdata/popeffect",
                                   disp_scenario = "s1")


sim_results2 <- extract_simulations(evaluations_path = "data_sim/evaluations2",
                                   estimates_path = "data_sim/estimates2",
                                   populationeffects_path = "data_sim/simdata2/popeffect",
                                   disp_scenario = "s2")


# Number of genes in each simulated data set


bind_rows(sim_results$populationeffects, 
          sim_results2$populationeffects) |> 
  # Filter to keep only 1 term
  filter(term == "conditionB") |> 
  mutate(effect = if_else(population_effect == 0, "true.negative", "true.positive")) |> 
  summarise(.by = c(dataset, size, disp_scenario, effect),
            n = n()) |>
  
  mutate(size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small (m = 54, n = 18)", 
                                  "Medium (m = 108, n = 36)", 
                                  "Large (m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         effect = factor(effect, levels = c("true.negative", "true.positive"), 
                         labels = c("True negative", "True positive"))) |> 
  
  ggplot(aes(size, n, fill = disp_scenario)) + 
  geom_point(position = position_jitter(width = 0.05), 
             size = 3, 
             shape = 21,
             alpha = 0.5) +
  labs(x = "Sample size (m samples, n participants)", 
       y = "Gene targets", 
       fill = "Dissersion scenario")  +
  theme_classic() + 
  
  
  
  scale_fill_manual(values = c(colors[c(1,5)])) +
  theme(legend.position = "bottom", 
        legend.title = element_blank(), 
        strip.background = element_blank(), 
        strip.text = element_text(hjust = 0)) + 
  facet_wrap(~ effect, scales = "free", ncol = 1)
```




```{r}
#| eval: false



# Number of genes per data set with true population effect
true_effects <- bind_rows(sim_results$populationeffects, 
          sim_results2$populationeffects) |> 
  mutate(effect = if_else(population_effect == 0, "true.negative", "true.positive")) |> 
  summarise(.by = c(term, dataset, size, disp_scenario, effect), 
            n = n()) 

# Savining for manuscript
saveRDS(true_effects, "data-out/true_effects.RDS")




# Combine all true/false effects
est_temp <- bind_rows(sim_results$estimates, 
          sim_results2$estimates) |> 
  # Retain only terms of interest.
  filter(term %in% c("conditionB", "timet3:conditionB")) |>
  # dataset was mis-named in the simulations.
  rename(dataset = datasets) |>
  select(- file)


est <- est_temp |> 
  inner_join(
    bind_rows(sim_results$populationeffects, 
              sim_results2$populationeffects) |>
               mutate(target = as.character(target)) |>
               select(-file)) |>

  select(target, term, estimate:p.value, population_effect, dataset, model, size, disp_scenario) |>

  mutate(.by = c(model, term, size, dataset, disp_scenario),
         fdr = p.adjust(p.value, method = "fdr")) |>
  mutate(true_effect = if_else(population_effect == 0, "neg", "pos"),
         identified_effect = if_else(fdr > 0.05, "neg", "pos"),
         true_positive = if_else(true_effect == "pos" &
                                   identified_effect == "pos", TRUE, FALSE),
         false_positive = if_else(true_effect == "neg" &
                                    identified_effect == "pos", TRUE, FALSE)) 


```



```{r}
#| eval: false



# 
convergence_stats <- bind_rows(sim_results$evaluations, 
          sim_results2$evaluations) |> 
  select(-file) |> 
  mutate(conv = if_else(model == "m3" & isSingular == TRUE, FALSE, 
                        if_else(model != "m3" & convergence != 0, FALSE, TRUE)), 
         conv.nonstrict = if_else(model != "m3" & convergence != 0, FALSE, TRUE)) |> 
  rename(dataset = datasets) |> 
  select(target, size, model, dataset, disp_scenario, isSingular, conv, conv.nonstrict) |> 

  inner_join(
    bind_rows(sim_results$populationeffects, 
              sim_results2$populationeffects) |>
               mutate(target = as.character(target)) |>
               select(-file) |> 
      mutate(effect = if_else(population_effect == 0, "n", "p")) |> 
      select(target, term, dataset, size, disp_scenario, effect) |> 
      pivot_wider(names_from = term, values_from = effect) |> 
      mutate(effect = paste0(conditionB, ":", `timet3:conditionB`)) |> 
      mutate(.by = c(dataset, size, effect, disp_scenario), 
             ntotal = n()) |> 
      select(-conditionB, - `timet3:conditionB`) 
  )  |> 
  
  summarise(.by = c(model, dataset, disp_scenario, effect, size), 
            conv = sum(conv),
            conv.nonstrict = sum(conv.nonstrict),
            ntotal = mean(ntotal)) 
  
  


  
saveRDS(convergence_stats, "data-out/covergence_stats.RDS" )


```


```{r}
#| eval: false
#| label: fig-convergence
#| fig-cap: "Number of models that provided estimates as a percentage of all potential targets. In (A), a strict criteria was used for convergence for the Gaussian model of transformed counts where models with singular fits where excluded as suggested by Vestal et al. (2022). In (B) all converged models are included."



p1 <- convergence_stats |> 

  mutate(success = 100 *  (conv / ntotal), 
         size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small (m = 54, n = 18)", 
                                  "Medium (m = 108, n = 36)", 
                                  "Large (m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"), 
                        labels = c("Negative binomial", 
                                   "Regularized Negative binomial", 
                                   "Gaussian transformed counts", 
                                   "Poisson OLRE", 
                                   "Regularized Poisson OLRE"))) |> 
  ggplot(aes(model, success, color = model)) +
  geom_point(position = position_jitter(width = 0.1)) + 
  facet_grid(disp_scenario ~ size) +
  
  labs(y = "Converged/non-singular models (%)", 
       x = "Model") +
  
  theme_classic() + 
  theme(panel.background = element_rect(fill = "gray95"), 
        strip.background = element_blank(), 
        strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(hjust = 0), 
        legend.title = element_blank(), 
        axis.text.x = element_blank()) + 
  scale_color_manual(values = colors)
  
  
  
p2 <-  convergence_stats |> 

  mutate(success = 100 *  (conv.nonstrict / ntotal), 
         size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small (m = 54, n = 18)", 
                                  "Medium (m = 108, n = 36)", 
                                  "Large (m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"), 
                        labels = c("Negative binomial", 
                                   "Regularized Negative binomial", 
                                   "Gaussian transformed counts", 
                                   "Poisson OLRE", 
                                   "Regularized Poisson OLRE"))) |> 
  ggplot(aes(model, success, color = model)) +
  geom_point(position = position_jitter(width = 0.1)) + 
  facet_grid(disp_scenario ~ size) +
  
  labs(y = "Converged models (%)", 
       x = "Model") +
  
  theme_classic() + 
  theme(panel.background = element_rect(fill = "gray95"), 
        strip.background = element_blank(), 
        strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(hjust = 0), 
        legend.title = element_blank(), 
        axis.text.x = element_blank(), 
        legend.position = "none")  + 
  scale_color_manual(values = colors)
  
plot_grid(p1, 
          plot_grid(p2, NULL, rel_widths = c(1, 0.32)), 
          ncol = 1)

```





```{r}
#| eval: false

# P-value distributions for null genes

est |>
  filter(population_effect == 0) |>
  
  mutate(model_coef = factor(paste0(model,"_", term), 
                             levels = c("m1_conditionB", 
                                        "m2_conditionB", 
                                        "m3_conditionB", 
                                        "m4_conditionB", 
                                        "m5_conditionB",
                                        "m1_timet3:conditionB", 
                                        "m2_timet3:conditionB",
                                        "m3_timet3:conditionB",
                                        "m4_timet3:conditionB",
                                        "m5_timet3:conditionB"
                                        ), 
                             labels = c("Model 1\nMain", 
                                        "Model 2\nMain",
                                        "Model 3\nMain",
                                        "Model 4\nMain",
                                        "Model 5\nMain",
                                        "Model 1\nInteraction",
                                        "Model 2\nInteraction",
                                        "Model 3\nInteraction",
                                        "Model 4\nInteraction",
                                        "Model 5\nInteraction"
                                        )), 
         size_disp = factor(paste0(size,"_", disp_scenario), 
                            levels = c("small_s1", "medium_s1", "large_s1", 
                                       "small_s2", "medium_s2", "large_s2"), 
                            labels = c("Small\nHigh dispersion", 
                                       "Medium\nHigh dispersion", 
                                       "Large\nHigh dispersion", 
                                       "Small\nLow dispersion", 
                                       "Medium\nLow dispersion", 
                                       "Large\nLow dispersion"
                                       ))) |> 
  
  ggplot(aes(p.value)) +
  geom_histogram(color="black", fill="grey40", binwidth = 0.025,
                 boundary = 0, closed = "left") +
  scale_x_continuous(breaks = c(0.05, 0.5, 1)) +
  facet_grid(model_coef ~ size_disp) + 
  theme_classic() + 
  labs(x = "Un-adjusted P-value", 
       y = "")









```




```{r}
#| eval: false


## Create averages

```




```{r}
#| eval: false





```




```{r}
#| eval: false

est |>

  filter(population_effect != 0) |>

  mutate(.by = c(model, size, term, dataset, disp_scenario),
         rank_true = rank(population_effect),
         rank_obs = rank(estimate)) |>

  summarise(.by = c(model, size, term, dataset, disp_scenario),
            cor = cor(rank_true, rank_obs)) |>

  mutate(size = factor(size, levels = c("small", "medium", "large"))) |>
  ggplot(aes(size, cor, group = paste(model,dataset),
             color = model)) +
  geom_line() +
  facet_grid(term ~ disp_scenario)



```



```{r}
#| eval: false
est |>

  filter(population_effect != 0) |>
  filter(dataset == 2, size == "medium") |>

  mutate(.by = c(model, term, dataset, disp_scenario),
         rank_true = rank(population_effect),
         rank_obs = rank(estimate)) |>

  ggplot(aes(rank_true, rank_obs)) +
  geom_point(alpha = 0.3) +
  facet_grid(paste(model) ~ paste(term, disp_scenario))




```


## Analyzing real-world data [@pillon2022]


```{r}
#| eval: false

models_data  <-  readRDS("data-out/pillon-models.RDS")


### Count number of successful models 
summaries <- lapply(models_data[1:5], seqwrap_summarise, verbose = FALSE)


targets_models <- bind_rows(
  summaries$m1_results$evaluations |> 
  mutate(model = "m1"), 
  summaries$m2_results$evaluations |> 
  mutate(model = "m2"), 
  summaries$m3_results$evaluations |> 
  mutate(model = "m3"), 
  summaries$m4_results$evaluations |> 
  mutate(model = "m4"), 
  summaries$m5_results$evaluations |> 
  mutate(model = "m5") ) |> 
  
  # Filter singular fits and bad convergence
  filter(!(model == "m3" & isSingular == TRUE)) |> 
  # Poisson and NB models do not have pdHess == FALSE or convergence != 0
  # but for failsafe
  filter(!(model %in% c("m1", "m2", "m4", "m5") & 
           (pdHess == FALSE | convergence != 0))) |> 
  group_by(model) |> 
  distinct(target)

## Sums number of successful targets  
targets_models |> 
  ungroup() |> 
  summarise(.by = model, 
            n = n()) |> 
  # Add total 
  mutate(total = nrow(models_data$filtered_counts), 
         estimated_prop = n / total) 
  



## Summaries
bind_rows(
  summaries$m1_results$summaries |> 
  mutate(model = "m1"), 
  summaries$m2_results$summaries |> 
  mutate(model = "m2"), 
  summaries$m3_results$summaries |> 
  mutate(model = "m3"), 
  summaries$m4_results$summaries |> 
  mutate(model = "m4"), 
  summaries$m5_results$summaries |> 
  mutate(model = "m5") ) |> 
  
  # This filter away singular fits from m3 
  inner_join(targets_models) |> 
  
  filter(term %in% c("groupT2D", "timepost:groupT2D", "timerec:groupT2D")) |> 
  
  # Adjust p-values and calculate MSD
  mutate(.by = c(model, term), 
         fdr = p.adjust(p.value, method = "fdr")) |> 
  # Approximate CI
  mutate(lwr = estimate - qnorm(0.975) * std.error, 
         upr = estimate + qnorm(0.972) * std.error, 
         msd = if_else(estimate > 0, lwr, -upr)) |> 

  
  print()
  

  
  
summaries$m1_results$summaries








```


