---
title: "seqwrap: an R package for flexible iterative fitting ofhigh dimensional data"
subtitle: "Supplementary material"
author:
  - name: Chidimma Phoebe Echebiri
    affiliation:
      - name: University of Inland Norway
        department: Department of Biotechnology, PhD programme in applied ecology and biotechnology
  - name: Rafi Ahmad
    affiliation: 
      - name: University of Inland Norway
        department: Department of Biotechnology
  - name: Stian Ellefsen
    affiliation: 
      - name: University of Inland Norway
        department: Department of Public Health and Sport Sciences 
  - name: Daniel Hammarström
    email: daniel.hammarstrom@inn.no
    affiliation: 
      - name: University of Inland Norway
        department: Department of Public Health and Sport Sciences 
    attributes:
      corresponding: true
format:
  typst:
    toc: true
csl: resources/bmc-bioinformatics.csl
---

## Fitting models with `seqwrap`

### Using target data for prior information in `glmmTMB` 


## Comparing `glmmSeq`, `lmerSeq` and `seqwrap`

Here, a basic comparison is made between lmerseq [@vestal2022] and Seqwrap using the [@pillon2022] gene expression data

### Data extraction/preprocessing

Here the Pillion gene expression data is downloaded and transformed using VST as required by lmerseq

```{r}
#| label: libraries and data preprocessing
#| code-fold: true
#| echo: false
library(lmerSeq)
library(DESeq2)
library(edgeR)
library(seqwraphelper)
library(dplyr)
library(tidyverse)
library(seqwrap)
library(purrr)
library(glmmSeq)
library(ggplot2)

# download the data from seqwraphelper
dat <- seqwraphelper::geo_data(gse = "GSE202295")




# Filter low expressed gened and combine after filtering
keep <- filterByExpr(
  dat$countdata[,-1],
  min.count = 10,
  min.total.count = 15,
  large.n = 10,
  min.prop = 0.7,
  group = paste(dat$metadata$group, dat$metadata$time))

countdat <-  dat$countdata[keep,]



metadat <- dat$metadata



# Check if sequence_ID matches. If False, make to match 
all(colnames(countdat)[-1] == metadat$seq_sample_id)



# Normalize the count data using DESeq2's VST
dds <- DESeqDataSetFromMatrix(countData = countdat[,-1],
                              colData = metadat,
                              design = ~ time * group)

dds <- DESeq(dds)
vsd.fixed <- varianceStabilizingTransformation(dds, blind=F)
vst_expr <- assay(vsd.fixed)





# Add as gene_ids to the variance_stabilised_counts
rownames(vst_expr) <- countdat$geneid

# subset to only the first fifteen genes
vst_expr <- vst_expr[1:15,]
```

Lmerseq has a lmerseq.fit() function that builds the model. Lmerseq by default uses 2 cores. This will be set in the other packages too

```{r}
#| label: lmerseq
#| code-fold: true
#| echo: false

# fit the model using lmerseq
fit.lmerSeq <- lmerSeq.fit(form = ~ time * group  + (1|id),
                           expr_mat = vst_expr,
                           sample_data = metadat,
                           REML = T)




coefs <- c("timepost", "timerec", "groupT2D", "timepost:groupT2D", "timerec:groupT2D")


lmerseq_summary <- coefs %>%
  map_df(~ lmerSeq.summary(
    lmerSeq_results = fit.lmerSeq,
    coefficient = .x,
    p_adj_method = "BH",
    ddf = "Satterthwaite",
    sort_results = FALSE
  )$summary_table %>%
    mutate(Coefficient = .x)
  )


saveRDS(lmerseq_summary, "data/lmerseq_model_summary.RDS")
```

### Seqwrap

Building same model with seqwrap

```{r}
#| label: Seqwrap
#| code-fold: true
#| echo: false

# make data suitable for seqwrap as it accepts only dataframes or lists
vst_expr_df <- data.frame(vst_expr) %>%
  tibble::rownames_to_column(var = "target")


# first validate input using seqwrap compose
seqwrap_val <- seqwrap_compose(
  data = vst_expr_df,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = lmerTest:::lmer,
  arguments = list(
    formula = y ~ time * group  + (1|id))
)


  seqwrap_model <- seqwrap(seqwrap_val,
                        return_models = FALSE,
                        cores = 2)

seqwrap_summary_df <- seqwrap_summarise(seqwrap_model)$summaries


saveRDS(seqwrap_summary_df, "data/seqwrap_model_summary.RDS")
```

### Visualisation

The estimates and pvalues of both models are to be visualised

```{r}
#| label: Visualisation
#| code-fold: true
#| echo: false


# Select
seqwrap_df <- seqwrap_summary_df %>%
  subset(!term %in% c("(Intercept)", "sd__(Intercept)", "sd__Observation")) %>%
  dplyr::select("target",  "estimate", "p.value", "term" ) %>%
  mutate(source = "Seqwrap")


lmerseq_df <- lmerseq_summary %>%
  dplyr::select(gene, Estimate, p_val_raw, Coefficient) %>%
  mutate(source = "lmerseq") %>%
  #rename to match the columns from seqwrap summary
  rename(target = gene,
         estimate = Estimate,
         p.value = p_val_raw,
         term = Coefficient)


merged <- merge(seqwrap_df, lmerseq_df, by = c("target", "term"), suffixes = c("_seqwrap", "_lmerseq"))

ggplot(merged, aes(x = estimate_seqwrap, y = estimate_lmerseq, color = target)) +
   geom_point(size = 2) +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
   facet_wrap(~ term)
   labs(title = "Comparison of Estimates", x = "seqwrap Estimate", y = "lmerseq Estimate") +
   theme_minimal()





   
   
   ggplot(merged, aes(x = p.value_seqwrap, y = p.value_lmerseq, color = target)) +
     geom_point(size = 2) +
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
     facet_wrap(~ term)
   labs(title = "Comparison of p values", x = "seqwrap p values", y = "lmerseq pvalues") +
     theme_minimal()


```



## Simulation results

```{r}
#| label: fig-genecounts-simulated
#| fig-cap: "Number of simulated gene targets (true negative and true positives) across dispersion scenarios and sample sizes."

library(tidyverse)
library(cowplot)
library(ggtext)
source("figures/figure-opts.R")

## Load data from simulations
source("R/simulation-functions.R")


# Extract simulations sorts out all simulations files and combines them.
# This is simulations with dispersion scenario 1.
sim_results <- extract_simulations(evaluations_path = "data_sim/evaluations",
                                   estimates_path = "data_sim/estimates",
                                   populationeffects_path = "data_sim/simdata/popeffect",
                                   disp_scenario = "s1")


sim_results2 <- extract_simulations(evaluations_path = "data_sim/evaluations2",
                                   estimates_path = "data_sim/estimates2",
                                   populationeffects_path = "data_sim/simdata2/popeffect",
                                   disp_scenario = "s2")


# Number of genes in each simulated data set


bind_rows(sim_results$populationeffects, 
          sim_results2$populationeffects) |> 
  # Filter to keep only 1 term
  filter(term == "conditionB") |> 
  mutate(effect = if_else(population_effect == 0, "true.negative", "true.positive")) |> 
  summarise(.by = c(dataset, size, disp_scenario, effect),
            n = n()) |>
  
  mutate(size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small (m = 54, n = 18)", 
                                  "Medium (m = 108, n = 36)", 
                                  "Large (m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         effect = factor(effect, levels = c("true.negative", "true.positive"), 
                         labels = c("True negative", "True positive"))) |> 
  
  ggplot(aes(size, n, fill = disp_scenario)) + 
  geom_point(position = position_jitter(width = 0.05), 
             size = 3, 
             shape = 21,
             alpha = 0.5) +
  labs(x = "Sample size (m samples, n participants)", 
       y = "Gene targets", 
       fill = "Dissersion scenario")  +
  theme_classic() + 
  
  
  
  scale_fill_manual(values = c(colors[c(1,5)])) +
  theme(legend.position = "bottom", 
        legend.title = element_blank(), 
        strip.background = element_blank(), 
        strip.text = element_text(hjust = 0)) + 
  facet_wrap(~ effect, scales = "free", ncol = 1)
```






```{r}

# Number of genes per data set with true population effect
true_effects <- bind_rows(sim_results$populationeffects, 
          sim_results2$populationeffects) |> 
  mutate(effect = if_else(population_effect == 0, "true.negative", "true.positive")) |> 
  summarise(.by = c(term, dataset, size, disp_scenario, effect), 
            n = n()) 

# Savining for manuscript
saveRDS(true_effects, "data-out/true_effects.RDS")




# Combine all true/false effects
est_temp <- bind_rows(sim_results$estimates, 
          sim_results2$estimates) |> 
  # Retain only terms of interest.
  filter(term %in% c("conditionB", "timet3:conditionB")) |>
  # dataset was mis-named in the simulations.
  rename(dataset = datasets) |>
  select(- file)


est <- est_temp |> 
  inner_join(
    bind_rows(sim_results$populationeffects, 
              sim_results2$populationeffects) |>
               mutate(target = as.character(target)) |>
               select(-file)) |>

  select(target, term, estimate:p.value, population_effect, dataset, model, size, disp_scenario) |>

  mutate(.by = c(model, term, size, dataset, disp_scenario),
         fdr = p.adjust(p.value, method = "fdr")) |>
  mutate(true_effect = if_else(population_effect == 0, "neg", "pos"),
         identified_effect = if_else(fdr > 0.05, "neg", "pos"),
         true_positive = if_else(true_effect == "pos" &
                                   identified_effect == "pos", TRUE, FALSE),
         false_positive = if_else(true_effect == "neg" &
                                    identified_effect == "pos", TRUE, FALSE)) 


```



```{r}
#| 



# 
convergence_stats <- bind_rows(sim_results$evaluations, 
          sim_results2$evaluations) |> 
  select(-file) |> 
  mutate(conv = if_else(model == "m3" & isSingular == TRUE, FALSE, 
                        if_else(model != "m3" & convergence != 0, FALSE, TRUE)), 
         conv.nonstrict = if_else(model != "m3" & convergence != 0, FALSE, TRUE)) |> 
  rename(dataset = datasets) |> 
  select(target, size, model, dataset, disp_scenario, isSingular, conv, conv.nonstrict) |> 

  inner_join(
    bind_rows(sim_results$populationeffects, 
              sim_results2$populationeffects) |>
               mutate(target = as.character(target)) |>
               select(-file) |> 
      mutate(effect = if_else(population_effect == 0, "n", "p")) |> 
      select(target, term, dataset, size, disp_scenario, effect) |> 
      pivot_wider(names_from = term, values_from = effect) |> 
      mutate(effect = paste0(conditionB, ":", `timet3:conditionB`)) |> 
      mutate(.by = c(dataset, size, effect, disp_scenario), 
             ntotal = n()) |> 
      select(-conditionB, - `timet3:conditionB`) 
  )  |> 
  
  summarise(.by = c(model, dataset, disp_scenario, effect, size), 
            conv = sum(conv),
            conv.nonstrict = sum(conv.nonstrict),
            ntotal = mean(ntotal)) 
  
  


  
saveRDS(convergence_stats, "data-out/covergence_stats.RDS" )


```


```{r}
#| label: fig-convergence
#| fig-cap: "Number of models that provided estimates as a percentage of all potential targets. In (A), a strict criteria was used for convergence for the Gaussian model of transformed counts where models with singular fits where excluded as suggested by Vestal et al. (2022). In (B) all converged models are included."



p1 <- convergence_stats |> 

  mutate(success = 100 *  (conv / ntotal), 
         size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small (m = 54, n = 18)", 
                                  "Medium (m = 108, n = 36)", 
                                  "Large (m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"), 
                        labels = c("Negative binomial", 
                                   "Regularized Negative binomial", 
                                   "Gaussian transformed counts", 
                                   "Poisson OLRE", 
                                   "Regularized Poisson OLRE"))) |> 
  ggplot(aes(model, success, color = model)) +
  geom_point(position = position_jitter(width = 0.1)) + 
  facet_grid(disp_scenario ~ size) +
  
  labs(y = "Converged/non-singular models (%)", 
       x = "Model") +
  
  theme_classic() + 
  theme(panel.background = element_rect(fill = "gray95"), 
        strip.background = element_blank(), 
        strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(hjust = 0), 
        legend.title = element_blank(), 
        axis.text.x = element_blank()) + 
  scale_color_manual(values = colors)
  
  
  
p2 <-  convergence_stats |> 

  mutate(success = 100 *  (conv.nonstrict / ntotal), 
         size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small (m = 54, n = 18)", 
                                  "Medium (m = 108, n = 36)", 
                                  "Large (m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"), 
                        labels = c("Negative binomial", 
                                   "Regularized Negative binomial", 
                                   "Gaussian transformed counts", 
                                   "Poisson OLRE", 
                                   "Regularized Poisson OLRE"))) |> 
  ggplot(aes(model, success, color = model)) +
  geom_point(position = position_jitter(width = 0.1)) + 
  facet_grid(disp_scenario ~ size) +
  
  labs(y = "Converged models (%)", 
       x = "Model") +
  
  theme_classic() + 
  theme(panel.background = element_rect(fill = "gray95"), 
        strip.background = element_blank(), 
        strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(hjust = 0), 
        legend.title = element_blank(), 
        axis.text.x = element_blank(), 
        legend.position = "none")  + 
  scale_color_manual(values = colors)
  
plot_grid(p1, 
          plot_grid(p2, NULL, rel_widths = c(1, 0.32)), 
          ncol = 1)

```





```{r}


# P-value distributions for null genes

est |>
  filter(population_effect == 0) |>
  
  mutate(model_coef = factor(paste0(model,"_", term), 
                             levels = c("m1_conditionB", 
                                        "m2_conditionB", 
                                        "m3_conditionB", 
                                        "m4_conditionB", 
                                        "m5_conditionB",
                                        "m1_timet3:conditionB", 
                                        "m2_timet3:conditionB",
                                        "m3_timet3:conditionB",
                                        "m4_timet3:conditionB",
                                        "m5_timet3:conditionB"
                                        ), 
                             labels = c("Model 1\nMain", 
                                        "Model 2\nMain",
                                        "Model 3\nMain",
                                        "Model 4\nMain",
                                        "Model 5\nMain",
                                        "Model 1\nInteraction",
                                        "Model 2\nInteraction",
                                        "Model 3\nInteraction",
                                        "Model 4\nInteraction",
                                        "Model 5\nInteraction"
                                        )), 
         size_disp = factor(paste0(size,"_", disp_scenario), 
                            levels = c("small_s1", "medium_s1", "large_s1", 
                                       "small_s2", "medium_s2", "large_s2"), 
                            labels = c("Small\nHigh dispersion", 
                                       "Medium\nHigh dispersion", 
                                       "Large\nHigh dispersion", 
                                       "Small\nLow dispersion", 
                                       "Medium\nLow dispersion", 
                                       "Large\nLow dispersion"
                                       ))) |> 
  
  ggplot(aes(p.value)) +
  geom_histogram(color="black", fill="grey40", binwidth = 0.025,
                 boundary = 0, closed = "left") +
  scale_x_continuous(breaks = c(0.05, 0.5, 1)) +
  facet_grid(model_coef ~ size_disp) + 
  theme_classic() + 
  labs(x = "Un-adjusted P-value", 
       y = "")









```




```{r}



## Create averages

```




```{r}






```




```{r}


est |>

  filter(population_effect != 0) |>

  mutate(.by = c(model, size, term, dataset, disp_scenario),
         rank_true = rank(population_effect),
         rank_obs = rank(estimate)) |>

  summarise(.by = c(model, size, term, dataset, disp_scenario),
            cor = cor(rank_true, rank_obs)) |>

  mutate(size = factor(size, levels = c("small", "medium", "large"))) |>
  ggplot(aes(size, cor, group = paste(model,dataset),
             color = model)) +
  geom_line() +
  facet_grid(term ~ disp_scenario)



```



```{r}

est |>

  filter(population_effect != 0) |>
  filter(dataset == 2, size == "medium") |>

  mutate(.by = c(model, term, dataset, disp_scenario),
         rank_true = rank(population_effect),
         rank_obs = rank(estimate)) |>

  ggplot(aes(rank_true, rank_obs)) +
  geom_point(alpha = 0.3) +
  facet_grid(paste(model) ~ paste(term, disp_scenario))




```




