---
title: "Case study: DNA methylation (GSE114763)"
format: html
bibliography: "resources/bibliography.bib"
csl: "resources/bmc-bioinformatics.csl"
editor_options: 
  chunk_output_type: console
---

## Introduction 

The following document outlines a partial reanalysis of data on CpG methylation status in human skeletal muscle in response to resistance exercise training, followed by detraining and subsequent retraining [@seaborne2018;@seaborne2018b]. Most of the relevant code is explicitly shown here, whereas the version of this case study included in the manuscript has been stripped of some details.


## Raw data and pre-processing

```{r}
#| label: dependencies
#| eval: true

library(GEOquery)      # For accessing the raw data
library(minfi)         # For pre-processing of raw data
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(maxprobes)     # For filtering cross-reactive probes 
library(seqwrap)


# Load color scale 
source(here::here("analysis/figures/figure-opts.R"))


```


Raw data are downloaded from the [Gene Expression Omnibus](https://www.ncbi.nlm.nih.gov/geo/) using the accession number [GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763). 


```{r}
#| label: paths
#| eval: true

# Prepare data download. 
# Note for use in the present repo: 
# raw_data and derived data are gitignored so safe for 
# large files. If code-re-use, check the download folder before committing 
# changes to git. 

# The GEO ID
gse_id   <- "GSE114763"

# Set download directories
raw_dir  <- here::here("analysis/data/raw_data", gse_id)
idat_dir <- file.path(raw_dir, "idat")
der_dir  <- here::here("analysis/data/derived_data")

dir.create(idat_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(der_dir,  recursive = TRUE, showWarnings = FALSE)

```

### Downloading data

Sample titles and characteristics are read from the SOFT text view of the series, which returns the metadata of all 40 samples. The supplementary archive (`GSE114763_RAW.tar`, ~722 MB) provides the IDAT files.

```{r}
#| label: download
#| eval: true


# The metadata (pheno_raw) and the raw data IDAT files are cached after 
# downloading once. 

# A function for retrying connection to GEO for downloading large data.
# This was created after issues with downloading the large zipped archive
# from GEO. 
retry_url <- function(expr, tries = 5, pause = 5) {
  call <- substitute(expr)          # re-evaluated on each attempt
  env  <- parent.frame()
  for (i in seq_len(tries)) {
    out <- try(eval(call, env), silent = TRUE)
    if (!inherits(out, "try-error")) return(out)
    if (i < tries) Sys.sleep(pause * i)
  }
  stop("failed after ", tries, " attempts: ",
       conditionMessage(attr(out, "condition")))
}

# Series-level metadata
#
# The metadata contains file names and titles include 
# time-points and participant information. 
pheno_file <- file.path(raw_dir, "pheno_raw.rds")

if (!file.exists(pheno_file)) {
  soft_url <- paste0("https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi",
                     "?acc=", gse_id, "&targ=gsm&form=text&view=brief")

  soft <- retry_url({
    res <- curl::curl_fetch_memory(soft_url)
    if (res$status_code != 200) stop("HTTP ", res$status_code)
    strsplit(rawToChar(res$content), "\r?\n")[[1]]
  })

  # one record per sample; "!Sample_<key> = <value>" lines within each
  kv <- tibble::tibble(record = cumsum(startsWith(soft, "^SAMPLE")),
                       line   = soft) |>
    dplyr::filter(startsWith(line, "!")) |>
    tidyr::separate(line, into = c("key", "value"), sep = " = ",
                    extra = "merge") |>
    dplyr::mutate(key = stringr::str_remove(key, "^!Sample_"))

  # characteristics are themselves "<key>: <value>" pairs
  characteristics <- kv |>
    dplyr::filter(key == "characteristics_ch1") |>
    tidyr::separate(value, into = c("key", "value"), sep = ": ",
                    extra = "merge") |>
    dplyr::filter(key %in% c("gender", "age", "weight", "height")) 

  # Cleaning the metadata for downstream processing. 
  pheno_raw <- kv |>
    dplyr::filter(key %in% c("geo_accession", "title")) |>
    dplyr::bind_rows(characteristics) |>
    tidyr::pivot_wider(id_cols = record, names_from = key,
                       values_from = value) |>
    dplyr::select(-record) |> 

        mutate(title_sep = title) |> 
        separate(title_sep, into = c("seq_sample_id", "timeid"), sep = ":") |>
        mutate(timeid = gsub(" Muscle_", "", timeid)) |> 
        mutate(timeid = gsub("Participant_", "", timeid)) |> 
        mutate(timeid = gsub("7wk_", "7wk", timeid)) |> 
        
         separate(timeid, into = c("time", "participant", "replicate"), sep = "_") |> 
      # Making time-point more obvious
        mutate(time = case_when(
              time == "Baseline" ~ "1_baseline", 
              time == "Acute" ~ "2_acute", 
              time == "7wkLoading" ~ "3_loading", 
              time == "7wkUnloading"  ~ "4_unloading", 
              time == "7wkRe-oading" ~ "5_reloading"
        )) |> 
        # Fixing meta-data details
        mutate(weight = gsub(" kg", "", weight), 
               height = gsub("cm", "", height), 
               weight = as.numeric(weight), 
               height = as.numeric(height), 
               age = as.numeric(age)) |> 
        # Fix a misprinted participant id
        mutate(participant = gsub("P006", "006", participant)) 
  

  saveRDS(pheno_raw, pheno_file)
}

pheno_raw <- readRDS(pheno_file)

# Double check 40 expected samples
stopifnot(nrow(pheno_raw) == 40)

# Download and extract raw IDAT files.
# IDATs are distributed gzipped; minfi reads decompressed files
tar_file <- file.path(raw_dir, paste0(gse_id, "_RAW.tar"))

if (!file.exists(tar_file)) {
  # ~722 MB; increase the timeout for slow connections
  options(timeout = 60 * 60)
  retry_url(GEOquery::getGEOSuppFiles(gse_id, baseDir = dirname(raw_dir),
                                      makeDirectory = TRUE))
}

if (length(list.files(idat_dir, pattern = "\\.idat$")) == 0) {
  utils::untar(tar_file, exdir = idat_dir)

  # Unzip idat files
  gz <- list.files(idat_dir, pattern = "\\.idat\\.gz$", full.names = TRUE)
  invisible(lapply(gz, R.utils::gunzip, overwrite = TRUE, remove = TRUE))
}

```

### Metadata summary

A total of 40 samples were downloaded from the Gene Expression Omnibus ([GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763)). The identified metadata (@tbl-meta-samples) corresponds to the metadata published in [@seaborne2018b] (see table 2 in @seaborne2018b).

Seaborne et al. removed two samples after preliminary analysis, (samples `SkM_Epi_Mem_1` and `SkM_Epi_Mem_39`), we will do the same.


```{r}
#| label: tbl-meta-samples
#| tbl-cap: "Sample metadata retreived from GEO ([GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763)) corresponding to @seaborne2018b".
#| echo: false


library(gt)

pheno_raw |> 
      summarise(.by = participant, 
                weight = mean(weight), 
                height = mean(height), 
                age = mean(age)) |> 
      mutate(participant = gsub("00", "", participant)) |> 
      select(participant, age, height, weight) |> 
      gt() |> 
      cols_label(participant = "Participant", 
                 age = "Age (years)", 
                 weight = "Weight (kg)", 
                 height = "Height (cm)")
      
      
```


## Case study description

```{r}
#| label: case-description
#| echo: false

# Double check gender 
# pheno_raw |> 
#      distinct(gender)
# Confirm all males


# Get sample metadata into text
sample_metadata <- pheno_raw |> 
      
      summarise(.by = participant, 
                weight = mean(weight), 
                height = mean(height), 
                age = mean(age)) |> 
      pivot_longer(weight:age) |> 
      summarise(.by = name, 
                m = mean(value), 
                s = sd(value)) |> 
      mutate(unit = c("kg", "cm", "years")) |> 
      mutate(stat = paste0(sprintf("%.1f", round(m, 1)), 
                           " (", 
                           sprintf("%.1f", round(s, 1)), 
                           ") ", 
                           unit)) 

age <- filter(sample_metadata, name == "age") |> pull(stat)
height <- filter(sample_metadata, name == "height") |> pull(stat)
weight <- filter(sample_metadata, name == "weight") |> pull(stat)

```


> Seaborne et al. [@seaborne2018;@seaborne2018b] investigated whether human skeletal muscle retains an epigenetic memory of earlier hypertrophy-inducing resistance training. For this aim, previously untrained male participants (age, [mean (SD)] `r age`; weight, `r weight`; height, `r height`) completed a sequential design with two seven-week periods of resistance-exercise training (RET) intersected with a 7-week unloading period without RET. RET periods included two weekly sessions targeting lower limbs (seven exercises) and one weekly session targeting the upper body (5 exercises; 4 sets of 10 repetitions per exercise). Skeletal muscle biopsies were sampled from m. vastus lateralis on five occasions: before the intervention, 30 min after the first exercise session, and after the completion of each of the 7-week periods. Genomic DNA was extracted from muscle biopsy samples (DNeasy Kit, Qiagen, Manchester, UK) and bisulfite converted before amplification, fragmentation, and hybridization using the Illumina Infinium Human MethylationEPIC Beadchip protocol [@seaborne2018;@seaborne2018b]. 


The Sentrix slide (chip) and array position are recovered from the IDAT file names (e.g. `GSM3149860_201496860025_R01C01_Grn.idat`) and retained as potential technical covariates in subsequent analyses. `Basename` is the file stem that `minfi` uses to locate the red/green channel pair.

The samples are distributed as described in [@seaborne2018;@seaborne2018b] with one participant dropping out from the study prior to the last sampling time-point. One participant have duplicate arrays at one time-point. After removal of samples based on preliminary QC-analysis (see @seaborne2018b), the full data set contains 38 samples with no technical duplicates. 

```{r}
#| label: tbl-metadata-count
#| eval: true
#| tbl-cap: "Samples per time-point and participant."


idat_stems <- list.files(idat_dir, pattern = "_Grn\\.idat$") |>
  stringr::str_remove("_Grn\\.idat$")

array_info <- tibble::tibble(stem = idat_stems) |>
  tidyr::separate(stem, into = c("geo_accession", "slide", "array"),
                  sep = "_", remove = FALSE) |>
  dplyr::mutate(Basename = file.path(idat_dir, stem))

metadata <- pheno_raw |>
      # Removing outlier sample, see "Quality Control of DNA Methylation Probes"
      # in @seaborne2018b
      filter(!grepl("SkM_Epi_Mem_1:", title), 
             !grepl("SkM_Epi_Mem_39:", title)) |> 
      
  tibble::as_tibble() |>
  dplyr::select(geo_accession,
                title,
                participant,
                gender,
                age, 
                time) |>
  dplyr::mutate(
    label = stringr::str_remove(title, "^SkM_Epi_Mem_\\d+:\\s*")) |>
  dplyr::left_join(array_info, by = "geo_accession")

stopifnot(!anyNA(metadata$time), !anyNA(metadata$participant))

# 38 arrays: 8 participants x 5 time points with missing values for on 
# participant.

metadata |> 
      summarise(.by = c(time, participant), 
                n = n()) |> 
      pivot_wider(names_from = time, values_from = n) |> 
      select(Participant = participant, 
             Baseline = `1_baseline`, 
             Acute = `2_acute`, 
             Loading = `3_loading`, 
             Unloading = `4_unloading`, 
             Reloading = `5_reloading`) |> 
      gt() |> 
      tab_spanner(label = "Sample time-points", columns = "Baseline":"Reloading")
      



```

### Reading IDAT and array-level quality control

We will use the `minfi` package [@fortin2014] to read the raw data and perform quality control, filtering and normalization. As a first step, array-level filtering is done using detection p-values where methylated and unmethylated positions are compared to negative control positions. This comparison produces a p-value used to identify problematic genomic positions. As shown below (@fig-array-qc), array-level aggregates show mean values < 0.01.    


```{r}
#| label: fig-array-qc
#| fig-cap: "Array-level quality control. Mean detection p-values and fraction of failed genomic positions per array subdivided by time-point."
#| eval: true

rgset <- minfi::read.metharray(metadata$Basename, extended = TRUE, force = TRUE)

# Check that orders correspond
stopifnot(identical(colnames(rgset), basename(metadata$Basename)))

colnames(rgset) <- metadata$geo_accession


det_p <- minfi::detectionP(rgset)

# 
qc <- tibble::tibble(
  geo_accession = colnames(det_p),
  mean_det_p    = colMeans(det_p),
  frac_failed   = colMeans(det_p > 0.01))

qc |> 
      inner_join(pheno_raw |> 
                       select(geo_accession, time)) |> 
      ggplot(aes(mean_det_p, frac_failed, color = time)) + 
      geom_point() +
      labs(x = "Array-level mean detection p-value", 
           y = "Fraction of failed genomic positions (p > 0.01)", 
           color = "Time-point")
```



```{r}

# Below is the suggested workflow for confirming sex. We expect only males and 
# the function indficate that only one sex is present which should suffice as 
# confirmation.
# predicted   <- minfi::getSex(minfi::mapToGenome(msetraw))
# qc$pred_sex <- predicted$predictedSex

# All samples pass quality control
keep_samples <- qc$geo_accession[qc$mean_det_p < 0.01]
# length(keep_samples)

rgset    <- rgset[, keep_samples]
det_p    <- det_p[, keep_samples]
metadata <- metadata |> dplyr::filter(geo_accession %in% keep_samples)
```

### Normalization and probe filtering

Functional normalization [@fortin2014] is used because the study design can be expected to result in global methylation differences. Functional normalization removes technical variation captured by the array control probes without assuming a common methylation distribution across samples [@fortin2014]. Probes are then removed if they (1) fail detection in any retained sample, (2) overlap common SNPs at the target CpG or single-base extension site, (3) appear on published lists of cross-reactive EPIC probes (using the [`maxprobes`](https://github.com/markgene/maxprobes/) package), or (iv) map to the sex chromosomes (all male cohort).

```{r}
#| label: normalise-filter
#| eval: true



# Functional normalization
gset <- minfi::preprocessFunnorm(rgset, ratioConvert = FALSE)

# Initial number of probes
nprobes_init <- dim(gset)[1]

## (i) detection
det_p  <- det_p[match(minfi::featureNames(gset), rownames(det_p)), ]
keep   <- rowSums(det_p < 0.01) == ncol(gset)
gset   <- gset[keep, ]

# Number of probes after filtering
nprobes_detectionfilter <- dim(gset)[1]

## (ii) SNP-affected probes 
gset <- minfi::dropLociWithSnps(gset, snps = c("SBE", "CpG"), maf = 0)

nprobes_snp <- dim(gset)[1]

## (iii) cross-reactive probes 
gset <- maxprobes::dropXreactiveLoci(gset)

nprobes_cross <- dim(gset)[1]

## (iv) sex chromosomes 
anno <- minfi::getAnnotation(gset)
gset <- gset[!anno$chr %in% c("chrX", "chrY"), ]

nprobes_final <- dim(gset)[1]


## Clean up memory 
rm(rgset); gc()



```

`r nprobes_init` normalized probes were filtered based on detection which retained `r nprobes_detectionfilter`, after SNP-overlap (n = `r nprobes_snp`), cross-reactivity (n = `r nprobes_cross`) and sex-chromosome mapping filtering, n = `r nprobes_final` remained in the data set. 

### Calculation of methylation quantities

An Illumina array detecting methylation values produces arbitrary signals for the unmethylated (U) and methylated (M) portion of the i^th^ genomic position. Converting these two values to a methylation score is done trough the calculation of a &beta; value.

$$
\beta_i = \frac{M_i}{M_i+U_i + c}
$$
The bias constant c, is added to the denominator to avoid large variations in the &beta; score when probe intensities are low [@bibikova2006]. This value was set to $c = 100$ to achieve regularization of the &beta; values in cases where both M and U were small [@bibikova2006], which is also the default in the `minfi` package when `type = "illumina"`. However, the successful choice of c is noted to depend on the pre-processing of the data (see `?minfi::getBeta`). 

Inspecting the normalized signals representing methylated and unmethylated probes indicate that the constant will only have a minor impact on most sites as both *M* and *U* are >> 100 @fig-raw-data-characteristics A.   

```{r}
#| label: fig-raw-data-characteristics
#| fig-cap: "Methylated and Unmethylated signals accross samples and a subset of the genomic sites (n = 1000) shows that most signals are well above 100 (A). Calculation of beta; values with different selections of *c* indicate minor differences in the overall distribution of values (B). Both figures have the relative density on the y axis (not shown)"

# Extract the methylated and unmethylated signals
M <- minfi::getMeth(gset)
U <- minfi::getUnmeth(gset)

# Set the number of probes
n <- 1000

# Figure A: Methylated and Unmethylated signals accross samples. 
set.seed(1)

# Sample sites 
sites <- sample(rownames(M), n)

figa <- data.frame(M)[sites,] |> 

   
      tibble::rownames_to_column(var = "id") |> 
      
      pivot_longer(starts_with("GSM"), values_to = "M") |> 
      
      inner_join(
            data.frame(U)[sites,] |> 
                  tibble::rownames_to_column(var = "id") |> 
                  pivot_longer(starts_with("GSM"), values_to = "U")     
      ) |> 
      pivot_longer(cols = M:U, 
                   names_to = "signal") |> 
      ggplot(aes(value, group = paste0(signal,name), 
                 color = signal)) + 
      geom_line(stat = "density") +
      scale_x_log10()  + 
      theme_classic() + 
      theme(axis.title.y = element_blank(), 
            axis.ticks.y = element_blank(), 
            axis.line.y = element_blank(), 
            axis.text.y = element_blank()) + 
      scale_color_manual(values = colors) + 
      labs(x = "Normalized signal (AU)", 
           color = "Signal type")


figb <- data.frame(M)[sites,] |> 
 
   
      tibble::rownames_to_column(var = "id") |> 
      
      pivot_longer(starts_with("GSM"), values_to = "M") |> 
      
      inner_join(
            data.frame(U)[sites,] |> 

                  tibble::rownames_to_column(var = "id") |> 
                  pivot_longer(starts_with("GSM"), values_to = "U")     
      ) |> 
      
     mutate(beta0 = M / (M + U + 0), 
            beta50 = M / (M + U + 50),
            beta100 = M / (M + U + 100)) |> 
      dplyr::select(id, name, beta0:beta100) |>

      pivot_longer(cols = starts_with("beta"), 
                   names_to = "constant", 
                   values_to = "beta") |> 
      mutate(constant = as.numeric(gsub("beta", "", constant))) |> 
      
      ggplot(aes(beta, group = constant, color = as.factor(constant))) +
      geom_line(stat = "density") +
       theme_classic() + 
      theme(axis.title.y = element_blank(), 
            axis.ticks.y = element_blank(), 
            axis.line.y = element_blank(), 
            axis.text.y = element_blank(), 
            axis.title.x = ggtext::element_markdown()) + 
      scale_color_manual(values = colors) + 
      labs(x = "&beta;-values", 
           color = "Offset constant")
      

cowplot::plot_grid(figa, figb, ncol = 1)

```

We decide to go for the default value of *c* = 100 for the calculation of the &beta; values. 
&beta;-values are restricted to the 0 to 1 space, and although this is an intuitive metric for understanding the methylation status at a specific genomic site (e.g., as a percentage methylation), it is an inconvenient quantity for a statistical model that expects normally distributed errors [@laird2010; @du2010]. Indeed, when applying a Gaussian model to identify differentially methylated genomic sites, a logit transformation of the &beta;-values results in more favorable diagnostics [@du2010]. The logit transformation of the &beta; values results in M-values. 

$$
M = \log_2 \left( \frac{\beta}{1-\beta} \right)
$$

The logit transformation (using a base 2 logarithm) puts the methylation values on a scale with no upper or lower bounds (-ininf;, &ininf;, @fig-beta-mval). This transformation is done on the &beta;-values directly to avoid missing c in the calculation.  


```{r}
#| label: fig-beta-mval
#| fig-cap: "The logit transform of &beta; values gives M-values."
#| echo: false


beta <- seq(from = 0.001, to = 0.999, by = 0.001)
Mval <- log2(beta/(1-beta))

data.frame(Mval, beta) |> 
      ggplot(aes(Mval, beta)) + geom_line() + 
      theme_classic() + 
      labs(x = "M-values", y = "&beta;-values") + 
      theme(axis.title.y = ggtext::element_markdown())

```

### Inferential error rates across models 

Next, we will compare the models using type I error rate, power and discrimination determined from permutation of within-participant relabeled data. This relabeling establishes a null distribution for statistical tests performed using the *within-participant design*. The design keeps the within-participant correlation. Our approach is similar to what was described in [@gadbury2008].

We will perform the experiment on permuted data from a subset of genomic sites distributed across different strata of &beta; values. 

A fixed number of sites is sampled from each &beta;-value stratum. Within every permutation, a random subset of those sites receives a *known* effect (adding a known average effect) while the remainder are left untouched. Sites without an effect estimate the *type I error rate*, sites with an effect estimate *power*.

We will use the *M-value scale* to add the effect to permuted data with effect assignment *redrawn in every permutation*. The &beta;-values are derived from the manipulated M-values. A shift of &delta; M units is exactly the fixed-effect parameter of the Gaussian model and proportionally to &delta;&times;log(2) for the linear-predictor shift of the &beta; model with the logit link. This means that no model is favored by the parameterization. Our approach differs here from @gadbury2008. We fix the effect sizes while @gadbury2008 uses empirical effect sizes from the data. Our simplified approach enables comparisons between model fitting procedures at known effect sizes and signs across strata of &beta; values.

In the code chunk below, a single site is used to show the workflow. &beta; values are calculated from the `gset` object and transformed to M values. We are using the first row of the M value matrix and join the values with the meta data. Within each participant, time is re-sampled to create a permuted time factor. Using the permuted time factor, an average effect (&delta; = 0.3) is added to all observations belonging to the permuted time-point `3_loading`. Data are back-transformed to the &beta; scale and both model types are fitted.

We will use a set of different inferential alternatives to establish the most appropriate model and test for our data set. In small samples, linear mixed model parameter p-values from the Wald test are anti-conservative [@luke2017]. When the restricted maximum likelihood (REML) is used in place of maximum likelihood and small sample size-corrections are used (Kenward-Roger ) estimate of the standard errors can be biased leading to inflated type 1 error rates when parameter estimates  are used for inference.   


```{r}

# Prepare data 
beta_vals <- getBeta(gset, offset = 100)
m_vals <- log2(beta_vals / (1 - beta_vals))




# Selecting a random site for testing
y <- data.frame(y = m_vals[3474,]) |> 
      tibble::rownames_to_column(var = "geo_accession" ) 

set.seed(1)
dat <- metadata |> 
      select(geo_accession, participant, time) |> 
      inner_join(y) |>
      # Permutation of time within participant
      group_by(participant) |> 
      mutate(time_permute = sample(time, replace = FALSE)) |> 
      ungroup() |> 
      mutate(msim = if_else(time_permute == "3_loading", 
             y + 0.3, y), 
             betasim = plogis(msim * log(2))) 


# A general purpose summary function for ML/REML fits
temp_summary_fun <- function(m) {
      
      is_reml <- m$modelInfo$REML
      # LRT p-value
      if(!is_reml) {
            lrt_mod <- drop1(m, test = "Chisq")
            lrt_p <- lrt_mod["time_permute", "Pr(>Chi)"]
            
      } else { lrt_p <- NA }
      # Wald p-value (restricted to paremeter 3_loading)
      wald_p <- coef(summary(m))$cond["time_permute3_loading", "Pr(>|z|)"]
      # Wald p-value Satterthwaite
      wald_p_satter <- coef(summary(m, ddf = "satterthwaite"))$cond["time_permute3_loading", "Pr(>|t|)"]
      
      data.frame(lrt_p = lrt_p, 
                 wald_p = wald_p, 
                 wald_p_satter = wald_p_satter)
      
}


# Fits 
fits <- list()
# Maximum likelihood M-values
fits[[1]] <- glmmTMB::glmmTMB(msim ~ time_permute + (1|participant), 
                          data = dat)


# Maximum likelihood beta regression
fits[[2]] <- glmmTMB::glmmTMB(betasim ~ time_permute + (1|participant), 
                          data = dat, 
                          family = glmmTMB::beta_family)

# REML M-values
fits[[3]]<- glmmTMB::glmmTMB(msim ~ time_permute + (1|participant), 
                           REML = TRUE,
                          data = dat)

# REML beta regression
fits[[4]] <- glmmTMB::glmmTMB(betasim ~ time_permute + (1|participant), 
                          data = dat, 
                          REML = TRUE,
                          family = glmmTMB::beta_family)



lapply(fits,temp_summary_fun) |> 
      bind_rows() |> 
      mutate(model = c("M-values",
                       "B-values",
                       "M-values",
                       "B-values"), 
             estimator = c("ML", "ML", "REML", "REML"))



es_mmod <- coef(summary(m_mod1))$cond["time_permute3_loading","Estimate"]
es_betamod<- coef(summary(beta_mod1))$cond["time_permute3_loading","Estimate"]


es_betamod <- es_betamod / log(2)
```

The estimated effects (`r round(es_mmod,2)` and `r round(es_betamod,2)`) for the model on M-values and &beta; values respectively are comparable but will differ due to compression in the &beta; values model.

The fixed effect sizes are multiplied by 1 or -1 in a balanced fashion to create the direction of the effect. At an extreme site (&beta;-values close to 0 or 1), adding or subtracting &delta; will either compress towards the boundary or move estimates towards 0.5. Since these scenarios are not equally detectable and the two models disagree about the variance function in extreme regions, we can evaluate the effect of selecting one model over the other.

The source code for the permutation experiment can be run in full using the script `analysis/R/methylation-error-permutation.R`. Below we check if files are available from a previous run, if not the script runs and stores data in `analysis/data/derived_data/permutation/`. 


```{r}

# list files in permutation results folder
# Versioned to match the script: permutation/ holds the earlier two-arm run
# (beta, m) with the old evaluation column names, permutation_v2/ the
# four-arm run (beta, beta_reml, m, m_reml_satt).
out_dir <- here::here("analysis/data/derived_data/permutation_v2")
perm_files <- list.files(out_dir, pattern = "^perm_\\d+\\.RDS$", full.names = TRUE)

if (length(perm_files) == 0) source(
      here::here("analysis/R/methylation-error-permutation.R")
      )

```

### Loading permutation study results

@tbl-permutation-evals shows the number of failed models among all model fits for the four fitted model specifications. A total of 500 sets of permutations where performed with 1000 sites in each set over 5 &beta; value strata.


```{r}
#| label: reading-permutation-data
#| eval: true
#| label: tbl-permutation-evals

### Reading the results back
spike_time <- "3_loading"

perm_files <- list.files(out_dir, pattern = "^perm_\\d+\\.RDS$", full.names = TRUE)
perm       <- lapply(perm_files, readRDS)

perm_summaries <- dplyr::bind_rows(lapply(perm, `[[`, "summaries"))
perm_designs   <- dplyr::bind_rows(lapply(perm, `[[`, "design"))
perm_evals     <- dplyr::bind_rows(lapply(perm, `[[`, "evaluations"))

# Which coefficients with the simulated effect.
#
# The added effect shifts only the samples labelled `spike_time` under the 
# permuted labels. `spike_time` is set to 3_loading and
# 1_baseline is the reference level.

# alt_terms also accepts and omnibus test (not used here)
# Instead, the omnibus test is saved per target in the eval slot.
alt_terms <- c(paste0("time_permute", spike_time))

# Attach the per-iteration effects (delta, sign, stratum) to every test.
# is_alt is ground truth for a (site, term) pair, not for a site.
perm_dat <- perm_summaries |>
      dplyr::inner_join(perm_designs,
                        by = c("iter", "target" = "id")) |>
      # Carry the per-fit convergence code across from the evaluation slot so
      # that failed fits can be dropped from the Wald error rates too. This
      # matters because the arms can fail at different rates, which would
      # quietly change which sites each arm is scored on.
      dplyr::inner_join(perm_evals |>
                              dplyr::select(iter, target, model, convergence),
                        by = c("iter", "target", "model")) |>
      # delta is signed, so the test must be != 0 -- `delta > 0` would classify
      # every downward-spiked site as a null and contaminate the reference.
      dplyr::mutate(is_alt = delta != 0 & term %in% alt_terms)

stopifnot(any(perm_dat$is_alt))

# The seventh inferential cell: REML standard errors scored against a NORMAL
# reference, i.e. the m_reml_satt fit without its Satterthwaite correction.
# It needs no fit of its own -- glmmTMB(REML = TRUE) on the Gaussian model and
# lmerTest::lmer agree to 2.9e-10 on the standard errors and exactly on the
# estimates -- so it is recovered here from the stored columns. Separating it
# out is what lets the REML effect and the reference-distribution effect be
# read off independently rather than as one bundled "lmerTest workflow".
perm_dat <- dplyr::bind_rows(
      perm_dat,
      perm_dat |>
            dplyr::filter(model == "m_reml_satt") |>
            dplyr::mutate(
                  model = "m_reml",
                  df = NA_real_,
                  p.value = 2 * stats::pnorm(-abs(estimate / std.error))
            ) |>
            # qval must be recomputed: BH is applied across sites within a
            # term, and these are different p-values from the ones the script
            # adjusted.
            dplyr::mutate(
                  .by = c(model, iter, term),
                  qval = stats::p.adjust(p.value, method = "BH")
            )
)

# Convergence failures are dropped silently by seqwrap_summarise (NULL entries
# are removed before rbind), so differential failure between the models would
# quietly change what is being compared.
#
# `singular` is the class-native degeneracy flag and is NOT comparable across
# arms: lme4::isSingular() fires on any zero variance component (common with 8
# participants) while glmmTMB reports only a non-positive-definite Hessian.
# Pooling the two would make the lmerTest arm look like it fails on a quarter
# of the sites when it is fitting fine. Failure is therefore defined as a
# non-zero convergence code or an unusable p-value; singularity is reported
# alongside as a separate, per-arm diagnostic.

perm_evals |>
      dplyr::summarise(.by = c(model),
                       n_fits     = dplyr::n(),
                       n_fail     = sum(convergence != 0 | is.na(pval)),
                       prop_fail  = mean(convergence != 0 | is.na(pval)),
                       n_singular = sum(singular),
                       n_pdHess   = sum(!singular)) |> 
      gt()
```


### Determining the type I error rate

The type I error rates are estimated from sites without added effects. The unaffected contrasts on a site with an added effect are nominally null, but a spiked site is not exchangeable with an unspiked one on *any* contrast: adding the effect changes the fitted mean of one time-point, and because the design is unbalanced (8/8/8/7/7 samples per time-point, with one participant contributing only three samples) the shift is not orthogonal to the estimated variance components. Under the &beta; likelihood the variance is in addition a function of the mean, so the common dispersion parameter, and with it the other contrasts, is affected as well. 



```{r}
#| label: fig-permutation-type1
#| eval: true


# This figure extracts the &beta; values and shows the strata 
# used in  the permutation study. 

# Extract the beta values  ##############

# Reloading gset to make code chunk independent of above
gset <- readRDS(
  here::here("analysis/data/derived_data/seaborne-gset-normalized.RDS")
)

# Extract beta values
B <- minfi::getBeta(gset, offset = 100)

# Pre-compute densities and save in data frame
d <- density(rowMeans(B), from = 0, to = 1)
dd <- data.frame(x = d$x, y = d$y) 

fig1 <- dd |> 
      ggplot(aes(x, y)) +
      
      # Fill areas under 
        geom_area(data = subset(dd, x <= 0.2), 
                  fill = colors[1], alpha = .6) +
        geom_area(data = subset(dd, x >= 0.2 & x <= 0.4),      
            fill = colors[2], alpha = .6) +
        geom_area(data = subset(dd, x >= 0.4 & x <= 0.6),      
            fill = colors[3], alpha = .6) +
        geom_area(data = subset(dd, x >= 0.6 & x <= 0.8),      
            fill = colors[4], alpha = .6) +
             geom_area(data = subset(dd, x >= 0.8 & x <= 1),      
            fill = colors[5], alpha = .6) +
      
      geom_line() +

       theme_classic() + 
      theme(axis.title.y = element_blank(), 
            axis.ticks.y = element_blank(), 
            axis.line.y = element_blank(), 
            axis.text.y = element_blank(), 
            axis.title.x = ggtext::element_markdown(), 
            strip.background = element_rect(fill = colors)) + 
      scale_color_manual(values = colors) +
      scale_x_continuous(breaks = seq(from = 0, to = 1, by = 0.2), 
                         expand = c(0,0)) +
      labs(x = "&beta;-values")
      


# Extract Type I error rates among null effects #################


# First we extract null targets among the wald-style parameters

null_dat <- perm_dat |> 
      # Keep null effects in non affected sites
     # filter(!(delta != 0 & term == "time_permute3_loading")) 

      # Alternative, keep only all nulls
      filter(delta == 0) 
      
s1 <- null_dat |>
      filter(convergence == 0, !is.na(p.value)) |> 

    dplyr::summarise(.by = c(model, betacat, iter, term),
                     n = dplyr::n(), 
                     sig = sum(p.value < 0.05),       
                     error = sig / n)

# Collapse strata and terms within an iteration: the permutation is the     
# resampling unit, so one error rate per iteration per model.
s1_iter <- s1 |>
      dplyr::summarise(.by = c(model, iter), error = mean(error))

model_error_type1 <- s1_iter |>
      dplyr::summarise(.by = c(model),
                       m.error = mean(error),
                       s.error = stats::sd(error),
                       n_iter  = dplyr::n(),
                       mcse    = s.error / sqrt(n_iter)) |>
      dplyr::mutate(lo = m.error - 1.96 * mcse,
                    hi = m.error + 1.96 * mcse)

t1e <- s1_iter |>
    dplyr::summarise(.by = c(model),
                     m.error = mean(error), 
                     s.error = stats::sd(error),
                     n_iter = dplyr::n(), 
                     mcse = s.error / sqrt(n_iter))


model_error_type1_betacat <- s1 |>
      dplyr::summarise(.by = c(model, iter, betacat), error = mean(error)) |> 
      dplyr::summarise(.by = c(model, betacat),
                       m.error = mean(error),
                       s.error = stats::sd(error),
                       n_iter  = dplyr::n(),
                       mcse    = s.error / sqrt(n_iter)) |>
      dplyr::mutate(lo = m.error - 1.96 * mcse,
                    hi = m.error + 1.96 * mcse) |> 
      mutate(type = "Wald")


# Next we extract the omnibus p-value from the LRT-Chi-square test
# `stat` is the omnibus test statistic: a chi-square LRT for the glmmTMB arms
# and a Satterthwaite F for m_reml_satt, so it is comparable across arms only
# through `pval`. `ddf` is the Satterthwaite denominator df, NA for the LRTs.
#
# The REML arms have NO valid omnibus (a restricted likelihood ratio across
# different fixed-effect designs is meaningless), so eval_fun records test =
# "none" and pval = NA for them. The !is.na(pval) filter below therefore drops
# them from this table by construction -- they contribute Wald contrasts only.
omnibus_null_dat <-  perm_evals |> 
      select(target, test, df, ddf, stat, pval, convergence, iter, model) |> 
      dplyr::inner_join(perm_designs,
                        by = c("iter", "target" = "id")) |> 
      filter(delta == 0) |> 
      # Failed fits carry no test; keeping them in the denominator would bias
      # the error rate downwards, and it can do so unevenly across arms.
      filter(convergence == 0, !is.na(pval))

# Omnibus "stage 1" aggergation
omnibus_s1 <- omnibus_null_dat |>
    dplyr::summarise(.by = c(model, betacat, iter),
                     n = sum(!is.na(pval)), 
                     sig = sum(pval < 0.05, na.rm = TRUE),       
                     error = sig / n)

# Collapse strata and terms within an iteration: the permutation is the     
# resampling unit, so one error rate per iteration per model.
omnibus_s1_iter <- omnibus_s1 |>
      dplyr::summarise(.by = c(model, iter), error = mean(error))

# Per model error rates
omnibus_t1e <- omnibus_s1_iter |>
    dplyr::summarise(.by = c(model),
                     m.error = mean(error), 
                     s.error = stats::sd(error),
                     n_iter = dplyr::n(), 
                     mcse = s.error / sqrt(n_iter))

# Per betacat and model, omnibus test
model_error_type1_betacat_omnibus <- omnibus_s1 |>
      dplyr::summarise(.by = c(model, iter, betacat), error = mean(error)) |> 
      dplyr::summarise(.by = c(model, betacat),
                       m.error = mean(error),
                       s.error = stats::sd(error),
                       n_iter  = dplyr::n(),
                       mcse    = s.error / sqrt(n_iter)) |>
      dplyr::mutate(lo = m.error - 1.96 * mcse,
                    hi = m.error + 1.96 * mcse) |> 
      mutate(type = "LRT")


# Combnine data for the plot
model_error_type1_betacat_combined <- bind_rows(
      model_error_type1_betacat, 
      model_error_type1_betacat_omnibus
)



# Only colour strips in x-direction
strip <- ggh4x::strip_themed(
      background_x = lapply(colors, \(x) element_rect(fill = scales::alpha(x, 0.3),
                                                  colour = NA)))

fig2 <- model_error_type1_betacat_combined |> 
      mutate(betacat = case_when(betacat == "xlow" ~ "0 - 0.2", 
                                 betacat == "low" ~ "0.2 - 0.4",  
                                 betacat == "mid" ~ "0.4 - 0.6", 
                                 betacat == "high" ~ "0.6 - 0.8", 
                                 betacat == "xhigh" ~ "0.8 - 1" 
                               ), 
             model = dplyr::recode(model,
                                   beta        = "&beta;<sub>ML</sub>",
                                   beta_reml   = "&beta;<sub>REML</sub>",
                                   m           = "M<sub>ML</sub>",
                                   m_reml      = "M<sub>REML</sub>",
                                   m_reml_satt = "M<sub>Satt</sub>")) |> 
      
      
      ggplot(aes(model,  100 * m.error, shape = type))  +
      
      geom_errorbar(aes(ymin = 100 * lo,
                        ymax = 100 * hi), 
                    width = 0.5, 
                    position = position_dodge(width = 0.2)) +
      geom_hline(yintercept = 5, lty = 2, 
                 color = "gray") + 
      
      geom_point(size = 3, fill = "gray50", 
                    position = position_dodge(width = 0.2)) +
      ggh4x::facet_wrap2(~ betacat, ncol = 5, 
                         strip = strip) +
             theme_classic() + 
      theme(axis.title.x = ggtext::element_markdown(),
            axis.text.x = ggtext::element_markdown(), 
            legend.position = "none")  +
      
      scale_shape_manual(values = c(21, 24)) + 
      labs(x = "Model", 
           y = "Type I error rate (%)")
      
######### False discovery rate ######################################


qval_errors <- null_dat |>
      filter(term != "3_loading") |> 
    dplyr::summarise(.by = c(model, betacat, iter, term),
                     n = dplyr::n(), 
                     sig = sum(qval < 0.05, na.rm = TRUE),       
                     error = sig / n)
qval_errors_iter <- qval_errors |>
      dplyr::summarise(.by = c(model, iter), error = mean(error))


model_error_fdr <- qval_errors_iter |>
      dplyr::summarise(.by = c(model),
                       m.error = mean(error),
                       s.error = stats::sd(error),
                       n_iter  = dplyr::n(),
                       mcse    = s.error / sqrt(n_iter)) |>
      dplyr::mutate(lo = m.error - 1.96 * mcse,
                    hi = m.error + 1.96 * mcse)


```





## Full model run

```{r}

# The annotation file
gset <- readRDS(here::here("analysis/data/derived_data/seaborne-gset-normalized.RDS"))
anno <- minfi::getAnnotation(gset)
nsites <- nrow(anno)

```


The unfiltered Illumina EPIC data has over 800 000 CpG positions. After filtering the available dataset contained `r nsites` CpG positions. Since the strategy here is to detect differentially methylated positions (DMP) we will fit a model per available position using the &beta;- and M-values and their respective model (generalized linear model with &beta; and normal likelihoods, respectively). 


### Computational time

```{r}
#| echo: false


# Elapsed time
beta_model_full <- readRDS(
      here::here("analysis/data/derived_data/beta-model-full.RDS"))
m_model_full <- readRDS(
      here::here("analysis/data/derived_data/m-model-full.RDS"))

beta_time <- beta_model_full@elapsed_time
m_time <- m_model_full@elapsed_time

beta_time <- round(beta_time[3] /60 /60, 1)
m_time <- round(m_time[3] /60 /60, 1)

# Number of sucessfull models 


############# NEEDS FIX ##########################

beta_model_full@errors



m_model_full@errors |> 
      distinct(message)
      print()


```

Running the full model took `r beta_time` and `r m_time` hours for the &beta; and M-models, respectively using 12 cores on a laptop computer (AMD Ryzen 5 7540U, 3.2 GHz, 16 GB RAM).

The &beta;-regression model returned estimates for all `r beta_errors_sum[1]` sites while the model on M-values flagged `r m_errors_sum[3]` models for convergence issues.



```{r}


m_model_sum <- readRDS(here::here("analysis/data/derived_data/m-model-full-sum.RDS"))
beta_model_sum <- readRDS(here::here("analysis/data/derived_data/beta-model-full-sum.RDS"))


beta_model_sum$summaries |> 
      filter(grepl("time", term)) |> 

      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time3_loading")  |> 
      filter(qval < 0.05) |> 
      arrange(qval)
       




m_model_sum$summaries |> 
      filter(grepl("time", term)) |> 
      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time3_loading") |> 
      filter(qval < 0.05) |> 
      arrange(qval)
   

# Read seaborn data 
sdat <- readxl::read_excel(
      path = here::here("analysis/data/resource_data/seaborne2018-suppl.xlsx"), 
                        sheet = 2) |> 
      select(target = `Column ID`, 
             p.value = `p-value(Time)`, 
             ratio_estimate = `MeanRatio(Baseline vs. Chronic Loading)`, 
             estimate = `Estimate(Baseline vs. Chronic Loading)`
             ) |> 
      print()

levels(factor(metadata$time))

  supp <- readxl::read_excel(
      path = here::here("analysis/data/resource_data/seaborne2018-suppl.xlsx"), sheet = 2)
  theirs <- supp |>
    select(target = `Column ID`,
           their_est = `Estimate(Baseline vs. Chronic Loading)`) |>
    mutate(their_dir = -sign(their_est))      # flip to loading-minus-baseline   

  ours <- m_model_sum$summaries |>
    filter(effect == "fixed", component == "cond", term == "time3_loading") |>   
    select(target, estimate)

  cmp <- inner_join(ours, theirs, by = "target")

  mean(sign(cmp$estimate) == cmp$their_dir)      
  cor(cmp$estimate, -cmp$their_est, method = "spearman")
  table(sign(cmp$estimate), cmp$their_dir)
            
            
      ggplot(aes(p.value, org.pval)) + geom_point()


      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time4_unloading") |> 

      filter(target == "cg15154546")
      arrange(p.value) |> 
      print()



```





















## AIC


```{r}


# Evaluation function with corrected (Jacobian) AIC for model comparison
eval_aic_beta <- function(m) {
    
      # This calculates the log likelihood for the 
      # beta model. Adds n (number of obs) for the 
      # model to double check no dropped obs. 
      # Adds bsum to confirm the same input
      
      beta <- as.numeric(m$frame[[1]])
      
      out <- data.frame(ll = as.numeric(logLik(m)), 
                        n = dim(m$frame)[1], 
                        bsum = sum(beta))
      
      return(out)
      
}

eval_aic_m <- function(m) {
      
      # This calculates the log likelihood for the 
      # M model. Adds n (number of obs) for the 
      # model to double check no dropped obs. 
      # Adds bsum to confirm the same input
      # The jacobian correction is added to the log likelihood.
      
      M <- as.numeric(m$frame[[1]])
      beta <- stats::plogis(M * log(2))
      # Jacobian correction
      jac <- -log(log(2)) - log(beta) - log(1-beta)
      # Corrected ll, n, and sum of beta values
      out <- data.frame(ll = as.numeric(logLik(m)) + sum(jac), 
                        n = dim(m$frame)[1], 
                        bsum = sum(beta))
      
       return(out)
      
}




# Setting up the models
bm1 <- seqwrap_compose(
      modelfun = glmmTMB::glmmTMB, 
      data = betadf, 
      metadata = metadata, 
      samplename = "geo_accession", 
      eval_fun = eval_aic_beta,
      arguments = list(
            formula = y ~ time + (1|participant),
            family = glmmTMB::beta_family)
)


mm1 <- seqwrap_compose(
      modelfun = glmmTMB::glmmTMB, 
      data = mdf, 
      metadata = metadata, 
      samplename = "geo_accession", 
      eval_fun = eval_aic_m,
      arguments = list(
            formula = y ~ time + (1|participant))
)




bm1_results <- seqwrap(bm1, 
                       subset = subset$row_id, 
                       return_models = FALSE, 
                       cores = parallel::detectCores())


mm1_results <- seqwrap(mm1, 
                       subset = subset$row_id,
                       return_models = FALSE, 
                       cores = parallel::detectCores())







saveRDS(bm1_results, here::here("analysis/data/derived_data/bm1_results.RDS"))
saveRDS(mm1_results, here::here("analysis/data/derived_data/mm1_results.RDS"))



bm1_sum <- seqwrap_summarise(bm1_results)
mm1_sum <- seqwrap_summarise(mm1_results)



aic_dat <- mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |>  
      pivot_wider(names_from = model, 
                  values_from = c(ll, bsum)) |> 
      # Compare beta sum to check
      mutate(comp = bsum_mm - bsum_bm) |> 
      # Calculate AIC delta
      mutate(delta_aic = -2 * (ll_mm - ll_bm)) |> 
      inner_join(subset |> 
                       rename(target = id)) 
     
   aic_dat |>    
      ggplot(aes(delta_aic)) + geom_line(stat = "density") + 
      facet_wrap(~ betacat) 
      

aicm <- lm(delta_aic ~ 0 + betacat, data = aic_dat)
summary(aicm)


mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |> 
            
      inner_join(subset |> 
                       rename(target = id)) |>  
      
     
      
      ggplot(aes(quantile)) + geom_histogram() + 
      facet_grid(betacat ~ model)


mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |> 
            
      inner_join(subset |> 
                       rename(target = id)) |> 
      inner_join(cden |> 
                       rename(target = id)) |> 
      
      select(target, betacat, quantile, model, intensity_cat) |> 

      ggplot(aes(quantile)) + geom_histogram() + 

      
      facet_wrap(intensity_cat ~ model)






mm1_sum$evaluations |> 
      rename(mm = pval) |> 
      inner_join(
            bm1_sum$evaluation |> 
                  rename(bm = pval)
      ) |> 
       inner_join(subset |> 
                       rename(target = id)) |> 
      ggplot(aes(bm, mm)) + geom_point() + 
      facet_wrap(~ betacat)



bm1_sum$evaluations


m <- bm1_results@models[[1]]
m <- mm1_results@models[[1]]


summary(m)


```






see here https://link.springer.com/article/10.1186/1471-2105-11-587







