---
title: "Case study: DNA methylation (GSE114763)"
format: html
bibliography: "resources/bibliography-methylation.bib"
csl: "resources/bmc-bioinformatics.csl"
editor_options: 
  chunk_output_type: console
---

## Introduction 

The following document outlines a partial reanalysis of data on CpG methylation status in human skeletal muscle in response to resistance exercise training, followed by detraining and subsequent retraining [@seaborne2018;@seaborne2018b]. All relevant code is explicitly shown here, whereas the version of this case study included in the manuscript has been stripped of some details.


## Raw data and pre-processing



```{r}
#| label: dependencies
#| eval: true

library(GEOquery)      # For accessing the raw data
library(minfi)         # For pre-processing of raw data
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(maxprobes)     # For filtering cross-reactive probes 
library(seqwrap)


# Load color scale 
source(here::here("analysis/figures/figure-opts.R"))


```


Raw data are downloaded from the [Gene Expression Omnibus](https://www.ncbi.nlm.nih.gov/geo/) using the accession number [GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763). 


```{r}
#| label: paths
#| eval: true

# Prepare data download. 
# Note for use in the present repo: 
# raw_data and derived data are gitignored so safe for 
# large files. If code-re-use, check the download folder before committing 
# changes to git. 

# The GEO ID
gse_id   <- "GSE114763"

# Set download directories
raw_dir  <- here::here("analysis/data/raw_data", gse_id)
idat_dir <- file.path(raw_dir, "idat")
der_dir  <- here::here("analysis/data/derived_data")

dir.create(idat_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(der_dir,  recursive = TRUE, showWarnings = FALSE)

```

### Downloading data

Sample titles and characteristics are read from the SOFT text view of the series, which returns the metadata of all 40 samples. The supplementary archive (`GSE114763_RAW.tar`, ~722 MB) provides the IDAT files.

```{r}
#| label: download
#| eval: true


# The metadata (pheno_raw) and the raw data IDAT files are cached after 
# downloading once. 



# A function for retrying connection to GEO for downloading large data.
# This was created after issues with downloading the large zipped archive
# from GEO. 
retry_url <- function(expr, tries = 5, pause = 5) {
  call <- substitute(expr)          # re-evaluated on each attempt
  env  <- parent.frame()
  for (i in seq_len(tries)) {
    out <- try(eval(call, env), silent = TRUE)
    if (!inherits(out, "try-error")) return(out)
    if (i < tries) Sys.sleep(pause * i)
  }
  stop("failed after ", tries, " attempts: ",
       conditionMessage(attr(out, "condition")))
}

# Series-level metadata
#
# The metadata contains file names and titles include 
# time-points and participant information. 
pheno_file <- file.path(raw_dir, "pheno_raw.rds")

if (!file.exists(pheno_file)) {
  soft_url <- paste0("https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi",
                     "?acc=", gse_id, "&targ=gsm&form=text&view=brief")

  soft <- retry_url({
    res <- curl::curl_fetch_memory(soft_url)
    if (res$status_code != 200) stop("HTTP ", res$status_code)
    strsplit(rawToChar(res$content), "\r?\n")[[1]]
  })

  # one record per sample; "!Sample_<key> = <value>" lines within each
  kv <- tibble::tibble(record = cumsum(startsWith(soft, "^SAMPLE")),
                       line   = soft) |>
    dplyr::filter(startsWith(line, "!")) |>
    tidyr::separate(line, into = c("key", "value"), sep = " = ",
                    extra = "merge") |>
    dplyr::mutate(key = stringr::str_remove(key, "^!Sample_"))

  # characteristics are themselves "<key>: <value>" pairs
  characteristics <- kv |>
    dplyr::filter(key == "characteristics_ch1") |>
    tidyr::separate(value, into = c("key", "value"), sep = ": ",
                    extra = "merge") |>
    dplyr::filter(key %in% c("gender", "age", "weight", "height")) 

  # Cleaning the metadata for downstream processing. 
  pheno_raw <- kv |>
    dplyr::filter(key %in% c("geo_accession", "title")) |>
    dplyr::bind_rows(characteristics) |>
    tidyr::pivot_wider(id_cols = record, names_from = key,
                       values_from = value) |>
    dplyr::select(-record) |> 

        mutate(title_sep = title) |> 
        separate(title_sep, into = c("seq_sample_id", "timeid"), sep = ":") |>
        mutate(timeid = gsub(" Muscle_", "", timeid)) |> 
        mutate(timeid = gsub("Participant_", "", timeid)) |> 
        mutate(timeid = gsub("7wk_", "7wk", timeid)) |> 
        
         separate(timeid, into = c("time", "participant", "replicate"), sep = "_") |> 
      # Making time-point more obvious
        mutate(time = case_when(
              time == "Baseline" ~ "1_baseline", 
              time == "Acute" ~ "2_acute", 
              time == "7wkLoading" ~ "3_loading", 
              time == "7wkUnloading"  ~ "4_unloading", 
              time == "7wkRe-oading" ~ "5_reloading"
        )) |> 
        # Fixing meta-data details
        mutate(weight = gsub(" kg", "", weight), 
               height = gsub("cm", "", height), 
               weight = as.numeric(weight), 
               height = as.numeric(height), 
               age = as.numeric(age)) |> 
        # Fix a misprinted participant id
        mutate(participant = gsub("P006", "006", participant)) 
  

  saveRDS(pheno_raw, pheno_file)
}

pheno_raw <- readRDS(pheno_file)

# Double check 40 expected samples
stopifnot(nrow(pheno_raw) == 40)

# Download and extract raw IDAT files.
# IDATs are distributed gzipped; minfi reads decompressed files
tar_file <- file.path(raw_dir, paste0(gse_id, "_RAW.tar"))

if (!file.exists(tar_file)) {
  # ~722 MB; increase the timeout for slow connections
  options(timeout = 60 * 60)
  retry_url(GEOquery::getGEOSuppFiles(gse_id, baseDir = dirname(raw_dir),
                                      makeDirectory = TRUE))
}

if (length(list.files(idat_dir, pattern = "\\.idat$")) == 0) {
  utils::untar(tar_file, exdir = idat_dir)

  # Unzip idat files
  gz <- list.files(idat_dir, pattern = "\\.idat\\.gz$", full.names = TRUE)
  invisible(lapply(gz, R.utils::gunzip, overwrite = TRUE, remove = TRUE))
}

```

### Metadata summary

A total of 40 samples were downloaded from the Gene Expression Omnibus ([GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763)). The identified metadata (@tbl-meta-samples) corresponds to the metadata published in [@seaborne2018b] (see table 2 in @seaborne2018b).

Seaborne et al. removed two samples after preliminary analysis, (samples `SkM_Epi_Mem_1` and `SkM_Epi_Mem_39`), we will do the same.


```{r}
#| label: tbl-meta-samples
#| tbl-cap: "Sample metadata retreived from GEO ([GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763)) corresponding to @seaborne2018b".
#| echo: false


library(gt)

pheno_raw |> 
      summarise(.by = participant, 
                weight = mean(weight), 
                height = mean(height), 
                age = mean(age)) |> 
      mutate(participant = gsub("00", "", participant)) |> 
      select(participant, age, height, weight) |> 
      gt() |> 
      cols_label(participant = "Participant", 
                 age = "Age (years)", 
                 weight = "Weight (kg)", 
                 height = "Height (cm)")
      
      
```


## Case study description


```{r}
#| label: case-description
#| echo: false

# Double check gender 
# pheno_raw |> 
#      distinct(gender)
# Confirm all males


# Get sample metadata into text
sample_metadata <- pheno_raw |> 
      
      summarise(.by = participant, 
                weight = mean(weight), 
                height = mean(height), 
                age = mean(age)) |> 
      pivot_longer(weight:age) |> 
      summarise(.by = name, 
                m = mean(value), 
                s = sd(value)) |> 
      mutate(unit = c("kg", "cm", "years")) |> 
      mutate(stat = paste0(sprintf("%.1f", round(m, 1)), 
                           " (", 
                           sprintf("%.1f", round(s, 1)), 
                           ") ", 
                           unit)) 

age <- filter(sample_metadata, name == "age") |> pull(stat)
height <- filter(sample_metadata, name == "height") |> pull(stat)
weight <- filter(sample_metadata, name == "weight") |> pull(stat)

```


> Seaborne et al. [@seaborne2018;@seaborne2018b] investigated whether human skeletal muscle retains an epigenetic memory of earlier hypertrophy-inducing resistance training. For this aim, previously untrained male participants (age, [mean (SD)] `r age`; weight, `r weight`; height, `r height`) completed a sequential design with two seven-week periods of resistance-exercise training (RET) intersected with a 7-week unloading period without RET. RET periods included two weekly sessions targeting lower limbs (seven exercises) and one weekly session targeting the upper body (5 exercises; 4 sets of 10 repetitions per exercise). Skeletal muscle biopsies were sampled from m. vastus lateralis on five occasions: before the intervention, 30 min after the first exercise session, and after the completion of each of the 7-week periods. Genomic DNA was extracted from muscle biopsy samples (DNeasy Kit, Qiagen, Manchester, UK) and bisulfite converted before amplification, fragmentation, and hybridization using the Illumina Infinium Human MethylationEPIC Beadchip protocol [@seaborne2018;@seaborne2018b]. 


The Sentrix slide (chip) and array position are recovered from the IDAT file names (e.g. `GSM3149860_201496860025_R01C01_Grn.idat`) and retained as potential technical covariates in subsequent analyses. `Basename` is the file stem that `minfi` uses to locate the red/green channel pair.

The samples are distributed as described in [@seaborne2018;@seaborne2018b] with one participant dropping out from the study prior to the last sampling time-point. One participant have duplicate arrays at one time-point. After removal of samples based on preliminary QC-analysis (see @seaborne2018b), the full data set contains 38 samples with no technical duplicates. 

```{r}
#| label: tbl-metadata-count
#| eval: true
#| tbl-cap: "Samples per time-point and participant."


idat_stems <- list.files(idat_dir, pattern = "_Grn\\.idat$") |>
  stringr::str_remove("_Grn\\.idat$")

array_info <- tibble::tibble(stem = idat_stems) |>
  tidyr::separate(stem, into = c("geo_accession", "slide", "array"),
                  sep = "_", remove = FALSE) |>
  dplyr::mutate(Basename = file.path(idat_dir, stem))

metadata <- pheno_raw |>
      # Removing outlier sample, see "Quality Control of DNA Methylation Probes"
      # in @seaborne2018b
      filter(!grepl("SkM_Epi_Mem_1:", title), 
             !grepl("SkM_Epi_Mem_39:", title)) |> 
      
  tibble::as_tibble() |>
  dplyr::select(geo_accession,
                title,
                participant,
                gender,
                age, 
                time) |>
  dplyr::mutate(
    label = stringr::str_remove(title, "^SkM_Epi_Mem_\\d+:\\s*")) |>
  dplyr::left_join(array_info, by = "geo_accession")

stopifnot(!anyNA(metadata$time), !anyNA(metadata$participant))

# 38 arrays: 8 participants x 5 time points with missing values for on 
# participant.

metadata |> 
      summarise(.by = c(time, participant), 
                n = n()) |> 
      pivot_wider(names_from = time, values_from = n) |> 
      select(Participant = participant, 
             Baseline = `1_baseline`, 
             Acute = `2_acute`, 
             Loading = `3_loading`, 
             Unloading = `4_unloading`, 
             Reloading = `5_reloading`) |> 
      gt() |> 
      tab_spanner(label = "Sample time-points", columns = "Baseline":"Reloading")
      



```

### Reading IDAT and array-level quality control

We will use the `minfi` package [@fortin2014] to read the raw data and perform quality control, filtering and normalization. As a first step, array-level filtering is done using detection p-values where methylated and unmethylated positions are compared to negative control positions. This comparison produces a p-value used to identify problematic genomic positions. As shown below (@fig-array-qc), array-level aggregates show mean values < 0.01.    


```{r}
#| label: fig-array-qc
#| fig-cap: "Array-level quality control. Mean detection p-values and fraction of failed genomic positions per array subdivided by time-point."
#| eval: true

rgset <- minfi::read.metharray(metadata$Basename, extended = TRUE, force = TRUE)

# Check that orders correspond
stopifnot(identical(colnames(rgset), basename(metadata$Basename)))

colnames(rgset) <- metadata$geo_accession


det_p <- minfi::detectionP(rgset)

# 
qc <- tibble::tibble(
  geo_accession = colnames(det_p),
  mean_det_p    = colMeans(det_p),
  frac_failed   = colMeans(det_p > 0.01))

qc |> 
      inner_join(pheno_raw |> 
                       select(geo_accession, time)) |> 
      ggplot(aes(mean_det_p, frac_failed, color = time)) + 
      geom_point() +
      labs(x = "Array-level mean detection p-value", 
           y = "Fraction of failed genomic positions (p > 0.01)", 
           color = "Time-point")
```



```{r}

# Below is the suggested workflow for confirming sex. We expect only males and 
# the function indficate that only one sex is present which should suffice as 
# confirmation.
# predicted   <- minfi::getSex(minfi::mapToGenome(msetraw))
# qc$pred_sex <- predicted$predictedSex

# All samples pass quality control
keep_samples <- qc$geo_accession[qc$mean_det_p < 0.01]
# length(keep_samples)

rgset    <- rgset[, keep_samples]
det_p    <- det_p[, keep_samples]
metadata <- metadata |> dplyr::filter(geo_accession %in% keep_samples)
```

### Normalization and probe filtering

Functional normalization [@fortin2014] is used because the study design can be expected to result in global methylation differences. Functional normalization removes technical variation captured by the array control probes without assuming a common methylation distribution across samples [@fortin2014]. Probes are then removed if they (1) fail detection in any retained sample, (2) overlap common SNPs at the target CpG or single-base extension site, (3) appear on published lists of cross-reactive EPIC probes (using the [`maxprobes`](https://github.com/markgene/maxprobes/) package), or (iv) map to the sex chromosomes (all male cohort).

```{r}
#| label: normalise-filter
#| eval: true



# Functional normalization
gset <- minfi::preprocessFunnorm(rgset, ratioConvert = FALSE)

# Initial number of probes
nprobes_init <- dim(gset)[1]

## (i) detection
det_p  <- det_p[match(minfi::featureNames(gset), rownames(det_p)), ]
keep   <- rowSums(det_p < 0.01) == ncol(gset)
gset   <- gset[keep, ]

# Number of probes after filtering
nprobes_detectionfilter <- dim(gset)[1]

## (ii) SNP-affected probes 
gset <- minfi::dropLociWithSnps(gset, snps = c("SBE", "CpG"), maf = 0)

nprobes_snp <- dim(gset)[1]

## (iii) cross-reactive probes 
gset <- maxprobes::dropXreactiveLoci(gset)

nprobes_cross <- dim(gset)[1]

## (iv) sex chromosomes 
anno <- minfi::getAnnotation(gset)
gset <- gset[!anno$chr %in% c("chrX", "chrY"), ]

nprobes_final <- dim(gset)[1]


## Clean up memory 
rm(rgset); gc()



```

`r nprobes_init` normalized probes were filtered based on detection which retained `r nprobes_detectionfilter`, after SNP-overlap (n = `r nprobes_snp`), cross-reactivity (n = `r nprobes_cross`) and sex-chromosome mapping filtering, n = `r nprobes_final` remained in the data set. 

### Calculation of methylation quantities

An Illumina array detecting methylation values produces arbitrary signals for the unmethylated (U) and methylated (M) portion of the i^th^ genomic position. Converting these two values to a methylation score is done trough the calculation of a &beta; value.

$$
\beta_i = \frac{M_i}{M_i+U_i + c}
$$
The bias constant c, is added to the denominator to avoid large variations in the &beta; score when probe intensities are low [@bibikova2006]. This value was set to $c = 100$ to achieve regularization of the &beta; values in cases where both M and U were small [@bibikova2006], which is also the default in the `minfi` package when `type = "illumina"`. However, the successful choice of c is noted to depend on the pre-processing of the data (see `?minfi::getBeta`). 

Inspecting the normalized signals representing methylated and unmethylated probes indicate that the constant will only have a minor impact on most sites as both *M* and *U* are >> 100 @fig-raw-data-characteristics A.   

```{r}
#| label: fig-raw-data-characteristics
#| fig-cap: "Methylated and Unmethylated signals accross samples and a subset of the genomic sites (n = 1000) shows that most signals are well above 100 (A). Calculation of beta; values with different selections of *c* indicate minor differences in the overall distribution of values (B). Both figures have the relative density on the y axis (not shown)"

# Extract the methylated and unmethylated signals
M <- minfi::getMeth(gset)
U <- minfi::getUnmeth(gset)

# Set the number of probes
n <- 1000

# Figure A: Methylated and Unmethylated signals accross samples. 
set.seed(1)

# Sample sites 
sites <- sample(rownames(M), n)

figa <- data.frame(M)[sites,] |> 

   
      tibble::rownames_to_column(var = "id") |> 
      
      pivot_longer(starts_with("GSM"), values_to = "M") |> 
      
      inner_join(
            data.frame(U)[sites,] |> 
                  tibble::rownames_to_column(var = "id") |> 
                  pivot_longer(starts_with("GSM"), values_to = "U")     
      ) |> 
      pivot_longer(cols = M:U, 
                   names_to = "signal") |> 
      ggplot(aes(value, group = paste0(signal,name), 
                 color = signal)) + 
      geom_line(stat = "density") +
      scale_x_log10()  + 
      theme_classic() + 
      theme(axis.title.y = element_blank(), 
            axis.ticks.y = element_blank(), 
            axis.line.y = element_blank(), 
            axis.text.y = element_blank()) + 
      scale_color_manual(values = colors) + 
      labs(x = "Normalized signal (AU)", 
           color = "Signal type")


figb <- data.frame(M)[sites,] |> 
 
   
      tibble::rownames_to_column(var = "id") |> 
      
      pivot_longer(starts_with("GSM"), values_to = "M") |> 
      
      inner_join(
            data.frame(U)[sites,] |> 

                  tibble::rownames_to_column(var = "id") |> 
                  pivot_longer(starts_with("GSM"), values_to = "U")     
      ) |> 
      
     mutate(beta0 = M / (M + U + 0), 
            beta50 = M / (M + U + 50),
            beta100 = M / (M + U + 100)) |> 
      dplyr::select(id, name, beta0:beta100) |>

      pivot_longer(cols = starts_with("beta"), 
                   names_to = "constant", 
                   values_to = "beta") |> 
      mutate(constant = as.numeric(gsub("beta", "", constant))) |> 
      
      ggplot(aes(beta, group = constant, color = as.factor(constant))) +
      geom_line(stat = "density") +
       theme_classic() + 
      theme(axis.title.y = element_blank(), 
            axis.ticks.y = element_blank(), 
            axis.line.y = element_blank(), 
            axis.text.y = element_blank(), 
            axis.title.x = ggtext::element_markdown()) + 
      scale_color_manual(values = colors) + 
      labs(x = "&beta;-values", 
           color = "Offset constant")
      

cowplot::plot_grid(figa, figb, ncol = 1)

```

We decide to go for the default value of *c* = 100 for the calculation of the &beta; values. 
&beta;-values are restricted to the 0 to 1 space, and although this is an intuitive metric for understanding the methylation status at a specific genomic site (e.g., as a percentage methylation), it is an inconvenient quantity for a statistical model that expects normally distributed errors [@laird2010; @du2010]. Indeed, when applying a Gaussian model to identify differentially methylated genomic sites, a logit transformation of the &beta;-values results in more favorable diagnostics [@du2010]. The logit transformation of the &beta; values results in M-values. 

$$
M = \log_2 \left( \frac{\beta}{1-\beta} \right)
$$

The logit transformation (using a base 2 logarithm) puts the methylation values on a scale with no upper or lower bounds (-ininf;, &ininf;, @fig-beta-mval). This transformation is done on the &beta;-values directly to avoid missing c in the calculation.  


```{r}
#| label: fig-beta-mval
#| fig-cap: "The logit transform of &beta; values gives M-values."
#| echo: false


beta <- seq(from = 0.001, to = 0.999, by = 0.001)
Mval <- log2(beta/(1-beta))

data.frame(Mval, beta) |> 
      ggplot(aes(Mval, beta)) + geom_line() + 
      theme_classic() + 
      labs(x = "M-values", y = "&beta;-values") + 
      theme(axis.title.y = ggtext::element_markdown())

```



```{r}
# Data preparation.

# Calculate the beta-values with the Illumina-style offset (c = 100)
beta_vals <- getBeta(gset, offset = 100)
# Derive M-values from the calculated beta values.
m_vals <- log2( beta_vals / (1-beta_vals) )


# No need to squeeze beta values away from the boundaries: (y(n - 1) + 0.5) / n
# n_obs     <- ncol(beta_vals)
# beta_vals <- (beta_vals * (n_obs - 1) + 0.5) / n_obs

betadf <- data.frame(beta_vals) |> 
      tibble::rownames_to_column(var = "id") |> 
      dplyr::select(id, starts_with("GSM")) 
      
mdf <- data.frame(m_vals) |> 
      tibble::rownames_to_column(var = "id") |> 
      dplyr::select(id, starts_with("GSM"))

```




### Type I and II error rates permutation study

**Scripts have moved to `analysis/R/methylation-error-permutation.R`.**

Next, we will compare the models using Type I and II error rates determined from permutation of within-participant relabeled data. This relabeling establishes a null distribution for statistical tests performed using the *within-participant design*. The design keeps the within-participant correlation.

We will perform the experiment on permutated data from a subset of genomic sites distributed across different strata of &beta; values. 

The permutation loop below produces two results from a single set of model fits. A fixed number of sites is sampled from each &beta;-value stratum. Within every permutation, a subset of those sites receives a *known* effect (adding a known average effect) while the remainder are left untouched. Sites without an effect estimate the *type I error rate*, sites with an effect estimate *power*. This will also give us an estimate of the *false discovery proportion*.

We will use the *M-value scale* to add the effect to permutated data with effect assignment *redrawn in every permutation*. The &beta;-values are derived from the manipulated M-values. A shift of &delta; M units is exactly the fixed-effect parameter of the Gaussian model and exactly the linear-predictor shift of the &beta; model with the logit link. This means that no model is favored by the parameterization.

The effect sizes are multiplied by 1 or -1 in a balanced fashion to create the direction of the effect. At an extreme site (&beta;-values close to 0 or 1), adding or subtracting &delta; will either compress towards the boundary or move estimates towards 0.5. Since these scenarios are not equally detectable and the two models disagree about the variance function in extreme regions, we can evaluate the effect of selecting one model over the other.



### The permutation loop

Each iteration writes a single RDS file, completed iterations are not refitted. The seed is set per iteration so that any single permutation can be reproduced on its own.

----- THIS IS WHERE I AM -----------------

### Loading permutation study results


```{r}
#| label: reading-permutation-data
#| eval: false

### Reading the results back
perm_files <- list.files(out_dir, pattern = "^perm_\\d+\\.RDS$", full.names = TRUE)
perm       <- lapply(perm_files, readRDS)

perm_summaries <- dplyr::bind_rows(lapply(perm, `[[`, "summaries"))
perm_designs   <- dplyr::bind_rows(lapply(perm, `[[`, "design"))
perm_evals     <- dplyr::bind_rows(lapply(perm, `[[`, "evaluations"))

# Which coefficients with the simulated effect.
#
# The spike shifts only the samples labelled `spike_time` under the permuted
# labels. With 1_baseline as the reference level, the only contrast with a
# non-zero true value is the one against spike_time. For a site with simulated
# effects in spike_time, all other parameters (2_acute, 4_unloading and 
# 5_reloading) are still null. The omnibus
# LRT is an alternative on a spiked site, because the time factor as a whole
# does have an effect.
alt_terms <- c(paste0("time_permute", spike_time), "time_omnibus")

# Attach the per-iteration truth (delta, sign, stratum) to every test.
# is_alt is ground truth for a (site, term) pair, not for a site.
perm_dat <- perm_summaries |>
      dplyr::inner_join(perm_designs,
                        by = c("iter", "target" = "id")) |>
      # delta is signed, so the test must be != 0 -- `delta > 0` would classify
      # every downward-spiked site as a null and contaminate the reference.
      dplyr::mutate(is_alt = delta != 0 & term %in% alt_terms)

stopifnot(any(perm_dat$is_alt))

# Convergence failures are dropped silently by seqwrap_summarise (NULL entries
# are removed before rbind), so differential failure between the two models
# would quietly change what is being compared. Report it.
perm_evals |>
      dplyr::summarise(.by = c(model),
                       n_fits     = dplyr::n(),
                       n_fail     = sum(convergence != 0 | !pdHess),
                       prop_fail  = mean(convergence != 0 | !pdHess))


```

### Type I error rate

Calibration is measured on the unspiked sites only. The unaffected contrasts on a
*spiked* site are nominally null too, but adding the effect changes the fitted mean
of one group, and under the beta likelihood the variance is a function of the mean --
so the shared dispersion parameter, and with it the other contrasts, are mildly
perturbed. Those tests are excluded from the reference distribution rather than
pooled into it.

```{r}
#| label: permutation-type1
#| eval: false

null_dat <- perm_dat |> dplyr::filter(delta == 0)

t1e <- null_dat |>
      dplyr::summarise(.by = c(model, term, betacat),
                       n      = dplyr::n(),
                       a_05   = mean(p.value < 0.05, na.rm = TRUE),
                       a_01   = mean(p.value < 0.01, na.rm = TRUE),
                       a_001  = mean(p.value < 0.001, na.rm = TRUE))


# Null p-value distribution by stratum
null_dat |>
      ggplot(aes(sample = p.value)) +
      stat_qq(distribution = stats::qunif) +
      stat_qq_line(distribution = stats::qunif) +
      facet_grid(betacat ~ model) +
      labs(x = "Expected quantile (uniform)", y = "Observed p-value")

```

### Power at matched empirical size

Comparing power at the nominal 5% level is confounded by size: a model whose actual
type I error is 7% will appear more powerful simply because it rejects more of
everything. The empirical 5% quantile of the null p-values, taken per model and
stratum from the *same* permutations, is used as the critical value instead, so both
models are compared while operating at exactly 5% empirical size.

```{r}
#| label: permutation-power
#| eval: false

crit <- null_dat |>
      dplyr::summarise(.by = c(model, term, betacat),
                       crit = stats::quantile(p.value, 0.05, na.rm = TRUE))

# is_alt, not delta != 0: only the contrast against spike_time (and the omnibus
# test) carries the effect. Grouping must keep `term`, otherwise the spiked
# contrast is pooled with the three unaffected ones and power is diluted towards
# the nominal level. `delta` is signed here, so upward and downward spikes are
# reported separately -- keep `delta_abs` too for the pooled view.
power_dat <- perm_dat |>
      dplyr::filter(is_alt) |>
      dplyr::inner_join(crit, by = c("model", "term", "betacat")) |>
      dplyr::summarise(.by = c(model, term, betacat, delta, delta_abs),
                       n             = dplyr::n(),
                       power_nominal = mean(p.value < 0.05, na.rm = TRUE),
                       power_matched = mean(p.value < crit, na.rm = TRUE))

# Power against the signed effect. At the extreme strata the two directions are
# not equally detectable -- an upward spike at an xhigh site compresses towards
# the boundary -- so asymmetry about zero is a result, not noise.
power_dat |>
      ggplot(aes(delta, power_matched, colour = model)) +
      geom_line(position = position_dodge(width = 0.02)) +
      geom_point(position = position_dodge(width = 0.02)) +
      facet_grid( ~ betacat) +
      scale_colour_manual(values = colors) +
      labs(x = "Simulated effect (M-value units, signed)",
           y = "Power at 5% empirical size",
           colour = "Model")

# Pooled over direction
power_dat |>
      dplyr::summarise(.by = c(model, term, betacat, delta_abs),
                       power_matched = stats::weighted.mean(power_matched, n),
                       n = sum(n)) |>
      ggplot(aes(delta_abs, power_matched, colour = model)) +
      geom_line(position = position_dodge(width = 0.02)) +
      geom_point(position = position_dodge(width = 0.02)) +
      facet_grid(term ~ betacat) +
      scale_colour_manual(values = colors) +
      labs(x = "Simulated effect (M-value units, magnitude)",
           y = "Power at 5% empirical size",
           colour = "Model")
```

### Realised false discovery proportion

Because true nulls and true alternatives are present in the same family of tests,
the Benjamini-Hochberg step can be scored directly. Note that this is a statement
about a family of `size * 5` sites, not about the full array: the BH threshold for
the smallest p-value scales with the number of tests, so this does not certify FDR
control at the ~700,000-site scale.

```{r}
#| label: permutation-fdp
#| eval: false

# Only the families that contain true alternatives can be scored as an FDR.
# A false discovery is a rejection of a (site, term) pair that is not is_alt.
fdp <- perm_dat |>
      dplyr::filter(term %in% alt_terms) |>
      dplyr::summarise(.by = c(model, term, iter),
                       R   = sum(qval < 0.05, na.rm = TRUE),
                       V   = sum(qval < 0.05 & !is_alt, na.rm = TRUE),
                       FDP = ifelse(R > 0, V / R, 0))

fdp |>
      dplyr::summarise(.by = c(model, term),
                       n_iter    = dplyr::n(),
                       mean_R    = mean(R),
                       FDR       = mean(FDP))

# The remaining contrasts are complete nulls -- no site carries an effect on
# them -- so every discovery is false and FDP is 1 whenever anything is
# rejected. There the quantity of interest is P(at least one rejection), which
# BH bounds by q under the complete null.
perm_dat |>
      dplyr::filter(!term %in% alt_terms) |>
      dplyr::summarise(.by = c(model, term, iter),
                       R = sum(qval < 0.05, na.rm = TRUE)) |>
      dplyr::summarise(.by = c(model, term),
                       n_iter = dplyr::n(),
                       p_any  = mean(R > 0))
```

### Agreement between the two models

```{r}
#| label: permutation-agreement
#| eval: false

perm_dat |>
      dplyr::filter(delta != 0) |>
      dplyr::select(iter, target, term, betacat, model,estimate, p.value) |>
      tidyr::pivot_wider(names_from = model, values_from = c(estimate,p.value)) |>
      
      print()
      
      ggplot(aes(-log10(m), -log10(beta))) +
      geom_point(alpha = 0.2) +
      geom_abline(slope = 1, intercept = 0) +
      facet_wrap(~ betacat) +
      labs(x = "M-value model, -log10(p)",
           y = "Beta model, -log10(p)")
```



## Full model run

```{r}

# The annotation file
gset <- readRDS(here::here("analysis/data/derived_data/seaborne-gset-normalized.RDS"))
anno <- minfi::getAnnotation(gset)
nsites <- nrow(anno)

```


The unfiltered Illumina EPIC data has over 800 000 CpG positions. After filtering the available dataset contained `r nsites` CpG positions. Since the strategy here is to detect differentially methylated positions (DMP) we will fit a model per available position using the &beta;- and M-values and their respective model (generalized linear model with &beta; and normal likelihoods, respectively). 


### Computational time

```{r}
#| echo: false


# Elapsed time
beta_model_full <- readRDS(
      here::here("analysis/data/derived_data/beta-model-full.RDS"))
m_model_full <- readRDS(
      here::here("analysis/data/derived_data/m-model-full.RDS"))

beta_time <- beta_model_full@elapsed_time
m_time <- m_model_full@elapsed_time

beta_time <- round(beta_time[3] /60 /60, 1)
m_time <- round(m_time[3] /60 /60, 1)

# Number of sucessfull models 
beta_errors_sum <- sapply(beta_model_full@errors, function(x) sum(!sapply(x, is.null)))
m_errors_sum <- sapply(m_model_full@errors, function(x) sum(!))

rm(beta_model_full)
rm(m_model_full)

gc()


# What were the errors
m_warnings <- m_model_full@errors |> 
      filter(!sapply(warnings_fit, is.null)) |> 
      pull(warnings_fit) 

```

Running the full model took `r beta_time` and `r m_time` hours for the &beta; and M-models, respectively using 12 cores on a laptop computer (AMD Ryzen 5 7540U, 3.2 GHz, 16 GB RAM).

The &beta;-regression model returned estimates for all `r beta_errors_sum[1]` sites while the model on M-values flagged `r m_errors_sum[3]` models for convergence issues.



```{r}


m_model_sum <- readRDS(here::here("analysis/data/derived_data/m-model-full-sum.RDS"))
beta_model_sum <- readRDS(here::here("analysis/data/derived_data/beta-model-full-sum.RDS"))


beta_model_sum$summaries |> 
      filter(grepl("time", term)) |> 

      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time3_loading") |> 
      




m_model_sum$summaries |> 
      filter(grepl("time", term)) |> 
      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time4_unloading") |> 
      filter(p.value < 0.05) |> 
      mutate(direction = if_else(sign(estimate) > 0, "pos", "neg")) |> 
      summarise(.by = direction, 
                n = n())


# Read seaborn data 
sdat <- readxl::read_excel(
      path = here::here("analysis/data/resource_data/seaborne2018-suppl.xlsx"), 
                        sheet = 2) |> 
      select(target = `Column ID`, 
             p.value = `p-value(Time)`, 
             ratio_estimate = `MeanRatio(Baseline vs. Chronic Loading)`, 
             estimate = `Estimate(Baseline vs. Chronic Loading)`
             ) |> 
      print()

levels(factor(metadata$time))

  supp <- readxl::read_excel(
      path = here::here("analysis/data/resource_data/seaborne2018-suppl.xlsx"), sheet = 2)
  theirs <- supp |>
    select(target = `Column ID`,
           their_est = `Estimate(Baseline vs. Chronic Loading)`) |>
    mutate(their_dir = -sign(their_est))      # flip to loading-minus-baseline   

  ours <- m_model_sum$summaries |>
    filter(effect == "fixed", component == "cond", term == "time3_loading") |>   
    select(target, estimate)

  cmp <- inner_join(ours, theirs, by = "target")

  mean(sign(cmp$estimate) == cmp$their_dir)      
  cor(cmp$estimate, -cmp$their_est, method = "spearman")
  table(sign(cmp$estimate), cmp$their_dir)
            
            
      ggplot(aes(p.value, org.pval)) + geom_point()


      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time4_unloading") |> 

      filter(target == "cg15154546")
      arrange(p.value) |> 
      print()



```





















## AIC


```{r}


# Evaluation function with corrected (Jacobian) AIC for model comparison
eval_aic_beta <- function(m) {
    
      # This calculates the log likelihood for the 
      # beta model. Adds n (number of obs) for the 
      # model to double check no dropped obs. 
      # Adds bsum to confirm the same input
      
      beta <- as.numeric(m$frame[[1]])
      
      out <- data.frame(ll = as.numeric(logLik(m)), 
                        n = dim(m$frame)[1], 
                        bsum = sum(beta))
      
      return(out)
      
}

eval_aic_m <- function(m) {
      
      # This calculates the log likelihood for the 
      # M model. Adds n (number of obs) for the 
      # model to double check no dropped obs. 
      # Adds bsum to confirm the same input
      # The jacobian correction is added to the log likelihood.
      
      M <- as.numeric(m$frame[[1]])
      beta <- stats::plogis(M * log(2))
      # Jacobian correction
      jac <- -log(log(2)) - log(beta) - log(1-beta)
      # Corrected ll, n, and sum of beta values
      out <- data.frame(ll = as.numeric(logLik(m)) + sum(jac), 
                        n = dim(m$frame)[1], 
                        bsum = sum(beta))
      
       return(out)
      
}




# Setting up the models
bm1 <- seqwrap_compose(
      modelfun = glmmTMB::glmmTMB, 
      data = betadf, 
      metadata = metadata, 
      samplename = "geo_accession", 
      eval_fun = eval_aic_beta,
      arguments = list(
            formula = y ~ time + (1|participant),
            family = glmmTMB::beta_family)
)


mm1 <- seqwrap_compose(
      modelfun = glmmTMB::glmmTMB, 
      data = mdf, 
      metadata = metadata, 
      samplename = "geo_accession", 
      eval_fun = eval_aic_m,
      arguments = list(
            formula = y ~ time + (1|participant))
)




bm1_results <- seqwrap(bm1, 
                       subset = subset$row_id, 
                       return_models = FALSE, 
                       cores = 12)


mm1_results <- seqwrap(mm1, 
                       subset = subset$row_id,
                       return_models = FALSE, 
                       cores = 12)







saveRDS(bm1_results, here::here("analysis/data/derived_data/bm1_results.RDS"))
saveRDS(mm1_results, here::here("analysis/data/derived_data/mm1_results.RDS"))



bm1_sum <- seqwrap_summarise(bm1_results)
mm1_sum <- seqwrap_summarise(mm1_results)



aic_dat <- mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |>  
      pivot_wider(names_from = model, 
                  values_from = c(ll, bsum)) |> 
      # Compare beta sum to check
      mutate(comp = bsum_mm - bsum_bm) |> 
      # Calculate AIC delta
      mutate(delta_aic = -2 * (ll_mm - ll_bm)) |> 
      inner_join(subset |> 
                       rename(target = id)) 
     
   aic_dat |>    
      ggplot(aes(delta_aic)) + geom_line(stat = "density") + 
      facet_wrap(~ betacat) 
      

aicm <- lm(delta_aic ~ 0 + betacat, data = aic_dat)
summary(aicm)


mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |> 
            
      inner_join(subset |> 
                       rename(target = id)) |>  
      
     
      
      ggplot(aes(quantile)) + geom_histogram() + 
      facet_grid(betacat ~ model)


mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |> 
            
      inner_join(subset |> 
                       rename(target = id)) |> 
      inner_join(cden |> 
                       rename(target = id)) |> 
      
      select(target, betacat, quantile, model, intensity_cat) |> 

      ggplot(aes(quantile)) + geom_histogram() + 

      
      facet_wrap(intensity_cat ~ model)






mm1_sum$evaluations |> 
      rename(mm = pval) |> 
      inner_join(
            bm1_sum$evaluation |> 
                  rename(bm = pval)
      ) |> 
       inner_join(subset |> 
                       rename(target = id)) |> 
      ggplot(aes(bm, mm)) + geom_point() + 
      facet_wrap(~ betacat)



bm1_sum$evaluations


m <- bm1_results@models[[1]]
m <- mm1_results@models[[1]]


summary(m)


```






see here https://link.springer.com/article/10.1186/1471-2105-11-587







