---
title: "Case study: DNA methylation (GSE114763)"
format: html
bibliography: "resources/bibliography-methylation.bib"
csl: "resources/bmc-bioinformatics.csl"
editor_options: 
  chunk_output_type: console
---

## Introduction 

The following document outlines a partial reanalysis of data on CpG methylation status in human skeletal muscle in response to resistance exercise training, followed by detraining and subsequent retraining [@seaborne2018;@seaborne2018b]. Most of the relevant code is explicitly shown here, whereas the version of this case study included in the manuscript has been stripped of some details.


## Raw data and pre-processing

```{r}
#| label: dependencies
#| eval: true

library(GEOquery)      # For accessing the raw data
library(minfi)         # For pre-processing of raw data
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(maxprobes)     # For filtering cross-reactive probes 
library(seqwrap)


# Load color scale 
source(here::here("analysis/figures/figure-opts.R"))


```


Raw data are downloaded from the [Gene Expression Omnibus](https://www.ncbi.nlm.nih.gov/geo/) using the accession number [GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763). 


```{r}
#| label: paths
#| eval: true

# Prepare data download. 
# Note for use in the present repo: 
# raw_data and derived data are gitignored so safe for 
# large files. If code-re-use, check the download folder before committing 
# changes to git. 

# The GEO ID
gse_id   <- "GSE114763"

# Set download directories
raw_dir  <- here::here("analysis/data/raw_data", gse_id)
idat_dir <- file.path(raw_dir, "idat")
der_dir  <- here::here("analysis/data/derived_data")

dir.create(idat_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(der_dir,  recursive = TRUE, showWarnings = FALSE)

```

### Downloading data

Sample titles and characteristics are read from the SOFT text view of the series, which returns the metadata of all 40 samples. The supplementary archive (`GSE114763_RAW.tar`, ~722 MB) provides the IDAT files.

```{r}
#| label: download
#| eval: true


# The metadata (pheno_raw) and the raw data IDAT files are cached after 
# downloading once. 

# A function for retrying connection to GEO for downloading large data.
# This was created after issues with downloading the large zipped archive
# from GEO. 
retry_url <- function(expr, tries = 5, pause = 5) {
  call <- substitute(expr)          # re-evaluated on each attempt
  env  <- parent.frame()
  for (i in seq_len(tries)) {
    out <- try(eval(call, env), silent = TRUE)
    if (!inherits(out, "try-error")) return(out)
    if (i < tries) Sys.sleep(pause * i)
  }
  stop("failed after ", tries, " attempts: ",
       conditionMessage(attr(out, "condition")))
}

# Series-level metadata
#
# The metadata contains file names and titles include 
# time-points and participant information. 
pheno_file <- file.path(raw_dir, "pheno_raw.rds")

if (!file.exists(pheno_file)) {
  soft_url <- paste0("https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi",
                     "?acc=", gse_id, "&targ=gsm&form=text&view=brief")

  soft <- retry_url({
    res <- curl::curl_fetch_memory(soft_url)
    if (res$status_code != 200) stop("HTTP ", res$status_code)
    strsplit(rawToChar(res$content), "\r?\n")[[1]]
  })

  # one record per sample; "!Sample_<key> = <value>" lines within each
  kv <- tibble::tibble(record = cumsum(startsWith(soft, "^SAMPLE")),
                       line   = soft) |>
    dplyr::filter(startsWith(line, "!")) |>
    tidyr::separate(line, into = c("key", "value"), sep = " = ",
                    extra = "merge") |>
    dplyr::mutate(key = stringr::str_remove(key, "^!Sample_"))

  # characteristics are themselves "<key>: <value>" pairs
  characteristics <- kv |>
    dplyr::filter(key == "characteristics_ch1") |>
    tidyr::separate(value, into = c("key", "value"), sep = ": ",
                    extra = "merge") |>
    dplyr::filter(key %in% c("gender", "age", "weight", "height")) 

  # Cleaning the metadata for downstream processing. 
  pheno_raw <- kv |>
    dplyr::filter(key %in% c("geo_accession", "title")) |>
    dplyr::bind_rows(characteristics) |>
    tidyr::pivot_wider(id_cols = record, names_from = key,
                       values_from = value) |>
    dplyr::select(-record) |> 

        mutate(title_sep = title) |> 
        separate(title_sep, into = c("seq_sample_id", "timeid"), sep = ":") |>
        mutate(timeid = gsub(" Muscle_", "", timeid)) |> 
        mutate(timeid = gsub("Participant_", "", timeid)) |> 
        mutate(timeid = gsub("7wk_", "7wk", timeid)) |> 
        
         separate(timeid, into = c("time", "participant", "replicate"), sep = "_") |> 
      # Making time-point more obvious
        mutate(time = case_when(
              time == "Baseline" ~ "1_baseline", 
              time == "Acute" ~ "2_acute", 
              time == "7wkLoading" ~ "3_loading", 
              time == "7wkUnloading"  ~ "4_unloading", 
              time == "7wkRe-oading" ~ "5_reloading"
        )) |> 
        # Fixing meta-data details
        mutate(weight = gsub(" kg", "", weight), 
               height = gsub("cm", "", height), 
               weight = as.numeric(weight), 
               height = as.numeric(height), 
               age = as.numeric(age)) |> 
        # Fix a misprinted participant id
        mutate(participant = gsub("P006", "006", participant)) 
  

  saveRDS(pheno_raw, pheno_file)
}

pheno_raw <- readRDS(pheno_file)

# Double check 40 expected samples
stopifnot(nrow(pheno_raw) == 40)

# Download and extract raw IDAT files.
# IDATs are distributed gzipped; minfi reads decompressed files
tar_file <- file.path(raw_dir, paste0(gse_id, "_RAW.tar"))

if (!file.exists(tar_file)) {
  # ~722 MB; increase the timeout for slow connections
  options(timeout = 60 * 60)
  retry_url(GEOquery::getGEOSuppFiles(gse_id, baseDir = dirname(raw_dir),
                                      makeDirectory = TRUE))
}

if (length(list.files(idat_dir, pattern = "\\.idat$")) == 0) {
  utils::untar(tar_file, exdir = idat_dir)

  # Unzip idat files
  gz <- list.files(idat_dir, pattern = "\\.idat\\.gz$", full.names = TRUE)
  invisible(lapply(gz, R.utils::gunzip, overwrite = TRUE, remove = TRUE))
}

```

### Metadata summary

A total of 40 samples were downloaded from the Gene Expression Omnibus ([GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763)). The identified metadata (@tbl-meta-samples) corresponds to the metadata published in [@seaborne2018b] (see table 2 in @seaborne2018b).

Seaborne et al. removed two samples after preliminary analysis, (samples `SkM_Epi_Mem_1` and `SkM_Epi_Mem_39`), we will do the same.


```{r}
#| label: tbl-meta-samples
#| tbl-cap: "Sample metadata retreived from GEO ([GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763)) corresponding to @seaborne2018b".
#| echo: false


library(gt)

pheno_raw |> 
      summarise(.by = participant, 
                weight = mean(weight), 
                height = mean(height), 
                age = mean(age)) |> 
      mutate(participant = gsub("00", "", participant)) |> 
      select(participant, age, height, weight) |> 
      gt() |> 
      cols_label(participant = "Participant", 
                 age = "Age (years)", 
                 weight = "Weight (kg)", 
                 height = "Height (cm)")
      
      
```


## Case study description

```{r}
#| label: case-description
#| echo: false

# Double check gender 
# pheno_raw |> 
#      distinct(gender)
# Confirm all males


# Get sample metadata into text
sample_metadata <- pheno_raw |> 
      
      summarise(.by = participant, 
                weight = mean(weight), 
                height = mean(height), 
                age = mean(age)) |> 
      pivot_longer(weight:age) |> 
      summarise(.by = name, 
                m = mean(value), 
                s = sd(value)) |> 
      mutate(unit = c("kg", "cm", "years")) |> 
      mutate(stat = paste0(sprintf("%.1f", round(m, 1)), 
                           " (", 
                           sprintf("%.1f", round(s, 1)), 
                           ") ", 
                           unit)) 

age <- filter(sample_metadata, name == "age") |> pull(stat)
height <- filter(sample_metadata, name == "height") |> pull(stat)
weight <- filter(sample_metadata, name == "weight") |> pull(stat)

```


> Seaborne et al. [@seaborne2018;@seaborne2018b] investigated whether human skeletal muscle retains an epigenetic memory of earlier hypertrophy-inducing resistance training. For this aim, previously untrained male participants (age, [mean (SD)] `r age`; weight, `r weight`; height, `r height`) completed a sequential design with two seven-week periods of resistance-exercise training (RET) intersected with a 7-week unloading period without RET. RET periods included two weekly sessions targeting lower limbs (seven exercises) and one weekly session targeting the upper body (5 exercises; 4 sets of 10 repetitions per exercise). Skeletal muscle biopsies were sampled from m. vastus lateralis on five occasions: before the intervention, 30 min after the first exercise session, and after the completion of each of the 7-week periods. Genomic DNA was extracted from muscle biopsy samples (DNeasy Kit, Qiagen, Manchester, UK) and bisulfite converted before amplification, fragmentation, and hybridization using the Illumina Infinium Human MethylationEPIC Beadchip protocol [@seaborne2018;@seaborne2018b]. 


The Sentrix slide (chip) and array position are recovered from the IDAT file names (e.g. `GSM3149860_201496860025_R01C01_Grn.idat`) and retained as potential technical covariates in subsequent analyses. `Basename` is the file stem that `minfi` uses to locate the red/green channel pair.

The samples are distributed as described in [@seaborne2018;@seaborne2018b] with one participant dropping out from the study prior to the last sampling time-point. One participant have duplicate arrays at one time-point. After removal of samples based on preliminary QC-analysis (see @seaborne2018b), the full data set contains 38 samples with no technical duplicates. 

```{r}
#| label: tbl-metadata-count
#| eval: true
#| tbl-cap: "Samples per time-point and participant."


idat_stems <- list.files(idat_dir, pattern = "_Grn\\.idat$") |>
  stringr::str_remove("_Grn\\.idat$")

array_info <- tibble::tibble(stem = idat_stems) |>
  tidyr::separate(stem, into = c("geo_accession", "slide", "array"),
                  sep = "_", remove = FALSE) |>
  dplyr::mutate(Basename = file.path(idat_dir, stem))

metadata <- pheno_raw |>
      # Removing outlier sample, see "Quality Control of DNA Methylation Probes"
      # in @seaborne2018b
      filter(!grepl("SkM_Epi_Mem_1:", title), 
             !grepl("SkM_Epi_Mem_39:", title)) |> 
      
  tibble::as_tibble() |>
  dplyr::select(geo_accession,
                title,
                participant,
                gender,
                age, 
                time) |>
  dplyr::mutate(
    label = stringr::str_remove(title, "^SkM_Epi_Mem_\\d+:\\s*")) |>
  dplyr::left_join(array_info, by = "geo_accession")

stopifnot(!anyNA(metadata$time), !anyNA(metadata$participant))

# 38 arrays: 8 participants x 5 time points with missing values for on 
# participant.

metadata |> 
      summarise(.by = c(time, participant), 
                n = n()) |> 
      pivot_wider(names_from = time, values_from = n) |> 
      select(Participant = participant, 
             Baseline = `1_baseline`, 
             Acute = `2_acute`, 
             Loading = `3_loading`, 
             Unloading = `4_unloading`, 
             Reloading = `5_reloading`) |> 
      gt() |> 
      tab_spanner(label = "Sample time-points", columns = "Baseline":"Reloading")
      



```

### Reading IDAT and array-level quality control

We will use the `minfi` package [@fortin2014] to read the raw data and perform quality control, filtering and normalization. As a first step, array-level filtering is done using detection p-values where methylated and unmethylated positions are compared to negative control positions. This comparison produces a p-value used to identify problematic genomic positions. As shown below (@fig-array-qc), array-level aggregates show mean values < 0.01.    


```{r}
#| label: fig-array-qc
#| fig-cap: "Array-level quality control. Mean detection p-values and fraction of failed genomic positions per array subdivided by time-point."
#| eval: true

rgset <- minfi::read.metharray(metadata$Basename, extended = TRUE, force = TRUE)

# Check that orders correspond
stopifnot(identical(colnames(rgset), basename(metadata$Basename)))

colnames(rgset) <- metadata$geo_accession


det_p <- minfi::detectionP(rgset)

# 
qc <- tibble::tibble(
  geo_accession = colnames(det_p),
  mean_det_p    = colMeans(det_p),
  frac_failed   = colMeans(det_p > 0.01))

qc |> 
      inner_join(pheno_raw |> 
                       select(geo_accession, time)) |> 
      ggplot(aes(mean_det_p, frac_failed, color = time)) + 
      geom_point() +
      labs(x = "Array-level mean detection p-value", 
           y = "Fraction of failed genomic positions (p > 0.01)", 
           color = "Time-point")
```



```{r}

# Below is the suggested workflow for confirming sex. We expect only males and 
# the function indficate that only one sex is present which should suffice as 
# confirmation.
# predicted   <- minfi::getSex(minfi::mapToGenome(msetraw))
# qc$pred_sex <- predicted$predictedSex

# All samples pass quality control
keep_samples <- qc$geo_accession[qc$mean_det_p < 0.01]
# length(keep_samples)

rgset    <- rgset[, keep_samples]
det_p    <- det_p[, keep_samples]
metadata <- metadata |> dplyr::filter(geo_accession %in% keep_samples)
```

### Normalization and probe filtering

Functional normalization [@fortin2014] is used because the study design can be expected to result in global methylation differences. Functional normalization removes technical variation captured by the array control probes without assuming a common methylation distribution across samples [@fortin2014]. Probes are then removed if they (1) fail detection in any retained sample, (2) overlap common SNPs at the target CpG or single-base extension site, (3) appear on published lists of cross-reactive EPIC probes (using the [`maxprobes`](https://github.com/markgene/maxprobes/) package), or (iv) map to the sex chromosomes (all male cohort).

```{r}
#| label: normalise-filter
#| eval: true



# Functional normalization
gset <- minfi::preprocessFunnorm(rgset, ratioConvert = FALSE)

# Initial number of probes
nprobes_init <- dim(gset)[1]

## (i) detection
det_p  <- det_p[match(minfi::featureNames(gset), rownames(det_p)), ]
keep   <- rowSums(det_p < 0.01) == ncol(gset)
gset   <- gset[keep, ]

# Number of probes after filtering
nprobes_detectionfilter <- dim(gset)[1]

## (ii) SNP-affected probes 
gset <- minfi::dropLociWithSnps(gset, snps = c("SBE", "CpG"), maf = 0)

nprobes_snp <- dim(gset)[1]

## (iii) cross-reactive probes 
gset <- maxprobes::dropXreactiveLoci(gset)

nprobes_cross <- dim(gset)[1]

## (iv) sex chromosomes 
anno <- minfi::getAnnotation(gset)
gset <- gset[!anno$chr %in% c("chrX", "chrY"), ]

nprobes_final <- dim(gset)[1]


## Clean up memory 
rm(rgset); gc()



```

`r nprobes_init` normalized probes were filtered based on detection which retained `r nprobes_detectionfilter`, after SNP-overlap (n = `r nprobes_snp`), cross-reactivity (n = `r nprobes_cross`) and sex-chromosome mapping filtering, n = `r nprobes_final` remained in the data set. 

### Calculation of methylation quantities

An Illumina array detecting methylation values produces arbitrary signals for the unmethylated (U) and methylated (M) portion of the i^th^ genomic position. Converting these two values to a methylation score is done trough the calculation of a &beta; value.

$$
\beta_i = \frac{M_i}{M_i+U_i + c}
$$
The bias constant c, is added to the denominator to avoid large variations in the &beta; score when probe intensities are low [@bibikova2006]. This value was set to $c = 100$ to achieve regularization of the &beta; values in cases where both M and U were small [@bibikova2006], which is also the default in the `minfi` package when `type = "illumina"`. However, the successful choice of c is noted to depend on the pre-processing of the data (see `?minfi::getBeta`). 

Inspecting the normalized signals representing methylated and unmethylated probes indicate that the constant will only have a minor impact on most sites as both *M* and *U* are >> 100 @fig-raw-data-characteristics A.   

```{r}
#| label: fig-raw-data-characteristics
#| fig-cap: "Methylated and Unmethylated signals accross samples and a subset of the genomic sites (n = 1000) shows that most signals are well above 100 (A). Calculation of beta; values with different selections of *c* indicate minor differences in the overall distribution of values (B). Both figures have the relative density on the y axis (not shown)"

# Extract the methylated and unmethylated signals
M <- minfi::getMeth(gset)
U <- minfi::getUnmeth(gset)

# Set the number of probes
n <- 1000

# Figure A: Methylated and Unmethylated signals accross samples. 
set.seed(1)

# Sample sites 
sites <- sample(rownames(M), n)

figa <- data.frame(M)[sites,] |> 

   
      tibble::rownames_to_column(var = "id") |> 
      
      pivot_longer(starts_with("GSM"), values_to = "M") |> 
      
      inner_join(
            data.frame(U)[sites,] |> 
                  tibble::rownames_to_column(var = "id") |> 
                  pivot_longer(starts_with("GSM"), values_to = "U")     
      ) |> 
      pivot_longer(cols = M:U, 
                   names_to = "signal") |> 
      ggplot(aes(value, group = paste0(signal,name), 
                 color = signal)) + 
      geom_line(stat = "density") +
      scale_x_log10()  + 
      theme_classic() + 
      theme(axis.title.y = element_blank(), 
            axis.ticks.y = element_blank(), 
            axis.line.y = element_blank(), 
            axis.text.y = element_blank()) + 
      scale_color_manual(values = colors) + 
      labs(x = "Normalized signal (AU)", 
           color = "Signal type")


figb <- data.frame(M)[sites,] |> 
 
   
      tibble::rownames_to_column(var = "id") |> 
      
      pivot_longer(starts_with("GSM"), values_to = "M") |> 
      
      inner_join(
            data.frame(U)[sites,] |> 

                  tibble::rownames_to_column(var = "id") |> 
                  pivot_longer(starts_with("GSM"), values_to = "U")     
      ) |> 
      
     mutate(beta0 = M / (M + U + 0), 
            beta50 = M / (M + U + 50),
            beta100 = M / (M + U + 100)) |> 
      dplyr::select(id, name, beta0:beta100) |>

      pivot_longer(cols = starts_with("beta"), 
                   names_to = "constant", 
                   values_to = "beta") |> 
      mutate(constant = as.numeric(gsub("beta", "", constant))) |> 
      
      ggplot(aes(beta, group = constant, color = as.factor(constant))) +
      geom_line(stat = "density") +
       theme_classic() + 
      theme(axis.title.y = element_blank(), 
            axis.ticks.y = element_blank(), 
            axis.line.y = element_blank(), 
            axis.text.y = element_blank(), 
            axis.title.x = ggtext::element_markdown()) + 
      scale_color_manual(values = colors) + 
      labs(x = "&beta;-values", 
           color = "Offset constant")
      

cowplot::plot_grid(figa, figb, ncol = 1)

```

We decide to go for the default value of *c* = 100 for the calculation of the &beta; values. 
&beta;-values are restricted to the 0 to 1 space, and although this is an intuitive metric for understanding the methylation status at a specific genomic site (e.g., as a percentage methylation), it is an inconvenient quantity for a statistical model that expects normally distributed errors [@laird2010; @du2010]. Indeed, when applying a Gaussian model to identify differentially methylated genomic sites, a logit transformation of the &beta;-values results in more favorable diagnostics [@du2010]. The logit transformation of the &beta; values results in M-values. 

$$
M = \log_2 \left( \frac{\beta}{1-\beta} \right)
$$

The logit transformation (using a base 2 logarithm) puts the methylation values on a scale with no upper or lower bounds (-ininf;, &ininf;, @fig-beta-mval). This transformation is done on the &beta;-values directly to avoid missing c in the calculation.  


```{r}
#| label: fig-beta-mval
#| fig-cap: "The logit transform of &beta; values gives M-values."
#| echo: false


beta <- seq(from = 0.001, to = 0.999, by = 0.001)
Mval <- log2(beta/(1-beta))

data.frame(Mval, beta) |> 
      ggplot(aes(Mval, beta)) + geom_line() + 
      theme_classic() + 
      labs(x = "M-values", y = "&beta;-values") + 
      theme(axis.title.y = ggtext::element_markdown())

```

### Type I and II error rates permutation study

Next, we will compare the models using Type I and II error rates determined from permutation of within-participant relabeled data. This relabeling establishes a null distribution for statistical tests performed using the *within-participant design*. The design keeps the within-participant correlation.

We will perform the experiment on permutated data from a subset of genomic sites distributed across different strata of &beta; values. 

The permutation loop below produces two results from a single set of model fits. A fixed number of sites is sampled from each &beta;-value stratum. Within every permutation, a subset of those sites receives a *known* effect (adding a known average effect) while the remainder are left untouched. Sites without an effect estimate the *type I error rate*, sites with an effect estimate *power*. This will also give us an estimate of the *false discovery proportion*.

We will use the *M-value scale* to add the effect to permutated data with effect assignment *redrawn in every permutation*. The &beta;-values are derived from the manipulated M-values. A shift of &delta; M units is exactly the fixed-effect parameter of the Gaussian model and proportionally to &delta;&times;log(2) for the linear-predictor shift of the &beta; model with the logit link. This means that no model is favored by the parameterization. 

In the code chunk below, a single site is used to show the workflow. &beta; values are calculated from the `gset` object and transformed to M values. We are using the first row of the M valu matrix and join the values with the meta data.  

Within each participant, time is re-sampled to create a permuted time factor. Using the permuted time factor, an average effect (&delta;=0.8) is added to all observations belonging to the permuted time-point `3_loading`. Data are back-transformed to the &beta; scale and both models are fitted.  


```{r}

beta_vals <- getBeta(gset, offset = 100)
m_vals <- log2(beta_vals / (1 - beta_vals))

y <- data.frame(y = m_vals[1,]) |> 
      tibble::rownames_to_column(var = "geo_accession" ) 

set.seed(1)
dat <- metadata |> 
      select(geo_accession, participant, time) |> 
      inner_join(y) |>
      group_by(participant) |> 
      mutate(time_permute = sample(time, replace = FALSE)) |> 
      ungroup() |> 
      mutate(msim = if_else(time_permute == "3_loading", 
             y + 0.8, y), 
             betasim = plogis(msim * log(2))) 


m_mod <- glmmTMB::glmmTMB(msim ~ time_permute + (1|participant), 
                          data = dat)

beta_mod <- glmmTMB::glmmTMB(betasim ~ time_permute + (1|participant), 
                          data = dat, 
                          family = glmmTMB::beta_family)

es_mmod <- coef(summary(m_mod))$cond["time_permute3_loading","Estimate"]
es_betamod<- coef(summary(beta_mod))$cond["time_permute3_loading","Estimate"]


es_betamod <- es_betamod / log(2)
```

The estimated effects (`r round(es_mmod,2)` and `r round(es_betamod,2)`) for the model on M-values and &beta; values respectively are comparable but will differ due to compression in the &beta; values model.

The fixed effect sizes are multiplied by 1 or -1 in a balanced fashion to create the direction of the effect. At an extreme site (&beta;-values close to 0 or 1), adding or subtracting &delta; will either compress towards the boundary or move estimates towards 0.5. Since these scenarios are not equally detectable and the two models disagree about the variance function in extreme regions, we can evaluate the effect of selecting one model over the other.

The source code for estimation of Type I and II error rates using a permutation experiment can be run in full using the script `analysis/R/methylation-error-permutation.R`. Below we check if files are available from a previous run, if not the script runs and stores data in `analysis/data/derived_data/permutation/`. 


```{r}

# list files in permutation results folder
out_dir <- here::here("analysis/data/derived_data/permutation")
perm_files <- list.files(out_dir, pattern = "^perm_\\d+\\.RDS$", full.names = TRUE)

if (length(perm_files) == 0) source(
      here::here("analysis/R/methylation-error-permutation.R")
      )

```

### Loading permutation study results

@tbl-permutation-evals shows the number of failed models among all model fits for the two model types. A total of 250 sets of permutations where performed with 1000 sites in each set over 5 &beta; value strata.


```{r}
#| label: reading-permutation-data
#| eval: true
#| label: tbl-permutation-evals

### Reading the results back

spike_time <- "3_loading"

perm_files <- list.files(out_dir, pattern = "^perm_\\d+\\.RDS$", full.names = TRUE)
perm       <- lapply(perm_files, readRDS)

perm_summaries <- dplyr::bind_rows(lapply(perm, `[[`, "summaries"))
perm_designs   <- dplyr::bind_rows(lapply(perm, `[[`, "design"))
perm_evals     <- dplyr::bind_rows(lapply(perm, `[[`, "evaluations"))

# Which coefficients with the simulated effect.
#
# The added effect shifts only the samples labelled `spike_time` under the 
# permuted labels. `spike_time` is set to 3_loading and
# 1_baseline is the reference level.

# alt_terms also accepts and omnibus test (not used here)
alt_terms <- c(paste0("time_permute", spike_time), "time_omnibus")

# Attach the per-iteration effects (delta, sign, stratum) to every test.
# is_alt is ground truth for a (site, term) pair, not for a site.
perm_dat <- perm_summaries |>
      dplyr::inner_join(perm_designs,
                        by = c("iter", "target" = "id")) |>
      # delta is signed, so the test must be != 0 -- `delta > 0` would classify
      # every downward-spiked site as a null and contaminate the reference.
      dplyr::mutate(is_alt = delta != 0 & term %in% alt_terms)

stopifnot(any(perm_dat$is_alt))

# Convergence failures are dropped silently by seqwrap_summarise (NULL entries
# are removed before rbind), so differential failure between the two models
# would quietly change what is being compared.

perm_evals |>
      dplyr::summarise(.by = c(model),
                       n_fits     = dplyr::n(),
                       n_fail     = sum(convergence != 0 | !pdHess),
                       prop_fail  = mean(convergence != 0 | !pdHess)) |> 
      gt()
```


### Calibration under the null

The type I error rates are estimated from sites without added effects. The unaffected contrasts on a site with an added effect are nominally null, but a spiked site is not exchangeable with an unspiked one on *any* contrast: adding the effect changes the fitted mean of one time-point, and because the design is unbalanced (8/8/8/7/7 samples per time-point, with one participant contributing only three samples) the shift is not orthogonal to the estimated variance components. Under the &beta; likelihood the variance is in addition a function of the mean, so the common dispersion parameter, and with it the other contrasts, is affected as well. 

```{r}
#| label: fig-permutation-type1
#| eval: true

null_dat <- perm_dat |> dplyr::filter(delta == 0)

t1e <- null_dat |>
      dplyr::summarise(.by = c(model, term, betacat),
                       n      = dplyr::n(),
                       a_05   = mean(p.value < 0.05, na.rm = TRUE),
                       a_01   = mean(p.value < 0.01, na.rm = TRUE),
                       a_001  = mean(p.value < 0.001, na.rm = TRUE)) 


t1e |> 
      ggplot(aes(betacat,  100 * a_05, fill = model))  +
      geom_point(shape = 21, 
                 position = position_jitter(width = 0.1)) +
      scale_y_continuous(limits = c(0, 10)) +
      geom_hline(yintercept = 5, color = "gray", 
                 lty = 2) +
      theme_classic() + 
      labs(x = "&beta; strata", 
           y = "Type I error rates (%)") + 
      theme(axis.title.x = ggtext::element_markdown())
      


# Null p-value distribution by stratum


#null_dat |>
#      ggplot(aes(sample = p.value)) +
#      stat_qq(distribution = stats::qunif) +
#      stat_qq_line(distribution = stats::qunif) +
#      facet_grid(betacat ~ model) +
#      labs(x = "Expected quantile (uniform)", y = "Observed p-value")

```

The two models are both miscalibrated (~ 2-3 %-points) in terms of type I errors (@fig-permutation-type1).

The miscalibration is not confined to conventional significance levels. It grows as the threshold is lowered, and it is in the extreme tail that a multiple-testing correction operates (@tbl-permutation-nulltail).

```{r}
#| label: tbl-permutation-nulltail
#| tbl-cap: "Observed proportion of null p-values below increasingly stringent thresholds. Valid p-values would give `observed` equal to `threshold`; `inflation` is the ratio of the two. The Benjamini-Hochberg step at *q* = 0.05 over a family of *m* tests cannot reject at all unless the smallest p-value falls below *q*/*m*, which is 6.25 &times; 10<sup>-5</sup> for the families used below."
#| eval: true

# Complete-null tests: contrasts that never receive a spike, on sites that
# carry no effect on any contrast.
null_p <- perm_dat |>
      dplyr::filter(!term %in% alt_terms, delta == 0)

dplyr::bind_rows(lapply(c(1e-3, 1e-4, 6.25e-5, 1e-5), function(t)
      null_p |>
            dplyr::summarise(.by = model,
                             threshold = t,
                             observed  = mean(p.value < t, na.rm = TRUE),
                             inflation = mean(p.value < t, na.rm = TRUE) / t))) |>
      tidyr::pivot_wider(names_from = model,
                         values_from = c(observed, inflation)) |>
      gt() |>
      gt::fmt_scientific(columns = c(threshold, dplyr::starts_with("observed")),
                         decimals = 2) |>
      gt::fmt_number(columns = dplyr::starts_with("inflation"), decimals = 1)
```

The consequence can be read directly from the contrasts that never receive a simulated effect. Once spiked sites are removed these form a *complete null*: every hypothesis in the family is true, so every rejection is false, FDP is 1 whenever anything is rejected, and the false discovery rate reduces to the probability of any rejection at all. Benjamini-Hochberg bounds this by *q*, and the bound is essentially attained under valid p-values -- for *m* = 800 and *q* = 0.05, $1 - (1 - q/m)^m$ = 0.049 (@tbl-permutation-nullbh).

```{r}
#| label: tbl-permutation-nullbh
#| tbl-cap: "Benjamini-Hochberg applied to complete-null families of `size * 5 * (1 - pi1)` sites, one per model, contrast and permutation. `p_any` is the realised false discovery rate and is bounded by *q* = 0.05. `p_any_indep` is what `mean_R` would imply if the tests within a family were independent; the gap between the two measures how strongly rejections cluster. `p_any_t` repeats the identical procedure after replacing the Wald *z* reference with a *t* reference on the effective degrees of freedom."
#| eval: true

null_families <- null_p |>
      # BH must be re-run on this subset: the stored qval was adjusted against
      # the full `size * 5` family and no longer applies. The same tests are
      # also referred to a t distribution on the effective degrees of freedom
      # (df = 12, the value that restores nominal size above) rather than the
      # Wald z that glmmTMB reports, holding everything else fixed.
      dplyr::mutate(.by = c(model, term, iter),
                    qval_null = stats::p.adjust(p.value, method = "BH"),
                    p_t       = 2 * stats::pt(-abs(estimate / std.error), df = 12),
                    qval_t    = stats::p.adjust(p_t, method = "BH")) |>
      dplyr::summarise(.by = c(model, term, iter),
                       m_tests = dplyr::n(),
                       R       = sum(qval_null < 0.05, na.rm = TRUE),
                       R_t     = sum(qval_t < 0.05, na.rm = TRUE))

null_families |>
      dplyr::summarise(.by = model,
                       n_families = dplyr::n(),
                       m_tests    = unique(m_tests),
                       mean_R     = mean(R),
                       p_any      = mean(R > 0),
                       # Under independence P(any) would follow from mean_R
                       # alone; the observed value is far smaller because all
                       # sites in a family share one permuted design.
                       p_any_indep     = 1 - exp(-mean(R)),
                       max_R           = max(R),
                       mean_R_when_any = mean(R[R > 0]),
                       p_any_t         = mean(R_t > 0)) |>
      gt() |>
      gt::fmt_number(decimals = 3)
```

Both models reject at least one true null in roughly a third of permutations, against a bound of 5 %. The failures are not a steady trickle: most families contain no rejection at all, while those that do contain several at once and occasionally more than a hundred, because every site in a family is fit against the same relabelling. Referring the identical tests to a *t* distribution on the effective degrees of freedom removes the excess entirely, which identifies the Wald reference distribution rather than the multiplicity correction as the cause. This is a property of Wald inference in a mixed model with eight participants, and both scales inherit it equally.


### Power at matched empirical size

Comparing power at the nominal 5% level is confounded by test size: a model whose actual type I error rate is 8% will appear more powerful simply because it calls more sites as differentially methylated irrespective of the simulated truth. Here we use the empirical 5% quantile of the null p-values, taken per model, contrast and stratum from the same permutations, as the critical value. Both models are therefore compared while operating at exactly 5% empirical size. Note that this matches the models on type I error rather than on the false discovery rate, which depends on the proportion of true alternatives in the family and is therefore a property of the experiment rather than of the model.

```{r}
#| label: fig-permutation-power
#| eval: true

crit <- null_dat |>
      dplyr::summarise(.by = c(model, term, betacat),
                       crit = stats::quantile(p.value, 0.05, na.rm = TRUE))

# is_alt, not delta != 0: only the contrast against spike_time (and the omnibus
# test) carries the effect. Grouping must keep `term`, otherwise the spiked
# contrast is pooled with the three unaffected ones and power is diluted towards
# the nominal level. `delta` is signed here, so upward and downward spikes are
# reported separately -- keep `delta_abs` too for the pooled view.
power_dat <- perm_dat |>
      dplyr::filter(is_alt) |>
      dplyr::inner_join(crit, by = c("model", "term", "betacat")) |>
      dplyr::summarise(.by = c(model, term, betacat, delta, delta_abs),
                       n             = dplyr::n(),
                       power_nominal = mean(p.value < 0.05, na.rm = TRUE),
                       power_matched = mean(p.value < crit, na.rm = TRUE))

# Power against the signed effect. At the extreme strata the two directions are
# not equally detectable -- an upward spike at an xhigh site compresses towards
# the boundary -- so asymmetry about zero is a result, not noise.
p1 <- power_dat |>
      ggplot(aes(delta, power_matched, colour = model)) +
      geom_line(position = position_dodge(width = 0.02)) +
      geom_point(position = position_dodge(width = 0.02)) +
      facet_grid( ~ betacat) +
      scale_colour_manual(values = colors) +
      labs(x = "Simulated effect (M-value units, signed)",
           y = "Power at 5% empirical size",
           colour = "Model")

# Pooled over direction
p2 <- power_dat |>
      dplyr::summarise(.by = c(model, term, betacat, delta_abs),
                       power_matched = stats::weighted.mean(power_matched, n),
                       n = sum(n)) |>
      ggplot(aes(delta_abs, power_matched, colour = model)) +
      geom_line(position = position_dodge(width = 0.02)) +
      geom_point(position = position_dodge(width = 0.02)) +
      facet_grid(term ~ betacat) +
      scale_colour_manual(values = colors) +
      labs(x = "Simulated effect (M-value units, magnitude)",
           y = "Power at 5% empirical size",
           colour = "Model")


cowplot::plot_grid(p1, p2, ncol = 1)

```

Models are comparable in terms of power (@fig-permutation-power). 



### Agreement between the two models

Type I error rates and power are averages over sites, and two models can match on both while disagreeing site by site. The permutation output allows the stronger, paired comparison: every site in every permutation is fit by both models, so their estimates and their p-values can be compared directly.

```{r}
#| label: tbl-permutation-agreement
#| tbl-cap: "Paired agreement between the two models on the spiked contrast, by &beta; stratum, separately for sites that received an effect and sites that did not. Estimates are compared after returning the &beta; coefficient to M-value units. `mad_` columns are mean absolute differences and `p90_logp` is the 90th percentile of the absolute difference in -log<sub>10</sub>(p). `discordant` is the proportion of sites the two models classify differently with each operating at its own 5% empirical size."
#| eval: true

# Each model is held at its own size-matched critical value, so a discordant
# call reflects the models disagreeing rather than one of them rejecting more.
crit_w <- crit |>
      dplyr::filter(term == paste0("time_permute", spike_time)) |>
      dplyr::select(-term) |>
      tidyr::pivot_wider(names_from = model, values_from = crit,
                         names_prefix = "crit_")

agree <- perm_dat |>
      dplyr::filter(term == paste0("time_permute", spike_time)) |>
      dplyr::select(iter, target, betacat, delta, model, estimate, p.value) |>
      tidyr::pivot_wider(names_from = model, values_from = c(estimate, p.value)) |>
      dplyr::inner_join(crit_w, by = "betacat") |>
      # The beta coefficient is a shift on the natural-log logit scale; dividing
      # by log(2) returns it to M-value units and makes the two comparable.
      dplyr::mutate(estimate_beta_M = estimate_beta / log(2),
                    lp_beta = -log10(p.value_beta),
                    lp_m    = -log10(p.value_m),
                    truth   = ifelse(delta == 0, "null", "spiked"))

agree |>
      dplyr::summarise(.by = c(truth, betacat),
                       n            = dplyr::n(),
                       r_estimate   = cor(estimate_beta_M, estimate_m),
                       mad_estimate = mean(abs(estimate_beta_M - estimate_m)),
                       r_logp       = cor(lp_beta, lp_m),
                       mad_logp     = mean(abs(lp_beta - lp_m)),
                       p90_logp     = stats::quantile(abs(lp_beta - lp_m), 0.9),
                       discordant   = mean((p.value_beta < crit_beta) !=
                                           (p.value_m < crit_m))) |>
      dplyr::arrange(truth, factor(betacat,
                                   c("xlow", "low", "mid", "high", "xhigh"))) |>
      gt() |>
      gt::fmt_number(decimals = 4)
```

Once the &beta; coefficient is returned to M-value units the two models recover the same effect: on sites carrying an effect the estimates correlate at *r* &ge; 0.9987 in every stratum and differ by 0.005 to 0.013 M units on average, against simulated effects of 0.1 to 0.8 (@tbl-permutation-agreement). The disagreement between the models is confined to the standard errors, and therefore to the p-values.

That disagreement is smallest in the middle of the &beta; range and grows towards both boundaries. On spiked sites the mean absolute difference in -log~10~(p) is 0.098 at `mid` against 0.46 at `xlow` and 0.58 at `xhigh`, and for the most extreme decile of `xhigh` sites the two models differ by more than 1.8 orders of magnitude (@fig-permutation-agreement). It also scales with the size of the effect: on unspiked sites the same quantity is 0.006 at `mid` and 0.020 to 0.037 elsewhere, roughly an order of magnitude smaller throughout. Both patterns are what the &beta; variance function predicts. The &beta; likelihood weights an observation by $\mu(1-\mu)/(1+\phi)$ while the Gaussian model weights every observation equally, so the two separate only where the fitted mean approaches 0 or 1 -- and a simulated effect at an already extreme site pushes it further into exactly that region.

```{r}
#| label: fig-permutation-agreement
#| fig-cap: "Paired -log<sub>10</sub>(p) for sites carrying a simulated effect, by &beta; stratum and by the direction of the effect. The line is the identity. The two models agree closely in the middle strata and fan out symmetrically towards both &beta; boundaries."
#| eval: true

agree |>
      dplyr::filter(truth == "spiked") |>
      dplyr::mutate(betacat = factor(betacat,
                                     c("xlow", "low", "mid", "high", "xhigh")),
                    direction = factor(sign(delta), levels = c(-1, 1),
                                       labels = c("Downward", "Upward"))) |>
      ggplot(aes(lp_m, lp_beta, color = direction)) +
      geom_abline(slope = 1, intercept = 0, colour = "gray") +
      geom_point(alpha = 0.2) +
      facet_wrap(~ betacat, nrow = 1) +
      scale_colour_manual(values = colors) +
      theme_classic() +
      labs(x = "M-value model, -log<sub>10</sub>(p)",
           y = "&beta; model, -log<sub>10</sub>(p)",
           colour = "Effect direction") +
      theme(axis.title.x = ggtext::element_markdown(),
            axis.title.y = ggtext::element_markdown())
```

The practical consequence is modest. With each model held at its own 5 % empirical size, the two disagree about whether to call a spiked site in 0.35 % of cases at `mid`, rising to 0.8 to 1.8 % in the other strata -- the same pattern, two- to fivefold, but on a base rate low enough that the choice of scale rarely changes a conclusion. The models differ in how confidently they speak, not in what they say.



## Full model run

```{r}

# The annotation file
gset <- readRDS(here::here("analysis/data/derived_data/seaborne-gset-normalized.RDS"))
anno <- minfi::getAnnotation(gset)
nsites <- nrow(anno)

```


The unfiltered Illumina EPIC data has over 800 000 CpG positions. After filtering the available dataset contained `r nsites` CpG positions. Since the strategy here is to detect differentially methylated positions (DMP) we will fit a model per available position using the &beta;- and M-values and their respective model (generalized linear model with &beta; and normal likelihoods, respectively). 


### Computational time

```{r}
#| echo: false


# Elapsed time
beta_model_full <- readRDS(
      here::here("analysis/data/derived_data/beta-model-full.RDS"))
m_model_full <- readRDS(
      here::here("analysis/data/derived_data/m-model-full.RDS"))

beta_time <- beta_model_full@elapsed_time
m_time <- m_model_full@elapsed_time

beta_time <- round(beta_time[3] /60 /60, 1)
m_time <- round(m_time[3] /60 /60, 1)

# Number of sucessfull models 


############# NEEDS FIX ##########################

beta_model_full@errors



m_model_full@errors |> 
      distinct(message)
      print()


```

Running the full model took `r beta_time` and `r m_time` hours for the &beta; and M-models, respectively using 12 cores on a laptop computer (AMD Ryzen 5 7540U, 3.2 GHz, 16 GB RAM).

The &beta;-regression model returned estimates for all `r beta_errors_sum[1]` sites while the model on M-values flagged `r m_errors_sum[3]` models for convergence issues.



```{r}


m_model_sum <- readRDS(here::here("analysis/data/derived_data/m-model-full-sum.RDS"))
beta_model_sum <- readRDS(here::here("analysis/data/derived_data/beta-model-full-sum.RDS"))


beta_model_sum$summaries |> 
      filter(grepl("time", term)) |> 

      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time3_loading")  |> 
      filter(qval < 0.05) |> 
      arrange(qval)
       




m_model_sum$summaries |> 
      filter(grepl("time", term)) |> 
      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time3_loading") |> 
      filter(qval < 0.05) |> 
      arrange(qval)
   

# Read seaborn data 
sdat <- readxl::read_excel(
      path = here::here("analysis/data/resource_data/seaborne2018-suppl.xlsx"), 
                        sheet = 2) |> 
      select(target = `Column ID`, 
             p.value = `p-value(Time)`, 
             ratio_estimate = `MeanRatio(Baseline vs. Chronic Loading)`, 
             estimate = `Estimate(Baseline vs. Chronic Loading)`
             ) |> 
      print()

levels(factor(metadata$time))

  supp <- readxl::read_excel(
      path = here::here("analysis/data/resource_data/seaborne2018-suppl.xlsx"), sheet = 2)
  theirs <- supp |>
    select(target = `Column ID`,
           their_est = `Estimate(Baseline vs. Chronic Loading)`) |>
    mutate(their_dir = -sign(their_est))      # flip to loading-minus-baseline   

  ours <- m_model_sum$summaries |>
    filter(effect == "fixed", component == "cond", term == "time3_loading") |>   
    select(target, estimate)

  cmp <- inner_join(ours, theirs, by = "target")

  mean(sign(cmp$estimate) == cmp$their_dir)      
  cor(cmp$estimate, -cmp$their_est, method = "spearman")
  table(sign(cmp$estimate), cmp$their_dir)
            
            
      ggplot(aes(p.value, org.pval)) + geom_point()


      dplyr::select(target, term, estimate, std.error, p.value) |> 
      mutate(.by = term, 
             qval = p.adjust(p.value, method = "fdr")) |> 
      
      filter(term == "time4_unloading") |> 

      filter(target == "cg15154546")
      arrange(p.value) |> 
      print()



```





















## AIC


```{r}


# Evaluation function with corrected (Jacobian) AIC for model comparison
eval_aic_beta <- function(m) {
    
      # This calculates the log likelihood for the 
      # beta model. Adds n (number of obs) for the 
      # model to double check no dropped obs. 
      # Adds bsum to confirm the same input
      
      beta <- as.numeric(m$frame[[1]])
      
      out <- data.frame(ll = as.numeric(logLik(m)), 
                        n = dim(m$frame)[1], 
                        bsum = sum(beta))
      
      return(out)
      
}

eval_aic_m <- function(m) {
      
      # This calculates the log likelihood for the 
      # M model. Adds n (number of obs) for the 
      # model to double check no dropped obs. 
      # Adds bsum to confirm the same input
      # The jacobian correction is added to the log likelihood.
      
      M <- as.numeric(m$frame[[1]])
      beta <- stats::plogis(M * log(2))
      # Jacobian correction
      jac <- -log(log(2)) - log(beta) - log(1-beta)
      # Corrected ll, n, and sum of beta values
      out <- data.frame(ll = as.numeric(logLik(m)) + sum(jac), 
                        n = dim(m$frame)[1], 
                        bsum = sum(beta))
      
       return(out)
      
}




# Setting up the models
bm1 <- seqwrap_compose(
      modelfun = glmmTMB::glmmTMB, 
      data = betadf, 
      metadata = metadata, 
      samplename = "geo_accession", 
      eval_fun = eval_aic_beta,
      arguments = list(
            formula = y ~ time + (1|participant),
            family = glmmTMB::beta_family)
)


mm1 <- seqwrap_compose(
      modelfun = glmmTMB::glmmTMB, 
      data = mdf, 
      metadata = metadata, 
      samplename = "geo_accession", 
      eval_fun = eval_aic_m,
      arguments = list(
            formula = y ~ time + (1|participant))
)




bm1_results <- seqwrap(bm1, 
                       subset = subset$row_id, 
                       return_models = FALSE, 
                       cores = 12)


mm1_results <- seqwrap(mm1, 
                       subset = subset$row_id,
                       return_models = FALSE, 
                       cores = 12)







saveRDS(bm1_results, here::here("analysis/data/derived_data/bm1_results.RDS"))
saveRDS(mm1_results, here::here("analysis/data/derived_data/mm1_results.RDS"))



bm1_sum <- seqwrap_summarise(bm1_results)
mm1_sum <- seqwrap_summarise(mm1_results)



aic_dat <- mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |>  
      pivot_wider(names_from = model, 
                  values_from = c(ll, bsum)) |> 
      # Compare beta sum to check
      mutate(comp = bsum_mm - bsum_bm) |> 
      # Calculate AIC delta
      mutate(delta_aic = -2 * (ll_mm - ll_bm)) |> 
      inner_join(subset |> 
                       rename(target = id)) 
     
   aic_dat |>    
      ggplot(aes(delta_aic)) + geom_line(stat = "density") + 
      facet_wrap(~ betacat) 
      

aicm <- lm(delta_aic ~ 0 + betacat, data = aic_dat)
summary(aicm)


mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |> 
            
      inner_join(subset |> 
                       rename(target = id)) |>  
      
     
      
      ggplot(aes(quantile)) + geom_histogram() + 
      facet_grid(betacat ~ model)


mm1_sum$evaluations |> 
      mutate(model = "mm") |> 
      bind_rows(
            bm1_sum$evaluations |> 
      mutate(model = "bm")) |> 
            
      inner_join(subset |> 
                       rename(target = id)) |> 
      inner_join(cden |> 
                       rename(target = id)) |> 
      
      select(target, betacat, quantile, model, intensity_cat) |> 

      ggplot(aes(quantile)) + geom_histogram() + 

      
      facet_wrap(intensity_cat ~ model)






mm1_sum$evaluations |> 
      rename(mm = pval) |> 
      inner_join(
            bm1_sum$evaluation |> 
                  rename(bm = pval)
      ) |> 
       inner_join(subset |> 
                       rename(target = id)) |> 
      ggplot(aes(bm, mm)) + geom_point() + 
      facet_wrap(~ betacat)



bm1_sum$evaluations


m <- bm1_results@models[[1]]
m <- mm1_results@models[[1]]


summary(m)


```






see here https://link.springer.com/article/10.1186/1471-2105-11-587







