---
title: "seqwrap: an R package for flexible iterative fitting of high-dimensional data"
subtitle: "Supplementary material"
author:
  - name: Chidimma Echebiri
    email: chidimma.echebiri@inn.no
    affiliation:
      - name: University of Inland Norway
        department: Department of Biotechnology, PhD programme in applied ecology and biotechnology
  - name: Rafi Ahmad
    email: rafi.ahmad@inn.no
    affiliation: 
      - name: University of Inland Norway
        department: Department of Biotechnology
  - name: Stian Ellefsen
    email: stian.ellefsen@inn.no
    affiliation: 
      - name: University of Inland Norway
        department: Department of Public Health and Sport Sciences 
  - name: Daniel Hammarström
    email: daniel.hammarstrom@inn.no
    affiliation: 
      - name: University of Inland Norway
        department: Department of Public Health and Sport Sciences 
    attributes:
      corresponding: true
format:
  typst:
    toc: true
    include-in-header:
      text: |
        #show figure.caption: set text(size: 9pt)
        #show outline: it => {
          v(2em)  // Add vertical space before TOC
          it
        }
csl: resources/bmc-bioinformatics.csl
bibliography: resources/bibliography.bib
editor_options: 
  chunk_output_type: console
---


## Installing seqwrap


The latest version of is available from [CRAN](https://CRAN.R-project.org/package=seqwrap). To install seqwrap: 

```{r}
#| eval: false

# Installing from CRAN
install.packages("seqwrap")

```

To install the developmental version from GitHub:

```{r}
#| eval: false

# Installing from Github
pak::pkg_install("trainome/seqwrap")

```


## Fitting models with seqwrap

The seqwrap package is designed to iterate over multiple targets in a high-dimensional data set to fit a user-defined regression model using a user-selected model fitting function. Each target may represent, e.g., a gene, an exon, a protein, a methylation site, or any other outcome from a high-throughput omics platform. Each target is measured across a set of samples, with a metadata set describing study covariates. In the basic case, data and metadata are combined in each iteration to fit the selected model for a specific target. 

Dungan et al. [@dungan2022] evaluated a senolytic treatment (dasatinib and quercetin) in mouse skeletal muscle subjected to mechanical overload through synergist ablation. The study included sham surgery and a vehicle control, allowing analysis of the effect of senolytic treatment and the interaction between mechanical overloading and senolytic treatment. RNA-seq was performed on 20 animals (5 per group). The data has been downloaded from the gene expression omnibus ([GSE195707](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE195707)) and are made available in the seqwrap package under `dungan_counts`. We will load the data and filter low-expression genes.


```{r}
#| labels: packages-and-data 
#| message: false
#| warning: false

# Load packages
library(tidyverse)
library(seqwrap)
library(edgeR)
library(gt)

# Load data 
metadata <- seqwrap::dungan_counts$metadata
counts <- seqwrap::dungan_counts$countdata

```

The metadata contains the sample ID corresponding to columns in the counts data set. We are using `seq_sample_id` as the column name here. The first rows of the prepared metadata are displayed in @tbl-meta-dungan.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-meta-dungan
#| tbl-cap: "Selected rows from the Dungan et al. metadata data frame."


metadata |> 
  slice_head(n = 2, by = c(treatment, surgery)) |> 
  gt() |> 
  opt_table_font(size = 12)

# Save seq_sample_id for display
samples <- metadata |> 
  slice_head(n = 2, by = c(treatment, surgery)) |> 
  pull(seq_sample_id)

```

The count data matrix can be manipulated as a data frame, with the first column indicating the targets (in this case, gene symbols). Importantly, additional column names have names corresponding to the `seq_sample_id` variable in the metadata. As an example, the first rows of the count data frame are shown for the same samples as in the metadata data frame (see @tbl-dungan-counts). seqwrap expects that targets are unique so before going further we should validate this in the data. When doing so we find that five targets have duplicate rows in the data set, for simplicity, we will remove them.   

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-dungan-counts
#| tbl-cap: "Selected rows from the Dungan et al. counts data frame."

counts |> 
  slice_head(n = 10) |> 
  dplyr::select(gene_name, all_of(samples)) |> 
  gt() |> 
  opt_table_font(size = 12)

```



```{r}

# Find targets with more than a unique row in the data
rm_targets <- counts |> 
  summarise(.by = gene_name, 
            n = n()) |> 
  filter(n > 1) |> 
  pull(gene_name)

# Filter out non-uniqe rows
counts <- counts |> 
  filter(!(gene_name %in% rm_targets))

```


```{r}
#| echo: false

# Count the number of targets
n_genes <- nrow(counts)

```

The data suggests a negative binomial model where the linear predictor can be used to assess the difference between senolytic treatment and control (treat; $\beta_{g,1}$), mechanical overloading compared to sham surgery (mov; $\beta_{g,2}$), and their interaction ($\beta_{g,3}$) for gene *g* (*g* = 1, ..., *m*, here *m* = `r n_genes`) and observation *i* (*i* = 1, ..., *n*, here *n* = 20 rows of the metadata; @eq-m1). We want to control for differences in sequencing depth by including an offset (*s*~i~). We will use the TMM method [@robinson2010] to calculate the effective library size.


$$
\begin{align}
y_{g,i} &\sim \operatorname{NB2}(\mu_{g,i}, \theta_g) \\
\operatorname{log}(\mu_{g,i}) &= \beta_{g,0} + \beta_{g,1}\text{treat}_i + \beta_{g,2}\text{mov}_i + \beta_{g,3}(\text{treat}_i \times  \text{mov}_i) + \operatorname{log}(s_i)
\end{align}
$$ {#eq-m1}

Below, we are using the edgeR package to calculate the library size normalization factor. Effective library sizes are then expressed as relative to the median effective library and log transformed.

```{r}

# Use EdgeR to calculate the TMM
y <- edgeR::DGEList(counts[,-1])
y <- edgeR::calcNormFactors(y)

# Combine the data into metadata
metadata <- metadata |> 
  inner_join(
    y$samples |> 
  tibble::rownames_to_column("seq_sample_id")
  ) |> 
   mutate(
     efflibsize = (lib.size * norm.factors) / median(lib.size * norm.factors),
     efflibsize = log(efflibsize)
   ) 


```


`seqwrap_compose()` is used to combine the data, metadata and the selected modelling function with its arguments into a `swobject`. A negative binomial model can be fitted using e.g., `glmmTMB()` from the glmmTMB package (`glmmTMB::glmmTMB()`) or `MASS::glm.nb()`, here we will use `glmmTMB()` as it can be extended in subsequent modelling steps. 

The specified `glmmTMB()` function (`modelfun` slot) needs a formula and a family function, these are supplied in the arguments slot. The count data are supplied in the `data` slot and the metadata in the `metadata` slot.  Notice that we use the namespace operator (`::`) to specify selected functions to avoid difficulties in parallel processing. The `samplename` indicates the column name in the metadata that gives sample names.   


```{r}

m1.1 <- seqwrap_compose(
  modelfun = glmmTMB::glmmTMB,
  data = counts, 
  metadata = metadata, 
  arguments = list(formula = y ~ treatment * surgery + offset(efflibsize), 
                family = glmmTMB::nbinom2),
  samplename = "seq_sample_id"
  )

```

Next, we will run the model using `seqwrap()`. We recommend using `subset` to specify a smaller number of targets when prototyping the model. Here, we will assess whether the model output matches expectations. By specifying `return_models = TRUE`, all model objects will be returned in the results object, which can be inspected if needed. Here we are using a single core for fitting the models, specified in the `cores` argument.

```{r}

m1.1_temp <- seqwrap(
  m1.1, 
  cores = 1, 
  subset = 1:24, 
  return_models = TRUE, 
  verbose = FALSE
)

```


`seqwrap()` will warn us that we are saving model objects. As we only use 24 targets (i.e., 24 model objects are saved in the environment) in prototyping we can ignore this warning. 

Following iterative fitting using `seqwrap()`, we can summarize the models using `seqwrap_summarise()`, which produces a list of two summaries. Below, we are saving the summaries in an object for later access. Running the `seqwrap_summarise()` function also gives us an overview of the model and successfully summarized targets.

In the `summaries` slot, model parameter values are returned (see @tbl-dungan-sum-temp). These are produced using `broom.mixed::tidy()` which is a general function to retrieve model parameters from many regression model packages in R. 

In the `evaluations` slot, each model is evaluated using tests from the [DHARMa package](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html) (see @tbl-dungan-eval-temp). 

```{r}

m1.1_temp_sum <- seqwrap_summarise(m1.1_temp, verbose = FALSE)

```


```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-dungan-sum-temp
#| tbl-cap: "The first ten rows from summarised preliminary models (model summaries)."


m1.1_temp_sum$summaries |> 
  slice_head(n = 10) |> 
  gt() |> 
  fmt_auto() |> 
  opt_table_font(size = 10)

```



```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-dungan-eval-temp
#| tbl-cap: "The first ten rows from summarised preliminary models (model evaluations). Values are p-values from the DHARMa package (see [here for details](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html))."


m1.1_temp_sum$evaluations |> 
  slice_head(n = 10) |> 
  gt() |> 
  opt_table_font(size = 12)

```


### Specifying summary and evaluation functions

We can specify custom functions for evaluating or summarizing the model object. seqwrap allows for two slots (summaries and evaluations). Each function should take the model object as an argument, and for compatibility with `seqwrap_summarise()`, return a data frame. For our next step, we would like to retrieve the dispersion parameter from each model. We will place this parameter in the evaluation function. We can extract the dispersion parameter on the log scale from the glmmTMB model object. We will construct a function around the extraction and include the observed counts from the data:


```{r}

dispersion_fun <- function(x) {
  
    out <- data.frame(
      dispersion =  data.frame(summary(x$sdr))["betadisp",1],
      dispersion.se = data.frame(summary(x$sdr))["betadisp",2],
      log_mu = log(mean(x$frame$y))
      )
  
    return(out)
  
}
```

Using our preliminary models, we can validate the function for one model.

```{r}
dispersion_fun(m1.1_temp@models[[1]])
```

The output is, as expected, a data frame with three columns. The dispersion parameter is expected to be on the log scale. To validate this we can directly access the estimated dispersion parameter from the model object using `sigma()`, which return the dispersion parameter on the natural scale. 

```{r}
log(sigma(m1.1_temp@models[[1]]))
```

For this particular task, seqwrap includes the function `dispersion_evaluation()` which can be used to extract dispersion parameters from a glmmTMB model object containing a negative binomial model. 

```{r}
seqwrap::dispersion_evaluation(m1.1_temp@models[[1]])
```

The reason for making your own summary or evaluation function could be, e.g., to extract custom contrasts or collect specific optimizer convergence diagnostics. Using a custom function, we could incorporate functionality from packages like marginaleffects [@arel-bundock_how_2024] or DHARMa [@DHARMa].

### Running the full model

When we are happy with preliminary performance on a subset of targets, we can initialize `seqwrap()` to iterate over all targets. We are updating the previous model with the evaluation function directly in `seqwrap()`. Notice that we also remove the `subset` and set `return_models` to `FALSE`. Saving all model objects will be too memory-consuming.

`seqwrap()` will give us a message about the expected update frequency of the progress bar and how many targets we are fitting (if `verbose = TRUE`).

We should expect some errors from modelling many thousands of targets. As seen in the note from `seqwrap()`, some models are associated with warnings from the modelling algorithm and/or errors from the evaluation function.


```{r}


m1.2_results <- seqwrap(
  m1.1,
  eval_fun = dispersion_evaluation,
  cores = parallel::detectCores(), 
  return_models = FALSE,
  verbose = TRUE
)


```


Errors can be inspected by accessing the errors slot. From the inspection below, warnings are at least to some extent related to model convergence.

```{r}

warnings <- m1.2_results@errors |>
  # one row per condition: keep warnings raised at the fit stage
  filter(stage == "fit", type == "warning") |>
  # save in object
  pull(message)

warnings[1]

```


### Using target data for prior information in glmmTMB 

A major benefit of canonical R packages for RNA-seq data modelling is the sharing of information between targets through shrinkage and empirical Bayes strategies. We suggest (see the main text) using priors in glmmTMB as convenient way of flexibly regularize models using an empirical Bayes strategy. In glmmTMB, the `priors` argument can be used to set priors on parameters. We will use the estimated parameters from the first model to set up the priors.

First, we need to extract the parameters. From the evaluation function we get the average log count and the dispersion parameter estimate. From the summaries, we will get the other parameters (summarized in @fig-dungan-estimates).


```{r}

m1.2_sum <- seqwrap_summarise(m1.2_results, verbose = FALSE)

```



```{r}
#| fig-cap: "Estimates from modelling the Dungan data. The dispersion trend is visualized using the default `ggplot`2::geom_smooth`."
#| echo: false
#| message: false
#| warning: false
#| label: fig-dungan-estimates

library(cowplot)
library(ggtext)


p1 <- m1.2_sum$evaluations |> 
  ggplot(aes(log_mu, dispersion)) + 
  geom_point(alpha = 0.2) + 
  geom_smooth(se = FALSE, color = "orange") + 
  theme_classic() +
  labs(x = "Average raw counts log(&mu;) ", 
       y = "Dispersion log(&theta;)") + 
  theme(axis.title.x = element_markdown(), 
        axis.title.y = element_markdown())
  

p2 <- m1.2_sum$summaries |> 
  
  ggplot(aes(estimate)) + 
  geom_density() + 
  facet_wrap(~ term, scales = "free") +
  theme_classic() + 
  labs(x = "Estimate", y = "")


plot_grid(
  plot_grid(NULL, p1, NULL, rel_heights = c(0.2, 1, 0.2), ncol = 1),
            p2, 
          ncol = 2, 
          rel_widths = c(1, 1.5))

```


Except for the dispersion values, we can easily summarize parameter values by their mean and standard deviation. A more elaborate summary of the dispersion values could include an estimate of the mean (log) dispersion, conditional on the average log counts. As the dispersion prior will be slightly different for each target (based on the observed counts), we will combine priors into target-specific data frames and store them as a list. `seqwrap()` will make target-specific list elements available in each iteration, making it possible to build the prior data frame for `glmmTMB()`.

To make this process more user friendly seqwrap (version > 0.8.1) includes a helper function for putting a list of priors together.

```{r}

# Summarise parameters from the main output and organizes a 
# list of target-specific priors
Prior_list <- seqwrap_priors(m1.2_results, data = counts)

```

We can inspect the resulting prior by printing the list object 

```{r}
Prior_list
```

We can also visually inspect the prior using the plot function.

```{r}
#| label: fig-prior-dungan
#| fig-cap: "Empirical Bayes priors estimated from the Dungan data set."
plot(Prior_list)
```

Again, we will combine data, metadata, and arguments using `seqwrap_compose()`. We are now also adding the priors to the arguments. The variables used in the priors data frame are made available from the target-specific list referenced in `targetdata = Prior_list`. Using `arguments = alist(... ` avoids evaluation of the list so that the data frame is not evaluated before being used in each iteration. Priors have been built corresponding what is expected by `glmmTMB()` (see [Priors in glmmTMB](https://glmmtmb.github.io/glmmTMB/articles/priors.html)).

```{r}


m1.3 <- seqwrap_compose(
  modelfun = glmmTMB::glmmTMB,
  data = counts, 
  metadata = metadata, 
  eval_fun = dispersion_evaluation, 
  arguments =  alist(
    formula = y ~ treatment * surgery + offset(efflibsize), 
    family = glmmTMB::nbinom2,
      priors = data.frame(
        prior = prior,
        class = class,
        coef = coef)), 
  targetdata = Prior_list
)


m1.3_temp <- seqwrap(
  m1.3, 
  subset = 1:24, 
  return_models = TRUE, 
  verbose = FALSE
)


summary(m1.3_temp@models[[1]])

```

Running the prototyping seems to work, the new set of models contains the prior information. We will proceed with the full data set.

```{r}

m1.3_results <- seqwrap(
  m1.3, 
  cores = parallel::detectCores(),
  return_models = FALSE, 
  verbose = TRUE
)

m1.3_sum <- seqwrap_summarise(m1.3_results, verbose = FALSE)

```

Regularization increases the number models without convergence issues, in fact all `r n_genes` targets have evaluations and summaries.

The summary data frame is easily manipulated in subsequent analysis steps to, e.g., adjust p-values and create basic visualizations of the results (@fig-dungan-final-estimates). Here we use the `fdrtool()` function from the fdrtool package to calculate a q-value for thresholding significant effects [@strimmer2008].

```{r}
#| fig-cap: "A visualization of results from the regularized model of the Dungan et al. data."
#| echo: false
#| message: false
#| warning: false
#| label: fig-dungan-final-estimates

# Using fdrtool
library(fdrtool)

# Extract z-scores for a term of interest
z <- m1.3_sum$summaries |> 
      filter(term == "treatmentSenolytic:surgeryOverload") |> 
      pull(statistic)

# Calculate  
res <- fdrtool(z, statistic = "normal", verbose = FALSE, plot = FALSE)

m1.3_sum$summaries |> 
  filter(term == "treatmentSenolytic:surgeryOverload") |> 
  mutate(qval = res$qval,
         sig = if_else(qval < 0.05, "significant", "non-significant"), 
         SYMBOL = if_else(sig == "significant", target, "")) |> 
      
  ggplot(aes(estimate, -log2(p.value), 
             color = sig)) + 
  geom_point(alpha = 0.4) + 
  
  geom_text(aes(label = SYMBOL), 
      show.legend = FALSE,
            position = position_nudge(y = 1)) +
  
  theme_classic() + 
  labs(x = "Estimate (log-scaled)", 
       y = "-log<sub>2</sub>(p-value)", 
       color = "q-value < 0.05", 
       subtitle = "Overload:Senolytic treatment") + 
  theme(axis.title.y = element_markdown()) 

  
  
```

The above example shows how to build models with seqwrap. Importantly, we make no claims about the validity of the final model but urge the user to assess assumptions in their data and models as part of a principled workflow.


## Comparing lmerSeq and seqwrap

lmerSeq is a package for fitting RNA-seq data using mixed-effects models assuming Gaussian errors on transformed count data. Here, we perform a basic comparison between lmerSeq [@vestal_lmerseq_2022] and seqwrap, using a subset from the gene expression data set presented in the main text from Pillon et al. [@pillon2022].

### Data preprocessing

Here the Pillion gene expression data is downloaded and transformed using VST transform from DESeq2 as suggested in [@vestal_lmerseq_2022].

```{r}
#| label: comp-data
#| echo: true
#| message: false
#| warning: false

# Need installation from github: https://github.com/stop-pre16/lmerSeq
library(lmerSeq) 
library(DESeq2)
library(edgeR)

library(dplyr)
library(tidyverse)
library(seqwrap)
library(purrr)
library(glmmSeq)
library(ggplot2)

# Load the data from seqwrappaper package
dat <- seqwrappaper::pillon_counts


# Filter low expressed genes and combine after filtering
keep <- filterByExpr(
  dat$countdata[,-1],
  min.count = 10,
  min.total.count = 15,
  large.n = 10,
  min.prop = 0.7,
  group = paste(dat$metadata$group, dat$metadata$time))

countdat <-  dat$countdata[keep,]

metadat <- dat$metadata


# Check if sequence_ID matches.
all(colnames(countdat)[-1] == metadat$seq_sample_id)


# Transform the count data using DESeq2's VST
dds <- DESeqDataSetFromMatrix(countData = countdat[,-1],
                              colData = metadat,
                              design = ~ time * group)

dds <- DESeq(dds)
vsd.fixed <- varianceStabilizingTransformation(dds, blind=F)
vst_expr <- assay(vsd.fixed)





# Add as gene_ids to the VST-transformed counts
rownames(vst_expr) <- countdat$geneid

# subset to only the first fifteen genes
vst_expr <- vst_expr[1:15,]

```


### Fitting models in lmerSeq

lmerSeq has a `lmerseq.fit()` function for building and fitting the model.

```{r}
#| label: lmerseq
#| echo: true
#| message: false
#| warning: false

# fit the model using lmerseq
fit.lmerSeq <- lmerSeq.fit(form = ~ time * group  + (1|id),
                           expr_mat = vst_expr,
                           sample_data = metadat,
                           parallel = FALSE,
                           REML = TRUE)



# Extract coefficients from the models 
coefs <- c("timepost", "timerec", "groupT2D", "timepost:groupT2D", "timerec:groupT2D")

# Extract all coefficients from lmerseq models
lmerseq_summary <- coefs |> 
  map_df(~ lmerSeq.summary(
    lmerSeq_results = fit.lmerSeq,
    coefficient = .x,
    p_adj_method = "BH",
    ddf = "Satterthwaite",
    sort_results = FALSE
  )$summary_table |> 
    mutate(Coefficient = .x)
  )



```


### Fitting models in transformed data in seqwrap

Below we are building the same model with seqwrap. We are using `lmerTest::lmer()` to get p-values from each model. lmerSeq filters targets from the output due to singular fit, here we add a small function that evaluates singular fits in the lme4-based model. We will use only models with non-singular fits in the comparison.   

```{r}
#| label: seqwrap
#| echo: true
#| message: false
#| warnings: false

# make data suitable for seqwrap as it accepts only data frames or lists
vst_expr_df <- data.frame(vst_expr) %>%
  tibble::rownames_to_column(var = "target")

# Creating a evaluation function for finding singular fits
singular_eval <- function(x) {
  out <- data.frame(isSingular = lme4::isSingular(x))
  return(out)
}



# first validate input using seqwrap compose
seqwrap_val <- seqwrap_compose(
  data = vst_expr_df,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = lmerTest::lmer,
  arguments = list(
    formula = y ~ time * group  + (1|id)), 
  eval_fun = singular_eval
)


seqwrap_model <- seqwrap(seqwrap_val,
                        return_models = FALSE,
                        cores = parallel::detectCores(), 
                        verbose = FALSE)



# Check for singular fits and list only targets without 
targets <- seqwrap_summarise(seqwrap_model, 
                             verbose = FALSE)$evaluations |> 
  filter(!isSingular) |> 
  pull(target)


seqwrap_summary_df <- seqwrap_summarise(seqwrap_model, 
                                        verbose = FALSE)$summaries |> 
  filter(target %in% targets)


```

### Comparing results from lmerSeq and seqwrap

Since both lmerSeq and seqwrap uses the same underlying model and data, the output estimates are the same. The estimates and p-values of both approaches are visualized in @fig-comp-visualisation

```{r}
#| label: fig-comp-visualisation
#| echo: false
#| fig-cap: "Comparing estimates from lmerSeq and seqwrap"
#| fig-height: 8


# Select relevant parameters and metrics
seqwrap_df <- seqwrap_summary_df %>%
  subset(!term %in% c("(Intercept)", "sd__(Intercept)", "sd__Observation")) %>%
  dplyr::select("target",  "estimate", "p.value", "term" ) %>%
  mutate(source = "seqwrap")


lmerseq_df <- lmerseq_summary |>
  dplyr::select(gene, Estimate, p_val_raw, Coefficient) |>
  mutate(source = "lmerseq") |>
  #rename to match the columns from seqwrap summary
  dplyr::rename(target = gene,
         estimate = Estimate,
         p.value = p_val_raw,
         term = Coefficient)



p1 <- bind_rows(seqwrap_df, 
          lmerseq_df) |> 
  pivot_wider(names_from = source, 
              values_from = c(estimate, p.value)) |> 
 ggplot(aes(x = estimate_seqwrap, y = estimate_lmerseq, color = target)) +
   geom_point(size = 2) +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
   facet_wrap(~ term) +
   labs(subtitle = "Comparison of Estimates", x = "seqwrap Estimate", y = "lmerseq Estimate", 
        color = "") +
   theme_classic()



p2 <-  bind_rows(seqwrap_df, 
          lmerseq_df) |> 
  pivot_wider(names_from = source, 
              values_from = c(estimate, p.value)) |> 
   ggplot(aes(x = p.value_seqwrap, y = p.value_lmerseq, color = target)) +
     geom_point(size = 2) +
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
     facet_wrap(~ term) +
   labs(subtitle = "Comparison of p-values", x = "seqwrap p-values", y = "lmerseq p-values") +
     theme_classic() + 
  theme(legend.position = "none")


plot_grid(p1, p2, ncol = 1)

```


## Fitting the negative binomial and Poisson OLRE models using seqwrap

In the analyses presented in the main text, we used a negative binomial model as the point of departure for the analyses. The model was fitted in `seqwrap()` using `glmmTMB()` with the `nbinom2()` family function. Below we show the code for fitting the model described in Eq. 1 (Negative binomial (non-informed)).

We used the TMM-method [@robinson2010] for calculating the effective library size before adding it to the metadata as log-transformed library sizes relative to the median-sized library. We are using the DHARMa package [@DHARMa] in a custom evaluation function (`sigma_summary2`), which also include additional metrics. 

```{r}
#| eval: false

### Code chunk not evaluated ###

# Set the number of cores
CORES <- parallel::detectCores()


# Load packages
library(seqwrap)
library(tidyverse)
library(edgeR)

# Loading the data. The Pillon 2022 data set is part of the seqwrap package as `pillon_counts`.
dat <- seqwrappaper::pillon_counts

# Filter low expressed genes and combine after filtering
keep <- filterByExpr(
  dat$countdata[,-1],
  min.count = 10,
  min.total.count = 15,
  large.n = 10,
  min.prop = 0.7,
  group = paste(dat$metadata$group, dat$metadata$time))



countdat <-  dat$countdata[keep,]
metadat <- dat$metadata


# Use EdgeR to calculate the TMM
y <- edgeR::DGEList(countdat[,-1])
y <- edgeR::calcNormFactors(y)

# Store effective library sizes 
libsize <- y$samples |>
    rownames_to_column(var = "seq_sample_id") |>
    select(- group)

# Combine all meta data
metadat <- dat$metadata |>
  inner_join(libsize, by = "seq_sample_id") |>
  mutate(group = factor(group, levels = c("NGT", "T2D")),
         time = factor(time, levels = c("basal", "post", "rec")),
         efflibsize = (lib.size * norm.factors)/median(lib.size * norm.factors),
         ln_efflibsize = log(efflibsize))

# Setting up the evaluation function
sigma_summary2 <- function(x) {

  if(is.null(x$fit$convergence)) {
    conv <- 1
  } else {
    conv <- x$fit$convergence[[1]]
  }

  # Saving SD of Obs Level Rand Eff in the Poisson model

  tidy_coef <- broom.mixed::tidy(x)

  if (any(tidy_coef$group %in% "seq_sample_id")) {
    olre.sd <- tidy_coef |>
      dplyr::filter(group == "seq_sample_id") |>
      dplyr::pull(estimate)
  } else { olre.sd <- NA }



  # Combine all in a data frame
  out <- data.frame(dispersion =  data.frame(summary(x$sdr))["betadisp",1],
                    dispersion.se = data.frame(summary(x$sdr))["betadisp",2],
                    olre.sd = olre.sd,
                    log_mu = mean(predict(x, type = "link")),
                    convergence = conv,
                    pdHess = x$sdr$pdHess,
                    aic = AIC(x))
  return(out)

}




# Setting up the model
m1 <- seqwrap_compose(
  data = countdat,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = glmmTMB::glmmTMB,
  eval_fun = sigma_summary2,
  arguments = list(
    formula = y ~ time * group + offset(ln_efflibsize) + (1|id),
    family = glmmTMB::nbinom2)
  )

m1_results <- seqwrap(m1,
                      return_models = FALSE,
                      cores = CORES)

```

The Poisson Observation-Level Random Effect (OLRE) model was fitted using `glmmTMB()` using the `stats::poisson()` family function. The same data and metadata as prepared above was used. The below code corresponds to Eq. 2 in the main text (Poisson OLRE (non-informed)). Notice that the sample id variable is used to introduce the observation-level random effect into the model.

```{r}
#| eval: false

### Code chunk not evaluated ###
m2 <- seqwrap_compose(
  data = countdat,
  metadata = metadat,
  samplename = "seq_sample_id",
  modelfun = glmmTMB::glmmTMB,
  eval_fun = sigma_summary2,
  targetdata = NULL,
  arguments = list(
    formula = y ~ time * group + 
      offset(ln_efflibsize) + 
      (1|id) +
      (1|seq_sample_id),
      family = stats::poisson)
  )


m2_results <- seqwrap(
      m2,
      return_models = FALSE,
      verbose = FALSE,
      subset = 1:16,
      cores = 1)

```


We regularized models by adding priors based on non-informed models. The general strategy was outlined above for the Dungan data example. Here follows the code for the negative binomial model. 

```{r}
#| eval: false

### Code not evaluated ###

# Prepare priors for the regularized version
Priors_list <- seqwrap_priors(m1_results, data = countdat)

# Fit regularized model 
  m1.reg <- seqwrap_compose(
    data =  countdat,
    metadata =  metadat,
    samplename = "seq_sample_id",
    modelfun = glmmTMB::glmmTMB,
    eval_fun = sigma_summary2,
    targetdata =  Priors_list,
    arguments = alist(
      formula =  y ~ time * group + offset(ln_efflibsize) + (1|id),
      family = glmmTMB::nbinom2,
      priors = data.frame(
        prior = prior,
        class = class,
        coef = coef) ))

  m1.reg_results <- seqwrap(
    m1.reg,
    return_models = FALSE,
    verbose = FALSE,
    #  subset = 1:50, # for prototyping
    cores = CORES)


```


## Analyzing DNA methylation data

The following outlines a partial reanalysis of data on CpG methylation status in human skeletal muscle in response to resistance exercise training, followed by detraining and subsequent retraining [@seaborne2018;@seaborne2018b]. Most of the relevant code is explicitly shown here. In contrast, the version of this case study included in the manuscript has been stripped of some details.


### Packages, raw data and preprocessing

```{r}
#| label: dependencies
#| eval: true
#| 

# Packages

library(GEOquery)      # For accessing the raw data
library(minfi)         # For pre-processing of raw data
library(dplyr)
library(tidyr)
library(gt)
library(ggplot2)
library(stringr)
library(maxprobes)     # For filtering cross-reactive probes 
library(seqwrap)
```


```{r}
#| echo: false
# Load color scale 
source(here::here("analysis/figures/figure-opts.R"))
```

Raw data are downloaded from the [Gene Expression Omnibus](https://www.ncbi.nlm.nih.gov/geo/) using the accession number [GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763) as IDAT files. See the file `analysis/methylation-case-study-dataprep.R` (available in the [project repository](https://github.com/trainome/seqwrappaper)) for details on downloading and preparation of the data.  

```{r}
#| label: paths
#| eval: true
#| echo: false

# Prepare data download. 
# Note for use in the present repo: 
# raw_data and derived data are gitignored so safe for 
# large files. If code-re-use, check the download folder before committing 
# changes to git. 

# The GEO ID
gse_id   <- "GSE114763"

# Set download directories
raw_dir  <- here::here("analysis/data/raw_data", gse_id)
idat_dir <- file.path(raw_dir, "idat")
der_dir  <- here::here("analysis/data/derived_data")

dir.create(idat_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(der_dir,  recursive = TRUE, showWarnings = FALSE)

# The metadata (pheno_raw) and the raw data IDAT files are cached after 
# downloading once. 

# A function for retrying connection to GEO for downloading large data.
# This was created after issues with downloading the large zipped archive
# from GEO. 
retry_url <- function(expr, tries = 5, pause = 5) {
  call <- substitute(expr)          # re-evaluated on each attempt
  env  <- parent.frame()
  for (i in seq_len(tries)) {
    out <- try(eval(call, env), silent = TRUE)
    if (!inherits(out, "try-error")) return(out)
    if (i < tries) Sys.sleep(pause * i)
  }
  stop("failed after ", tries, " attempts: ",
       conditionMessage(attr(out, "condition")))
}

# Series-level metadata
#
# The metadata contains file names and titles include 
# time-points and participant information. 
pheno_file <- file.path(raw_dir, "pheno_raw.rds")

if (!file.exists(pheno_file)) {
  soft_url <- paste0("https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi",
                     "?acc=", gse_id, "&targ=gsm&form=text&view=brief")

  soft <- retry_url({
    res <- curl::curl_fetch_memory(soft_url)
    if (res$status_code != 200) stop("HTTP ", res$status_code)
    strsplit(rawToChar(res$content), "\r?\n")[[1]]
  })

  # one record per sample; "!Sample_<key> = <value>" lines within each
  kv <- tibble::tibble(record = cumsum(startsWith(soft, "^SAMPLE")),
                       line   = soft) |>
    dplyr::filter(startsWith(line, "!")) |>
    tidyr::separate(line, into = c("key", "value"), sep = " = ",
                    extra = "merge") |>
    dplyr::mutate(key = stringr::str_remove(key, "^!Sample_"))

  # characteristics are themselves "<key>: <value>" pairs
  characteristics <- kv |>
    dplyr::filter(key == "characteristics_ch1") |>
    tidyr::separate(value, into = c("key", "value"), sep = ": ",
                    extra = "merge") |>
    dplyr::filter(key %in% c("gender", "age", "weight", "height")) 

  # Cleaning the metadata for downstream processing. 
  pheno_raw <- kv |>
    dplyr::filter(key %in% c("geo_accession", "title")) |>
    dplyr::bind_rows(characteristics) |>
    tidyr::pivot_wider(id_cols = record, names_from = key,
                       values_from = value) |>
    dplyr::select(-record) |> 

        mutate(title_sep = title) |> 
        separate(title_sep, into = c("seq_sample_id", "timeid"), sep = ":") |>
        mutate(timeid = gsub(" Muscle_", "", timeid)) |> 
        mutate(timeid = gsub("Participant_", "", timeid)) |> 
        mutate(timeid = gsub("7wk_", "7wk", timeid)) |> 
        
         separate(timeid, into = c("time", "participant", "replicate"), sep = "_") |> 
      # Making time-point more obvious
        mutate(time = case_when(
              time == "Baseline" ~ "1_baseline", 
              time == "Acute" ~ "2_acute", 
              time == "7wkLoading" ~ "3_loading", 
              time == "7wkUnloading"  ~ "4_unloading", 
              time == "7wkRe-oading" ~ "5_reloading"
        )) |> 
        # Fixing meta-data details
        mutate(weight = gsub(" kg", "", weight), 
               height = gsub("cm", "", height), 
               weight = as.numeric(weight), 
               height = as.numeric(height), 
               age = as.numeric(age)) |> 
        # Fix a misprinted participant id
        mutate(participant = gsub("P006", "006", participant)) 
  

  saveRDS(pheno_raw, pheno_file)
}

pheno_raw <- readRDS(pheno_file)

# Double check 40 expected samples
stopifnot(nrow(pheno_raw) == 40)

# Download and extract raw IDAT files.
# IDATs are distributed gzipped; minfi reads decompressed files
tar_file <- file.path(raw_dir, paste0(gse_id, "_RAW.tar"))

if (!file.exists(tar_file)) {
  # ~722 MB; increase the timeout for slow connections
  options(timeout = 60 * 60)
  retry_url(GEOquery::getGEOSuppFiles(gse_id, baseDir = dirname(raw_dir),
                                      makeDirectory = TRUE))
}

if (length(list.files(idat_dir, pattern = "\\.idat$")) == 0) {
  utils::untar(tar_file, exdir = idat_dir)

  # Unzip idat files
  gz <- list.files(idat_dir, pattern = "\\.idat\\.gz$", full.names = TRUE)
  invisible(lapply(gz, R.utils::gunzip, overwrite = TRUE, remove = TRUE))
}

```

The data downloading process gives us a metadata set describing the participants (@tbl-meta-samples).

```{r}
#| label: tbl-meta-samples
#| tbl-cap: "Sample metadata retreived from GEO ([GSE114763](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114763)) corresponding to @seaborne2018b."
#| echo: false


pheno_raw |> 
      summarise(.by = participant, 
                weight = mean(weight), 
                height = mean(height), 
                age = mean(age)) |> 
      mutate(participant = gsub("00", "", participant)) |> 
      select(participant, age, height, weight) |> 
      gt() |> 
      cols_label(participant = "Participant", 
                 age = "Age (years)", 
                 weight = "Weight (kg)", 
                 height = "Height (cm)")
      
      
```

Checking available IDAT files gives us an overview over the available data over the time-points of the study (@tbl-metadata-count). 

```{r}
#| label: tbl-metadata-count
#| eval: true
#| tbl-cap: "Samples per time-point and participant."


idat_stems <- list.files(idat_dir, pattern = "_Grn\\.idat$") |>
  stringr::str_remove("_Grn\\.idat$")

array_info <- tibble::tibble(stem = idat_stems) |>
  tidyr::separate(stem, into = c("geo_accession", "slide", "array"),
                  sep = "_", remove = FALSE) |>
  dplyr::mutate(Basename = file.path(idat_dir, stem))

metadata <- pheno_raw |>
      # Removing outlier sample, see "Quality Control of DNA Methylation Probes"
      # in @seaborne2018b
      filter(!grepl("SkM_Epi_Mem_1:", title), 
             !grepl("SkM_Epi_Mem_39:", title)) |> 
      
  tibble::as_tibble() |>
  dplyr::select(geo_accession,
                title,
                participant,
                gender,
                age, 
                time) |>
  dplyr::mutate(
    label = stringr::str_remove(title, "^SkM_Epi_Mem_\\d+:\\s*")) |>
  dplyr::left_join(array_info, by = "geo_accession")

stopifnot(!anyNA(metadata$time), !anyNA(metadata$participant))

# 38 arrays: 8 participants x 5 time points with missing values for on 
# participant.

metadata |> 
      summarise(.by = c(time, participant), 
                n = n()) |> 
      pivot_wider(names_from = time, values_from = n) |> 
      select(Participant = participant, 
             Baseline = `1_baseline`, 
             Acute = `2_acute`, 
             Loading = `3_loading`, 
             Unloading = `4_unloading`, 
             Reloading = `5_reloading`) |> 
      gt() |> 
      tab_spanner(label = "Sample time-points", columns = "Baseline":"Reloading")
      



```

We will use the minfi package [@aryee2014] to read the raw data and perform quality control, filtering and normalization. As a first step, array-level filtering is done using detection p-values where methylated and unmethylated positions are compared to negative control positions. This comparison produces a p-value used to identify problematic genomic positions. As shown below (@fig-array-qc), array-level aggregates show mean values < 0.01. 

In the code below, IDAT file names (e.g., `r idat_stems[1]`) are combined with the file tree in `Basename`. `minfi::read.metharray()` iterates over all file names in this vector and imports the data. 


```{r}
#| label: fig-array-qc
#| fig-cap: "Array-level quality control. Mean detection p-values and fraction of failed genomic positions per array subdivided by time-point."
#| eval: true
#| echo: true

rgset <- minfi::read.metharray(metadata$Basename, 
                               extended = TRUE, 
                               force = TRUE)

# Check that orders correspond
stopifnot(identical(colnames(rgset), basename(metadata$Basename)))

colnames(rgset) <- metadata$geo_accession

# Perform detection check  
det_p <- minfi::detectionP(rgset)

qc <- tibble::tibble(
  geo_accession = colnames(det_p),
  mean_det_p    = colMeans(det_p),
  frac_failed   = colMeans(det_p > 0.01))

qc |> 
      inner_join(pheno_raw |> 
                       select(geo_accession, time)) |> 
      ggplot(aes(mean_det_p, frac_failed, color = time)) + 
      geom_point() +
      labs(x = "Array-level mean detection p-value", 
           y = "Fraction of failed genomic positions (p > 0.01)", 
           color = "Time-point")
```

Next, we filter the data using the detection p-value criteria.

```{r}
# All samples pass quality control
keep_samples <- qc$geo_accession[qc$mean_det_p < 0.01]
# length(keep_samples)
rgset    <- rgset[, keep_samples]
det_p    <- det_p[, keep_samples]
metadata <- metadata |> dplyr::filter(geo_accession %in% keep_samples)
```

### Normalization and probe filtering 

Between-array quantile normalization, as implemented by `minfi::preprocessQuantile()` [@aryee2014], is used. An assumption about the data is the we do not expect genome-wide differences in methylation patterns between time-points. Probes are then removed if they (1) fail detection in any retained sample, (2) overlap common SNPs at the target CpG or single-base extension site, (3) appear on published lists of cross-reactive EPIC probes (using the [`maxprobes`](https://github.com/markgene/maxprobes/) package), or (iv) map to the sex chromosomes (all male cohort). 

The same pipeline is run by `analysis/R/methylation-case-study-dataprep.R`, which writes the result to `analysis/data/derived_data/seaborne-gset-quantile.RDS`. We use the resulting "gset" in the permutation study and for the full model run.

```{r}
#| label: normalise-filter
#| eval: true

# Between-array quantile normalization.
#
# NOTE THE RETURN CLASS. preprocessFunnorm(ratioConvert = FALSE) returned a
# GenomicMethylSet, which still carries the methylated and unmethylated
# intensities, so beta could be asked for as getBeta(gset, offset = 100).
# preprocessQuantile() returns a GenomicRatioSet built as
# GenomicRatioSet(Beta = NULL, M = log2(Meth / Unmeth)): there are no
# intensities in it and no offset anywhere, and getBeta() on such an object
# derives beta from M by the inverse logit and ignores `offset`.
# Everything downstream therefore derives beta from M explicitly.
gset <- minfi::preprocessQuantile(
  rgset,
  fixOutliers = TRUE,
  quantileNormalize = TRUE,
  stratified = TRUE,
  # All samples are male and the sex chromosomes are dropped below, so the sex
  # prediction this would otherwise run has nothing to decide.
  sex = rep("M", ncol(rgset))
)

stopifnot(is(gset, "GenomicRatioSet"))

# Initial number of probes
nprobes_init <- dim(gset)[1]

## (i) detection
det_p  <- det_p[match(minfi::featureNames(gset), rownames(det_p)), ]
keep   <- rowSums(det_p < 0.01) == ncol(gset)
gset   <- gset[keep, ]

# Number of probes after filtering
nprobes_detectionfilter <- dim(gset)[1]

## (ii) SNP-affected probes 
gset <- minfi::dropLociWithSnps(gset, snps = c("SBE", "CpG"), maf = 0)

nprobes_snp <- dim(gset)[1]

## (iii) cross-reactive probes 
gset <- maxprobes::dropXreactiveLoci(gset)

nprobes_cross <- dim(gset)[1]

## (iv) sex chromosomes 
anno <- minfi::getAnnotation(gset)
gset <- gset[!anno$chr %in% c("chrX", "chrY"), ]

nprobes_final <- dim(gset)[1]

# The check the beta arm depends on. M is what a GenomicRatioSet stores; beta
# is the inverse logit of it, which is inside the OPEN interval (0, 1) for
# every finite M. This is what replaces the guarantee the Illumina offset used
# to provide, and it is asserted rather than assumed: a non-finite M would come
# from an intensity of exactly zero surviving fixOutliers, and would reach the
# beta family as a response of exactly 0 or 1, which it cannot take.
m_check    <- minfi::getM(gset)
beta_check <- 2^m_check / (1 + 2^m_check)

stopifnot(all(is.finite(m_check)), all(beta_check > 0 & beta_check < 1))

# The observed distance from the boundary, reported below.
beta_min <- min(beta_check)
beta_max <- max(beta_check)
m_range  <- range(m_check)

rm(m_check, beta_check)

```


`r nprobes_init` normalized probes were filtered based on detection which retained `r nprobes_detectionfilter`, after SNP-overlap (n = `r nprobes_snp`), cross-reactivity (n = `r nprobes_cross`) and sex-chromosome mapping filtering, n = `r nprobes_final` remained in the data set. 

### Calculation of methylation quantities

An Illumina array detecting methylation values produces arbitrary signals for the unmethylated (U) and methylated (M) portion of the i^th^ genomic position. Converting these two values to a methylation score can be done using the calculation of a &beta; value.

$$
\beta_i = \frac{M_i}{M_i+U_i}
$$
A bias constant c, is added to the denominator to avoid large variations in the &beta; score when probe intensities are low [@bibikova2006]. A common choice is $c = 100$ [@bibikova2006], which is also the default in the `minfi` package when `type = "illumina"`. However, the successful choice of c is noted to depend on the pre-processing of the data (see `?minfi::getBeta`). Since we are using the `minfi::preprocessQuantile()` function we get the M values directly ($M = \log_2(\text{Meth}/\text{Unmeth})$) without the involvement of the c constant. The &beta; values are instead derived from the normalized M values using the inverse logit: 

$$
\beta_i = \frac{2^{M_i}}{1 + 2^{M_i}}
$$
Here &beta; is restricted to the interval $(0, 1)$ and although this is an intuitive metric for understanding the methylation status at a specific genomic site (e.g., as a percentage methylation), it is an inconvenient quantity for a statistical model that expects normally distributed errors [@laird2010; @du2010]. Indeed, when applying a Gaussian model to identify differentially methylated genomic sites, a logit transformation of the &beta;-values (M values) results in more favorable diagnostics [@du2010].

The observed M-values in the data set span `r sprintf("%.2f", m_range[1])` to `r sprintf("%.2f", m_range[2])`, giving &beta; values between `r sprintf("%.3g", beta_min)` and `r sprintf("%.6f", beta_max)`, i.e. `r sprintf("%.3g", beta_min)` and `r sprintf("%.3g", 1 - beta_max)` away from the two boundaries making it possible to use beta regression to model &beta; values. We can move back and forth between M and &beta; values, to get M from &beta; we use:

$$
M = \log_2 \left( \frac{\beta}{1-\beta} \right)
$$

The relationship between the two transformations can be visualized as shown in @fig-beta-mval.

```{r}
#| label: fig-beta-mval
#| fig-cap: "The logit transform of &beta; values gives M-values."
#| echo: false


beta <- seq(from = 0.001, to = 0.999, by = 0.001)
Mval <- log2(beta/(1-beta))

data.frame(Mval, beta) |> 
      ggplot(aes(Mval, beta)) + geom_line() + 
      theme_classic() + 
      labs(x = "M-values", y = "&beta;-values") + 
      theme(axis.title.y = ggtext::element_markdown())

```

In conclusions, two potential routes for modeling methylation values exists. We can model M values as having normally distributed errors (Gaussian model), or we could model the &beta; values using beta regression. Next we evaluate these model choices. 

### Inferential error rates across models 

We will compare the models using type I error rate and power determined from permutation of within-participant relabeled data. This relabeling establishes a null distribution for statistical tests performed using the within-participant design. The design keeps the within-participant correlation. Our approach is similar to what was described in [@gadbury2008].

We will perform the experiment on permuted data from a subset of genomic sites distributed across different strata of &beta; values. 

A fixed number of sites is sampled from each &beta;-value stratum. Within every permutation, a random subset of those sites receives a known effect (adding a known average effect) while the remainder are left untouched (null-sites). Sites without an effect estimate the type I error rate, sites with an added effect are used to estimate power.

We will use the M-value scale to add the effect to permuted data with effect assignment redrawn in every permutation. The &beta;-values are derived from the manipulated M-values. A shift of &delta; M units is exactly the fixed-effect parameter of the Gaussian model and proportionally to &delta;&times;log(2) for the linear-predictor shift of the &beta; model with the logit link. This means that no model is favored by the parameterization. Our approach differs here from @gadbury2008. We fix the effect sizes while @gadbury2008 uses empirical effect sizes from the data. Our simplified approach enables comparisons between model fitting procedures at known effect sizes and signs across strata of &beta; values.

In the code chunk below, a single site is used to show the workflow. M values are taken from the `gset` object and &beta; values are derived from them. We are using a randomly selected row of the M value matrix and join the values with the meta data. Within each participant, time is re-sampled to create a permuted time factor. Using the permuted time factor, an average effect (&delta; = 0.3) is added to all observations belonging to the permuted time-point `3_loading`. Data are back-transformed to the &beta; scale and both model types are fitted.

We will use a set of different inferential alternatives to establish the most appropriate model and test for our data set. In small samples, linear mixed model parameter p-values from the Wald test are anti-conservative [@luke2017]. When the restricted maximum likelihood (REML) is used in place of maximum likelihood and small sample size-corrections are used (Kenward-Roger or Satterthwaite approximations of degrees of freedom) the type I error rates are controlled at the nominal level [@luke2017]. However, for the generalized linear mixed model (GLMM) with link functions, such as beta regression, the restricted maximum likelihood is not clearly defined [@maestrini2024]. When comparing different REML implementations, @maestrini2024 found that GLMM fitted with REML leads to a reduction in small sample bias and that the use of REML probably should be more widely adopted. The choice of algorithm can be made based on software availability [@maestrini2024]. REML is implemented also for GLMM in the glmmTMB package with the option `REML=TRUE`. For Gaussian models, the results from `glmmTMB` replicates the results of `lme4::lmer`. However, the documentation also says that performance of REML in non-Gaussian models should be verified by simulations (see `?glmmTMB::glmmTMB`).

In addition to REML for mitigation of small sample bias, the Satterthwaite correction is widely adopted in Gaussian mixed-effects models to approximate the denominator degrees of freedom for t-tests of parameter estimates [@kuznetsova2017]. Similarly to REML estimation for GLMM's, the Satterthwaite approximation is indicated for use "at your own risk" (see `?glmmTMB::dof_satt`).

```{r}
# Illustrating data preparation and model settings for a set of 
# candidate models for the methylation data set.

# Prepare data.
#
# We first load M values and derive beta derived from them.
m_vals <- minfi::getM(gset)
beta_vals <- 2^m_vals / (1 + 2^m_vals)

# Setting the seed for site selection and time permutation
set.seed(1)

# Selecting a random site for testing
y <- data.frame(y = m_vals[sample(1:nrow(m_vals), 1),]) |> 
      tibble::rownames_to_column(var = "geo_accession" ) 

# Organizing the data into a format suitable for single model 
# fit. 
dat <- metadata |> 
      dplyr::select(geo_accession, participant, time) |> 
      inner_join(y) |>
      # Permutation of time within participant
      group_by(participant) |> 
      mutate(time_permute = sample(time, replace = FALSE)) |> 
      ungroup() |> 
      mutate(msim = if_else(time_permute == "3_loading", 
             y + 0.3, y), 
             betasim = plogis(msim * log(2))) 


# A general purpose summary function for ML/REML fits.
#
# This function is similar to the `summary_fun` in
# analysis/R/methylation-error-permutation.R). 
temp_summary_fun <- function(m) {

      # Wald p-value (restricted to paremeter 3_loading)
      wald_p <- coef(summary(m))$cond["time_permute3_loading", "Pr(>|z|)"]
      # Wald p-value Satterthwaite
      wald_p_satter <- coef(summary(m, ddf = "satterthwaite"))$cond["time_permute3_loading", "Pr(>|t|)"]

      data.frame(wald_p = wald_p,
                 wald_p_satter = wald_p_satter)

}


# Storing fits in a list 
fits <- list()

# Maximum likelihood M-values
fits[[1]] <- glmmTMB::glmmTMB(msim ~ time_permute + (1|participant), 
                          data = dat)


# Maximum likelihood beta regression
fits[[2]] <- glmmTMB::glmmTMB(betasim ~ time_permute + (1|participant), 
                          data = dat, 
                          family = glmmTMB::beta_family)

# REML M-values
fits[[3]]<- glmmTMB::glmmTMB(msim ~ time_permute + (1|participant), 
                           REML = TRUE,
                          data = dat)

# REML beta regression
fits[[4]] <- glmmTMB::glmmTMB(betasim ~ time_permute + (1|participant), 
                          data = dat, 
                          REML = TRUE,
                          family = glmmTMB::beta_family)



out <- lapply(fits,temp_summary_fun) |> 
      bind_rows() |> 
      mutate(model = c("M-values",
                       "B-values",
                       "M-values",
                       "B-values"), 
             estimator = c("ML", "ML", "REML", "REML"))
```


```{r}
#| tbl-cap: "P-values from a set of model specifications for a specific CpG site"
#| label: tbl-prel-results-meth
#| echo: false

out |>
            dplyr::select(model,estimator, wald_p, wald_p_satter) |>
      gt() |>

      gt::cols_label(wald_p = "Wald z/t-test",
                     wald_p_satter = md("Wald t-test w.<br>Satterthwaite DDF"), 
                     estimator = "Estimator", 
                     model = "Model") 
```




```{r}
# Illustrating the 
es_mmod <- coef(summary(fits[[3]]))$cond["time_permute3_loading","Estimate"]
es_betamod<- coef(summary(fits[[4]]))$cond["time_permute3_loading","Estimate"]

es_betamod <- es_betamod / log(2)

```

The different model specifications produces different p-values (@tbl-prel-results-meth), although, we will need the full permutation experiment to determine any systematic differences. The transformation of the estimates follow the same logic as the transformation of the raw data. The estimated effects (`r round(es_mmod,2)` and `r round(es_betamod,2)`) for the model on M-values and &beta; values respectively are comparable but will differ due to compression in the &beta; values model.

The source code for the full permutation experiment can be found in the script `analysis/R/methylation-error-permutation.R` in the [project repository](https://github.com/trainome/seqwrappaper).


## Simulation results from the RNA-seq analysis

Below follows supplementary results from the simulation study presented in the main text. 

```{r}
#| label: load-sim-results
#| echo: false
#| message: false
#| warning: false


library(cowplot)

source(here::here("analysis/figures/figure-opts.R"))

## Load data from simulations
source(here::here("analysis/R/simulation-functions.R"))

# Extract simulations sorts out all simulations files and combines them.
# This is simulations with dispersion scenario 1.
sim_results <- extract_simulations(
  evaluations_path = here::here("analysis/data/raw_data/evaluations"),
  estimates_path = here::here("analysis/data/raw_data/estimates"),
  populationeffects_path = here::here(
    "analysis/data/raw_data/simdata/popeffect"
  ),
  disp_scenario = "s1"
)


sim_results2 <- extract_simulations(
  evaluations_path = here::here("analysis/data/raw_data/evaluations2"),
  estimates_path = here::here("analysis/data/raw_data/estimates2"),
  populationeffects_path = here::here(
    "analysis/data/raw_data/simdata2/popeffect"
  ),
  disp_scenario = "s2"
)



```




```{r}
#| label: fig-genecounts-simulated
#| fig-cap: "Number of simulated gene targets (true negative and true positives) across dispersion scenarios and sample sizes."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6


# Number of genes in each simulated data set
bind_rows(sim_results$populationeffects, 
          sim_results2$populationeffects) |> 
  # Filter to keep only 1 term
  filter(term == "conditionB") |> 
  mutate(effect = if_else(population_effect == 0, "true.negative", "true.positive")) |> 
  summarise(.by = c(dataset, size, disp_scenario, effect),
            n = n()) |>
  
  mutate(size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small (m = 54, n = 18)", 
                                  "Medium (m = 108, n = 36)", 
                                  "Large (m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         effect = factor(effect, levels = c("true.negative", "true.positive"), 
                         labels = c("True negative", "True positive"))) |> 
  
  ggplot(aes(size, n, fill = disp_scenario)) + 
  geom_point(position = position_jitter(width = 0.05), 
             size = 3, 
             shape = 21,
             alpha = 0.5) +
  labs(x = "Sample size (m samples, n participants)", 
       y = "Gene targets", 
       fill = "Dissersion scenario")  +
  theme_classic() + 
  
  
  
  scale_fill_manual(values = c(colors[c(1,5)])) +
  theme(legend.position = "bottom", 
        legend.title = element_blank(), 
        strip.background = element_blank(), 
        strip.text = element_text(hjust = 0)) + 
  facet_wrap(~ effect, scales = "free", ncol = 1)

```




```{r}
#| echo: false
#| message: false
#| warning: false


# Number of genes per data set with true population effect
true_effects <- bind_rows(sim_results$populationeffects, 
          sim_results2$populationeffects) |> 
  mutate(effect = if_else(population_effect == 0, "true.negative", "true.positive")) |> 
  summarise(.by = c(term, dataset, size, disp_scenario, effect), 
            n = n()) 


# Combine all true/false effects
est_temp <- bind_rows(sim_results$estimates, 
          sim_results2$estimates) |> 
  # Retain only terms of interest.
  filter(term %in% c("conditionB", "timet3:conditionB")) |>
  # dataset was mis-named in the simulations.
  dplyr::rename(dataset = datasets) |>
  dplyr::select(- file)


est <- est_temp |> 
  inner_join(
    bind_rows(sim_results$populationeffects, 
              sim_results2$populationeffects) |>
               mutate(target = as.character(target)) |>
               dplyr::select(-file)) |>

  dplyr::select(target, term, estimate:p.value, population_effect, dataset, model, size, disp_scenario) |>

  mutate(.by = c(model, term, size, dataset, disp_scenario),
         fdr = p.adjust(p.value, method = "fdr")) |>
  mutate(true_effect = if_else(population_effect == 0, "neg", "pos"),
         identified_effect = if_else(fdr > 0.05, "neg", "pos"),
         true_positive = if_else(true_effect == "pos" &
                                   identified_effect == "pos", TRUE, FALSE),
         false_positive = if_else(true_effect == "neg" &
                                    identified_effect == "pos", TRUE, FALSE)) 


```



```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false


convergence_stats <- bind_rows(sim_results$evaluations, 
          sim_results2$evaluations) |> 
  dplyr::select(-file) |> 
  mutate(conv = if_else(model == "m3" & isSingular == TRUE, FALSE, 
                        if_else(model != "m3" & convergence != 0, FALSE, TRUE)), 
         conv.nonstrict = if_else(model != "m3" & convergence != 0, FALSE, TRUE)) |> 
  dplyr::rename(dataset = datasets) |> 
   dplyr::select(target, size, model, dataset, disp_scenario, isSingular, conv, conv.nonstrict) |> 

  inner_join(
    bind_rows(sim_results$populationeffects, 
              sim_results2$populationeffects) |>
               mutate(target = as.character(target)) |>
                dplyr::select(-file) |> 
      mutate(effect = if_else(population_effect == 0, "n", "p")) |> 
       dplyr::select(target, term, dataset, size, disp_scenario, effect) |> 
      pivot_wider(names_from = term, values_from = effect) |> 
      mutate(effect = paste0(conditionB, ":", `timet3:conditionB`)) |> 
      mutate(.by = c(dataset, size, effect, disp_scenario), 
             ntotal = n()) |> 
       dplyr::select(-conditionB, - `timet3:conditionB`) 
  )  |> 
  
  summarise(.by = c(model, dataset, disp_scenario, effect, size), 
            conv = sum(conv),
            conv.nonstrict = sum(conv.nonstrict),
            ntotal = mean(ntotal)) 

# Save for manuscript  
saveRDS(convergence_stats, here::here("analysis/data/derived_data/convergence_stats.RDS"))

```


```{r}
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| label: fig-convergence
#| fig-cap: "Number of models that provided estimates as a percentage of all potential targets. In (A), a strict criteria was used for convergence for the Gaussian model of transformed counts where models with singular fits where excluded as suggested by Vestal et al. (2022). In (B) all converged models are included."
#| fig-height: 8



p1 <- convergence_stats |> 

  mutate(success = 100 *  (conv / ntotal), 
         size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small\n(m = 54, n = 18)", 
                                  "Medium\n(m = 108, n = 36)", 
                                  "Large\n(m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"), 
                        labels = c("Negative binomial", 
                                   "Regularized Negative binomial", 
                                   "Gaussian transformed counts", 
                                   "Poisson OLRE", 
                                   "Regularized Poisson OLRE"))) |> 
  ggplot(aes(model, success, color = model)) +
  geom_point(position = position_jitter(width = 0.1)) + 
  facet_grid(disp_scenario ~ size) +
  
  labs(y = "Converged/non-singular models (%)", 
       x = "Model") +
  
  theme_classic() + 
  theme(panel.background = element_rect(fill = "gray95"), 
        strip.background = element_blank(), 
        strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(hjust = 0), 
        legend.title = element_blank(), 
        axis.text.x = element_blank(), 
        legend.position = "bottom", 
        legend.box = "vertical") + 
  scale_color_manual(values = colors) + 
  guides(color=guide_legend(nrow=3, byrow=TRUE))
  
  
  
p2 <-  convergence_stats |> 

  mutate(success = 100 *  (conv.nonstrict / ntotal), 
         size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small\n(m = 54, n = 18)", 
                                  "Medium\n(m = 108, n = 36)", 
                                  "Large\n(m = 216, n = 72)")), 
         disp_scenario = factor(disp_scenario, 
                                levels = c("s1", "s2"), 
                                labels = c("High dispersion\nvariability", 
                                           "Low dispersion\nvariability")), 
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"), 
                        labels = c("Negative binomial", 
                                   "Regularized Negative binomial", 
                                   "Gaussian transformed counts", 
                                   "Poisson OLRE", 
                                   "Regularized Poisson OLRE"))) |> 
  ggplot(aes(model, success, color = model)) +
  geom_point(position = position_jitter(width = 0.1)) + 
  facet_grid(disp_scenario ~ size) +
  
  labs(y = "Converged models (%)", 
       x = "Model") +
  
  theme_classic() + 
  theme(panel.background = element_rect(fill = "gray95"), 
        strip.background = element_blank(), 
        strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(hjust = 0), 
        legend.title = element_blank(), 
        axis.text.x = element_blank(), 
        legend.position = "none")  + 
  scale_color_manual(values = colors) 
  
plot_grid(p1, 
          p2, 
          ncol = 1, 
          rel_heights = c(1, 0.8),
          labels = c("A", "B"))  

```





```{r}
#| label: fig-pdist
#| fig-cap: "Distribution of unadjusted p-values from all null-effects in models of simulated data. Non-uniform distributions indicate mis-specified models, i.e. models that do not capture variation in the data. Models are the negative binomial (NB), regularized negative binomial (NB-R), the Gaussian model of transformed counts (Gaussian), the Poisson observation-level random effect model (POLRE) and the regularized Poisson observation-level random effect model (POLRE-R)."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 6


# P-value distributions for null genes

est |>
  filter(population_effect == 0) |>
  
  mutate(model_coef = factor(paste0(model,"_", term), 
                             levels = c("m1_conditionB", 
                                        "m2_conditionB", 
                                        "m3_conditionB", 
                                        "m4_conditionB", 
                                        "m5_conditionB",
                                        "m1_timet3:conditionB", 
                                        "m2_timet3:conditionB",
                                        "m3_timet3:conditionB",
                                        "m4_timet3:conditionB",
                                        "m5_timet3:conditionB"
                                        ), 
                             labels = c("NB\nMain", 
                                        "NB-R\nMain",
                                        "Gaussian\nMain",
                                        "POLRE\nMain",
                                        "POLRE-R\nMain",
                                        "NB\nInteraction",
                                        "NB-R\nInteraction",
                                        "Gaussian\nInteraction",
                                        "POLRE\nInteraction",
                                        "POLRE-R\nInteraction"
                                        )), 
         size_disp = factor(paste0(size,"_", disp_scenario), 
                            levels = c("small_s2", "medium_s2", "large_s2", 
                                       "small_s1", "medium_s1", "large_s1"), 
                            labels = c("Small\nLow dispersion", 
                                       "Medium\nLow dispersion", 
                                       "Large\nLow dispersion", 
                                       "Small\nHigh dispersion", 
                                       "Medium\nHigh dispersion", 
                                       "Large\nHigh dispersion"
                                       ))) |> 
  
  ggplot(aes(p.value)) +
  geom_histogram(color="black", fill=colors[2], binwidth = 0.025,
                 boundary = 0, closed = "left", 
                 linewidth = 0.1) +
  scale_x_continuous(breaks = c(0.05, 0.5, 1)) +
  facet_grid(model_coef ~ size_disp) + 
  theme_classic() + 
  theme(strip.background = element_blank(), 
        strip.text = element_text(size = 8), 
        axis.text = element_text(size = 8)) +
  labs(x = "Un-adjusted P-value", 
       y = "")



```






```{r}
#| label: fig-cor-sim
#| fig-cap: "Average correlations between simulated population effects and estimates as a function of sample size in two parameters (Main and interaction effects) and two dispersion scenarios."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 6


# Model averages
mod_avg <- est |>

  filter(population_effect != 0) |>

  mutate(.by = c(model, size, term, dataset, disp_scenario),
         rank_true = rank(population_effect),
         rank_obs = rank(estimate)) |>

  summarise(.by = c(model, size, term, disp_scenario),
            cor = cor(rank_true, rank_obs)) |>

 mutate(size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small", "Medium", "Large")),
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"),
                        labels = c("Negative binomial (non-informed)",
                                   "Regularized Negative binomial",
                                   "Gaussian transformed counts",
                                   "Poisson OLRE (non-informed)",
                                   "Regularized Poisson OLRE")), 
         disp_scenario = factor(disp_scenario, levels = c("s2", "s1"), 
                                labels = c("Low variability", 
                                           "High variability")), 
        term = factor(term, levels = c("conditionB", "timet3:conditionB"), 
                      labels = c("Main effect", 
                                 "Interaction effect"))) 



est |>

  filter(population_effect != 0) |>

  mutate(.by = c(model, size, term, dataset, disp_scenario),
         rank_true = rank(population_effect),
         rank_obs = rank(estimate)) |>

  summarise(.by = c(model, size, term, dataset, disp_scenario),
            cor = cor(rank_true, rank_obs)) |>

 mutate(size = factor(size, levels = c("small", "medium", "large"), 
                       labels = c("Small", "Medium", "Large")),
         model = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"),
                        labels = c("Negative binomial (non-informed)",
                                   "Regularized Negative binomial",
                                   "Gaussian transformed counts",
                                   "Poisson OLRE (non-informed)",
                                   "Regularized Poisson OLRE")), 
         disp_scenario = factor(disp_scenario, levels = c("s2", "s1"), 
                                labels = c("Low variability", 
                                           "High variability")), 
        term = factor(term, levels = c("conditionB", "timet3:conditionB"), 
                      labels = c("Main effect", 
                                 "Interaction effect")))  |>
  
  ggplot(aes(size, cor, group = paste(dataset, model),
             color = model)) +
  
    
  geom_line(alpha = 0.3) +
  
  geom_point(data = mod_avg, 
             color = "black",
             aes(group = NULL, shape = model, fill = model), 
             size = 3) +
  
  scale_fill_manual(values = c(colors[1], colors[2], colors[4], colors[1], colors[2])) +
  scale_color_manual(values = c(colors[1], colors[2], colors[4], colors[1], colors[2], 
                                colors[1], colors[2], colors[4], colors[1], colors[2])) +
  scale_shape_manual(values = c(22,22, 23, 25, 25)) +
  

  facet_grid(term ~ disp_scenario) +
  theme_classic() + 
  theme(strip.background = element_blank(), 
        legend.title = element_blank(), 
        axis.title.x = element_blank()) +
  labs(y = "Average correlation (r)")



```



```{r}
#| eval: false
#| echo: false

est |>

  filter(population_effect != 0) |>
  filter(dataset == 2, size == "medium") |>

  mutate(.by = c(model, term, dataset, disp_scenario),
         rank_true = rank(population_effect),
         rank_obs = rank(estimate)) |>

  ggplot(aes(rank_true, rank_obs)) +
  geom_point(alpha = 0.3) +
  facet_grid(paste(model) ~ paste(term, disp_scenario))




```


## Supplementary tables


```{r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Gene set enrichment analysis of fold-changes in all models. For each model the top (and bottom) ranked gene sets were selected based on -log~10~(FDR) &times; NES. The sign of the normalized effect size (NES) indicate enrichment at top (positive fold-change in T2D vs. control) and bottom (negative fold change) ranked genes."
#| label: tbl-gsea



library(gt)
library(stringr)


# To get the gsea results we need to run the figure 4 script
if(!file.exists(here::here("analysis/data/derived_data/gsea_results.RDS"))) {
      source(here::here("analysis/figures/figure-4.R"))
} 
      

gsea_results <- readRDS(here::here("analysis/data/derived_data/gsea_results.RDS"))


gsea_up <- gsea_results |> 
  
  slice_max(order_by = (-log10(p.adjust) * NES), n = 2, by = model) |> 
  dplyr::select(-core_enrichment) |> 
    mutate(model_descr = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"),
                        labels = c("Negative binomial (non-informed)",
                                   "Regularized Negative binomial",
                                   "Gaussian transformed counts",
                                   "Poisson OLRE (non-informed)",
                                   "Regularized Poisson OLRE"))) 

gsea_down <- gsea_results |> 
  
  slice_max(order_by = -(-log10(p.adjust) * NES), n = 2, by = model) |> 
  dplyr::select(-core_enrichment) |> 
    mutate(model_descr = factor(model, levels = c("m1", "m2", "m3", "m4", "m5"),
                        labels = c("Negative binomial (non-informed)",
                                   "Regularized Negative binomial",
                                   "Gaussian transformed counts",
                                   "Poisson OLRE (non-informed)",
                                   "Regularized Poisson OLRE"))) 



bind_rows(gsea_up, gsea_down) |>  
          mutate(NES = round(NES, 2)) |>
  dplyr::select(model_descr,ID, Description, setSize, NES, p.adjust) |> 
  mutate(Description = str_to_title(Description)) |> 
  
 gt(groupname_col = "model_descr") |> 
  fmt_scientific(columns = p.adjust) |> 
  opt_table_font(size = 10) |> 
  cols_label("ID"=  "Gene ontology ID", 
             "setSize" = "Set size", 
             "p.adjust" = "FDR")






```


## Comparing elapsed time 

Choosing the most appropriate modeling strategy for high-dimensional data can entail a high computational cost. `seqwrap()` records the elapsed time for fitting a set of models. This estimate can assist in selecting modeling strategies. To retrieve the elapsed time:

```{r}
# Example of elapsed time using different modeling strategies. 
library(seqwrap)

set.seed(1) 
dat <- simcounts(n_genes = 100)

# Store data sets
d <- dat$data
md <- dat$metadata


mod1 <- seqwrap_compose(data = d, 
                metadata = md, 
                modelfun = stats::lm, 
                arguments = list(formula = y ~ x), 
                samplename = "sample")

mod2 <- seqwrap_compose(data = d, 
                metadata = md, 
                modelfun = lme4::lmer, 
                arguments = list(formula = y ~ x + (1|cluster)), 
                samplename = "sample")

mod3 <- seqwrap_compose(data = d, 
                metadata = md, 
                modelfun = glmmTMB::glmmTMB, 
                arguments = list(formula = y ~ x + (1|cluster)), 
                samplename = "sample")

results1 <- seqwrap(mod1, verbose = FALSE)
results2 <- seqwrap(mod2, verbose = FALSE)
results3 <- seqwrap(mod3, verbose = FALSE)

results1@elapsed_time
results2@elapsed_time
results3@elapsed_time

```

Below, we are comparing three different algorithms, `lme4::glmer.nb()`, `glmmTMB()` and `glmmTMB()` using a non-default optimizer. Using a subset gives indications on which algorithm is the fastest (@fig-timing). Differnt modeling packages are expected to have different computational costs depending on the model definition and the data [@brooks_glmmtmb_nodate]. Options within packages can also influence elapsed times, see e.g., the [lme4 performance tips](https://cran.r-project.org/web/packages/lme4/vignettes/lmerperf.html) vignette. 


```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-timing
#| fig-cap: "Comparing elapsed time between three different algorithms using the same basic model (negative binomial on simulate count data with 8 clusters and 2 observations each). "


dat <- simcounts(n_genes = 1000)

d <- dat$data
md <- dat$metadata

md$libsize <- log(colSums(d[,-1]))

available_cores <- parallel::detectCores()

size <- c(62, 125, 250, 500)
cores <- c(2, available_cores, available_cores, available_cores)

time_results <- data.frame(size = size, 
                      mod1 = rep(NA, length(size)), 
                      mod2 = rep(NA, length(size)), 
                      mod3 = rep(NA, length(size)))

for(i in seq_along(size)) {
  
  d_sub <- d[1:size[i], ]
  
  
  
  mod1 <- seqwrap_compose(data = d_sub, 
                metadata = md, 
                modelfun = lme4::glmer.nb, 
                arguments = list(formula = y ~ x + (1|cluster) + offset(libsize)), 
                samplename = "sample")
  
  mod2 <- seqwrap_compose(data = d_sub, 
                metadata = md, 
                modelfun = glmmTMB::glmmTMB, 
                arguments = list(formula = y ~ x + (1|cluster) + offset(libsize), 
                                 family = glmmTMB::nbinom2), 
                samplename = "sample")
  
  mod3 <- seqwrap_compose(data = d_sub, 
                metadata = md, 
                modelfun = glmmTMB::glmmTMB, 
                arguments = list(formula = y ~ x + (1|cluster) + offset(libsize), 
                                 family = glmmTMB::nbinom2, 
                                 control=glmmTMB::glmmTMBControl(optimizer=optim,
                                                        optArgs=list(method="BFGS"))), 
                samplename = "sample")
  
  
  
  
  
  
  results1 <- seqwrap(mod1, verbose = FALSE, cores = cores[i])
  results2 <- seqwrap(mod2, verbose = FALSE, cores = cores[i])
  results3 <- seqwrap(mod3, verbose = FALSE, cores = cores[i])
  
  
  time_results[i,2] <- results1@elapsed_time[3]
  time_results[i,3] <- results2@elapsed_time[3]
  time_results[i,4] <- results3@elapsed_time[3]

}



time_results |> 
  pivot_longer(cols = c(mod1, mod2, mod3)) |> 
  
  mutate(name = factor(name, levels = c("mod1", "mod2", "mod3"), 
                       labels = c("lme4::glmer.nb", 
                                  "glmmTMB (defualt optimizer)", 
                                  "glmmTMB optimizer: 'optim' - method 'BFGS'"))) |> 
  
  ggplot(aes(size, value, group = name, color = name)) + 
  geom_line() +
  geom_point() + 
  theme_classic() + 
  labs(color = "", 
       x = "Number of targets in data set", 
       y = "Computation time (sec)")
  
  



```














