---
title: "Pillon et al. Re-analysis"
format: html
editor_options: 
  chunk_output_type: console
---

Pillon et al. [-@pillon2022] investigated gene expression profiles in individuals with type 2 diabetes and healthy control individuals in response to a standardized exercise session. RNA was extracted from muscle biopsy samples obtained before and immediately after exercise, as well as after a three-hour recovery. Paired-end RNA-sequencing was performed, and the data were submitted to the Gene Expression Omnibus (GSE202295). Here we have downloaded the available raw counts matrix data, which has been processed into counts from raw reads by NCBI.




After filtering low-expression genes (`filterByExpr` from the `edgeR` package), the total library size was quantified and normalized using the trimmed mean of M-values [@robinson2010]. Normalized library sizes were further expressed as a ratio to the median normalized library size and log-transformed. 

```{r}
#| label: setup-and-data-prep
#| code-fold: true

library(seqwraphelper)
library(seqwrap)
library(tidyverse)
library(ggtext)
library(glmmTMB)
library(edgeR)
library(DESeq2)


dat <- seqwraphelper::geo_data(gse = "GSE202295")



all(dat$metadata$seq_sample_id == colnames(dat$countdata[,-1]))

# Combine all gene counts after filtering
keep <- filterByExpr(
  dat$countdata[,-1], 
  min.count = 10, 
  min.total.count = 15, 
  large.n = 10, 
  min.prop = 0.7, 
  group = paste(dat$metadata$group, dat$metadata$time))

countdat <-  dat$countdata[keep,] 


# Use EdgeR to calculate the TMM
y <- edgeR::DGEList(countdat[,-1])
y <- edgeR::calcNormFactors(y)

# Store library sizes
libsize <- y$samples |> 
  rownames_to_column(var = "seq_sample_id") |> 
  select(- group)

# Combine all meta data
metadat <- dat$metadata |> 
  inner_join(libsize, by = "seq_sample_id") |> 
  mutate(group = factor(group, levels = c("NGT", "T2D")), 
         time = factor(time, levels = c("basal", "post", "rec")), 
         efflibsize = (lib.size * norm.factors) / median(lib.size * norm.factors),
         ln_efflibsize = log(efflibsize))



```

### A preliminary model

Using the data from Pillon et al. [-@pillon2022], we are interested in baseline differences in gene expression profiles between individuals with type 2 diabetes and healthy controls. Additionally, we want to determine differences in responses to the exercise session. The data have been collected using a repeated measures design where samples are obtained from the same individuals at three time-points. Using a mixed-effects model we could first get estimates for the baseline differences and the differences between groups in changes from baseline at while controlling for correlations among samples due to the repeated design. The deterministic part of our model will thus be:

$$

\eta_i = \beta_0 + \beta_1 \text{group}_{\text{diabetes}} + \beta_2 \text{time}_{t2} + \beta_3 \text{time}_{t3} +\\
\beta_4 ( \text{time}_{t2}\times\text{group}_{\text{diabetes}}) + \beta_5 ( \text{time}_{t3}\times\text{group}_{\text{diabetes}}) +
b_{0[i,j]}
$$
Where $b_{0[i,j]}$ represents a varying intercept per participant ($b_0 \sim \operatorname{Normal}(0, \sigma_b)$).

Count data are traditionally modeled as following a conditional Poisson or negative binomial distribution. However, count data have also been fitted as normally distributed models after transformation.   


We first used a preliminary negative binomial model to estimate distributions of parameter values seen in our data. In our preliminary model we modeled the data ($y$) using a conditional negative binomial distribution

$$
y \sim \operatorname{NB}(\mu, \phi)
$$
where the average count depends on the deterministic model using a log link $\mu = \operatorname{exp}(\eta)$. Using estimates from the preliminary model we simulated data with a similar data structure and the observed mean-dispersion relationship as observed in our data. Using the simulated data we will compare three modelling strategies, (1) a naive negative binomial model, (2) a two-step model with regularization applied to parameter values using prior distributions estimated from model 1, and (3) a model using a conditional normal distribution where the count data has been transformed to remove the relationship between the mean and variance (variance stabilizing transformation, available in `DESeq2`). 

To get 

```{r}
#| label: defining-an-eval-function
#| code-fold: true 


# Summary function to return the dispersion parameter with SE on the log scale
# mean(predict(x, type = "link)) will give us the predicted log counts.
# We will put this in the eval fun to also get estimates of the parameters in 
# the generic summary function.
sigma_summary <- function(x) {

  out <- data.frame(dispersion =  data.frame(summary(x$sdr))["betadisp",1],
                    dispersion.se = data.frame(summary(x$sdr))["betadisp",2],
                    log_mu = mean(predict(x, type = "link"))) 
  return(out)
  
}
```







```{r}
m1 <- seqwrap_compose(
  data = countdat, 
  metadata = metadat, 
  samplename = "seq_sample_id", 
  modelfun = glmmTMB, 
  eval_fun = sigma_summary,
  arguments = list(
    formula = y ~ time * group + offset(ln_efflibsize) + (1|id),
    family = glmmTMB::nbinom2)
  )




m1_results <- seqwrap(m1, 
                      return_models = FALSE, 
                      subset = 1:500, 
                      cores = 10) 




m1_sum <- seqwrap_summarise(m1_results)

```



Simulating data with known 

```{r}


m1_sum$summaries |> 
  select(target, term, estimate) |> 
  bind_rows(
    m1_sum$evaluations |> 
  mutate(term = "dispersion", 
         dispersion = dispersion) |> 
  select(target, term, estimate = dispersion)
  ) |> 
  ggplot(aes(estimate)) + geom_density() + 
  facet_wrap(~term, scales = "free")


metadat |> 
  summarise(.by = c(group, time), 
            n = n() )


# Fit a trend to the dispersion data from m1
# save the data in a convenient format.
dispersion_dat <- m1_sum$evaluations 



# Fit a loess model
trend_model <- loess(dispersion ~ log_mu, 
                     data = dispersion_dat, 
                     span = 0.3, 
                     weights = 1/(dispersion.se^2))





# Set the number of truly different genes
nullgenes <- 100
condB_true <- 50
condB_time2_true <- 50


# The simcounts2 simulation function will provied 

simdat <- simcounts2(n1 = 17,
                          n2 = 20, 
                     beta = runif(nullgenes + condB_true + condB_time2_true,
                                  min = 1, 
                                  max = 7.5),
                     conditionB = c(rep(0, nullgenes),
                                    sample(c(0.4, -0.4), condB_true, replace = TRUE), 
                                    rep(0, condB_time2_true)),
                     timet2 = rnorm(nullgenes + condB_true + condB_time2_true,
                                    0, 
                                    0.15),
                     timet3 = rnorm(nullgenes + condB_true + condB_time2_true,
                                    0, 
                                    0.25),
                     conditionB_timet2 = rep(0, nullgenes + 
                                               condB_true + 
                                               condB_time2_true),
                     conditionB_timet3 = c(rep(0, nullgenes), 
                                           rep(0, condB_true),
                                            sample(c(0.4, -0.4),
                                                   condB_time2_true,
                                                   replace = TRUE)),
                     b0 = rlnorm(nullgenes + condB_true + condB_time2_true,
                                 log(0.2), 
                                 0.75),
                     b1 = rlnorm(nullgenes + condB_true + condB_time2_true,
                                 log(0.1), 
                                 0.5),
                     b2 = rlnorm(nullgenes + condB_true + condB_time2_true,
                                 log(0.1), 
                                 0.5), 
                     phi_model = trend_model)



# VST transformation ###################################################
# The VST transformation 
dds <- DESeqDataSetFromMatrix(countData = as.matrix(simdat$counts[,-1]), 
                       colData = simdat$metadata, 
                       design = ~ time * condition)

dds <- DESeq(dds, quiet = TRUE)
vst_mat <- assay(varianceStabilizingTransformation(dds, blind = FALSE))
vst_dat <- cbind(data.frame(gene = simdat$counts[,1], as.data.frame(vst_mat)))

# Model 1 ##########################
# Fitting a naive Neg-Binom model ##
simdat$metadata$ln_libsize <- log(  simdat$metadata$library_size / median(simdat$metadata$library_size)  )


ms1  <- seqwrap_compose(
  data = simdat$counts, 
  metadata = simdat$metadata, 
  samplename = "seq_sample_id", 
  modelfun = glmmTMB::glmmTMB, 
  eval_fun = sigma_summary,
  arguments = list(
    formula = y ~ time * condition + offset(ln_libsize) + (1|id),
    family = glmmTMB::nbinom2)
  )



ms1_results <- seqwrap(
  ms1,
  return_models = FALSE, 
  subset = 1:50, 
  cores = 10) 



ms1_sum <- seqwrap_summarise(ms1_results)




# Model 2 ###############################################################
# An informed GLMM using data-wide-informed estimates to inform priors ##

## Fitting a model for the mean-dispersion relationship 

# Fit a trend to the dispersion data from m1
# save the data in a convenient format.
dispersion_dat <- ms1_sum$evaluations 

# Fit a loess model
trend_model <- loess(dispersion ~ log_mu, 
                     data = dispersion_dat, 
                     span = 0.3, 
                     weights = 1/(dispersion.se^2))



target_dat <- dispersion_dat |> 
  mutate(global_sd = 1 , 
         dispersion_prior = paste0("normal(", round(dispersion, 3),",", round(global_sd, 3), ")"), 
                                  Class = "fixef_disp", 
                                  Coef = "1")



ms2 <- seqwrap_compose(
  data = simdat$counts, 
  metadata = simdat$metadata, 
  samplename = "seq_sample_id", 
  modelfun = glmmTMB::glmmTMB, 
  eval_fun = sigma_summary,
                targetdata =  data.frame(Prior = target_dat$dispersion_prior,
                                        Class = rep("fixef_disp",nrow(target_dat)), 
                                        Coef = rep("1", nrow(target_dat))),
                arguments = alist(
                  formula = y ~ time * condition + offset(ln_libsize) + (1|id),
                  family = glmmTMB::nbinom2,
                  priors = data.frame(
                    prior = Prior, 
                    class = Class, 
                    coef = Coef) ))

ms2_results <- seqwrap(
  ms2,
  return_models = FALSE, 
  subset = 1:50, 
  cores = 10) 


ms2_sum <- seqwrap_summarise(ms2_results)


# Model 3 ###############################################################
# A Gaussian model on VST-transformed count data ########################

ms3 <-  seqwrap_compose(
  data = vst_dat, 
  metadata = simdat$metadata, 
  samplename = "seq_sample_id", 
  modelfun = lme4::lmer, 
  eval_fun = NULL,
  arguments = list(
    formula = y ~ time * condition + (1|id))
  )

ms3_results <- seqwrap(
  ms3,
  return_models = FALSE, 
  subset = 1:50, 
  cores = 10) 



ms3_sum <- seqwrap_summarise(ms3_results)





library(GGally)

plot(seq(from = 0, to = 2, by = 0.01),
      dlnorm(seq(from = 0, to = 2, by = 0.01), log(0.1), 0.5), 
     type = "l")
  
  rlnorm(ng, log(1.25), 0.1)



m1_sum$summaries |> 
  select(target, term, estimate) |> 
  bind_rows(
    m1_sum$evaluations |> 
  mutate(term = "dispersion", 
         dispersion = dispersion) |> 
      select(target, term, estimate = dispersion)) |>
  arrange(target) |> 

  pivot_wider(names_from = term, 
              values_from = estimate) |> 
  select(-target) |> 
  ggpairs()






```



















```{r}




# Predicted values
dispersion_dat$trend <- predict(trend_model)
# Residuals
dispersion_dat$resid <- dispersion_dat$dispersion - dispersion_dat$trend

# Prior variance
prior_var <- var(dispersion_dat$resid)


# Shrink values
dispersion_dat <- dispersion_dat |> 
  mutate(
    # posterior var
    posterior_var = dispersion.se^2, 
    # Shrinkage weights
    shrink_weights = posterior_var / (prior_var + posterior_var), 
    # Shrink
    dispersion_est = (1 - shrink_weights) * dispersion + shrink_weights * trend
  )

# Dispersion SE is related to log(mu)
dispersion_dat |> 
  ggplot(aes(log_mu, dispersion.se)) + geom_point() +
  coord_cartesian(ylim = c(0, 2)) 



# Shrinking the dispersion towards 

dispersion_dat |>

  ggplot(aes(log_mu, dispersion)) + 

  geom_pointrange(aes(ymin = dispersion - dispersion.se, 
                      ymax = dispersion + dispersion.se)) + 
  
  geom_line(aes(log_mu, trend), 
            color = "blue", 
            linewidth = 2) +
  
  geom_point(aes(log_mu, dispersion_est), 
             color = "red") +
  
  
  coord_cartesian(ylim = c(0, 40)) 
  

# Is shrinkage 
dispersion_dat |> 
  ggplot(aes(dispersion_est - dispersion)) + geom_density() +
  coord_cartesian(xlim = c(-3, 3))
  



# Using shrunken dispersion values

# Only genes with estimates
countdat2 <- countdat |> 
  filter(geneid %in% dispersion_dat$target)

# 



dispersion_dat <- dispersion_dat |> 
  mutate(global_sd = sd(resid), 
         Prior = paste0("normal(", round(dispersion_est, 3),",", round(global_sd, 3), ")"), 
                                  Class = "fixef_disp", 
                                  Coef = "1")




m2 <- seqwrap_compose(data = countdat2, 
                metadata = metadat, 
                samplename = "seq_sample_id", 
                modelfun = glmmTMB, 
                eval_fun = sigma_summary,
                targetdata = data.frame(sigma_est = dispersion_dat$dispersion_est),
                arguments = list(formula = y ~ time * group + offset(ln_efflibsize) + (1|id), 
                                 family = glmmTMB::nbinom2,
                                 
                                 start = alist(betadisp = sigma_est), 
                                 map = list(betadisp = factor(NA))
                                 ))




m3 <- seqwrap_compose(data = countdat2, 
                metadata = metadat, 
                samplename = "seq_sample_id", 
                modelfun = glmmTMB::glmmTMB, 
               # eval_fun = sigma_summary,
                targetdata =  data.frame(Prior = dispersion_dat$Prior,
                                        Class = rep("fixef_disp",nrow(dispersion_dat)), 
                                        Coef = rep("1", nrow(dispersion_dat))),
                arguments = alist(formula = y ~ time * group + offset(ln_efflibsize) + (1|id), 
                                 family = glmmTMB::nbinom2,
                                 priors = data.frame(prior = Prior, 
                                                    class = Class, 
                                                    coef = Coef) ))




m2_results <- seqwrap(m2, 
                      return_models = TRUE, 
                    #  subset = 1:50, 
                      cores = 10) 


m3_results <- seqwrap(m3, 
                      return_models = TRUE, 
                    #  subset = 1:50, 
                      cores = 10) 

summary(m3_results@models[[1]])
summary(m2_results@models[[1]])


m3_results@errors[2,2][[1]]

m2_sum <- seqwrap_summarise(m3_results)


m2_sum$evaluations 


m2_sum$summaries |> 
  filter(term == "timerec:groupT2D")  |> 
  select(target, 
         estimate.m2 = estimate, 
         pval.m2 = p.value) |> 
  inner_join( m1_sum$summaries |> 
  filter(term == "timerec:groupT2D") |> 
  filter(target %in% dispersion_dat$target) |> 
    select(target, 
           estimate.m1 = estimate, 
           pval.m1 = p.value), 
  by = "target") |> 
  
  ggplot(aes(estimate.m2, -log10(pval.m2))) + 
  
  geom_point(color = "red") + 
  geom_point(aes(estimate.m1, -log10(pval.m1)), 
             color = "black") +
  
  geom_segment(aes(x = estimate.m1, xend = estimate.m2, 
                   y = -log10(pval.m1), yend = -log10(pval.m2)))
  
  
  m2_sum$summaries |> 
  filter(term == "timerec:groupT2D")  |> 
  select(target, 
         estimate.m2 = estimate, 
         pval.m2 = p.value) |> 
  inner_join( m1_sum$summaries |> 
  filter(term == "timerec:groupT2D") |> 
  filter(target %in% dispersion_dat$target) |> 
    select(target, 
           estimate.m1 = estimate, 
           pval.m1 = p.value), 
  by = "target") |> 
    ggplot(aes( -log10(pval.m1),  -log10(pval.m2))) + geom_point() + 
    geom_abline(intercept = 0, slope = 1)

  
  

   



summary(m2_results@models[[1]])
  
  exp(eval_dat$sigma_est[1])
  
  m2_results@errors[1,2][[1]]
  
  
  
    




```






